### 2023-01-16 docs/chapter_03.qmd
add README and docs

---
title: "Chapter 3"
format: html
---

```{mermaid}
gantt
    dateFormat  YYYY-MM-DD
    title       Capstone implementation timeline

    section Pre-capstone proposal
    Chapter 01            :done,    pre1, 2023-01-10,2023-01-12
    Chapter 02            :active,  pre2, 2023-01-17, 3d
    Chapter 03            :         pre3, 2023-01-19, 2d

    section Data
    Data ingestion pipelines            :datapipe, after pre3, 2023-01-23

    section Modelling
    Establish set of models             :active, m1, after pre3, 1w
    Establish model performance evaluation metrics :active, m2, after a1, 1w
    Implement model training pipelines  :modpipe, after pre3  , 2w
    Train model A                       :after modpipe, 1w
    Train model B                       :after modpipe, 1w
    Train model C                       :after modpipe, 1w
    Evaluate models                     :eval, after modpipe, 4w

    section Capstone paper
    Draft findings                      :draft, after eval, 2w
    Create figures                      :figs, after draft, 1w
    Copyediting and revision            :final, after figs, 2w
```

### 2023-01-17 src/foresight/dags/ingestion/assets.py
add directories and stubs for ingestion pipeline

@@ -1,0 +1,117 @@
+import pandas as pd
+from dagster import Definitions
+from dagster import AssetSelection
+from dagster import MonthlyPartitionsDefinition
+from dagster import HourlyPartitionsDefinition
+from dagster import TimeWindowPartitionsDefinition
+from dagster import MultiPartitionsDefinition
+from dagster import StaticPartitionsDefinition
+from dagster import TimeWindowPartitionMapping
+from dagster import AssetIn
+from dagster import Output
+from dagster import define_asset_job
+from dagster import asset
+from dagster import ExpectationResult
+from dagster import MetadataValue
+
+def calculate_bytes(df):
+    return None
+
+"""
+fifteen_minutes_partitions_def = MultiPartitionsDefinition({
+    "quarterhour": TimeWindowPartitionsDefinition(start="20200101010101", fmt='%Y%m%d%H%M%S', cron_schedule="*/15 * * * *"),
+    "location": StaticPartitionsDefinition(["ETH", "MAM", "NGR"]),})
+
+monthly_partitions_def = MultiPartitionsDefinition({
+    "month": MonthlyPartitionsDefinition(start_date="2020-01-01"),
+    "location": StaticPartitionsDefinition(["ETH", "MAM", "NGR"]),})
+"""
+fifteen_minutes_partitions_def = TimeWindowPartitionsDefinition(start="20200101000000",
+                                                                fmt='%Y%m%d%H%M%S',
+                                                                cron_schedule="*/15 * * * *")
+monthly_partitions_def = MonthlyPartitionsDefinition(start_date="2020-01-01")
+
+@asset(partitions_def=monthly_partitions_def, code_version="1")
+def acled(context) -> pd.DataFrame:
+    df = pd.DataFrame()
+    context.log_event(
+        ExpectationResult(
+            success=len(df) > 0,
+            description="ensure dataframe has rows",
+            metadata={
+                "text_metadata": "Text-based metadata for this event",
+                "dashboard_url": MetadataValue.url(
+                    "http://mycoolsite.com/url_for_my_data"
+                ),
+                "raw_count": len(df),
+                "size (bytes)": calculate_bytes(df),
+            },
+        )
+    )
+    yield Output(df)
+
+@asset(
+    #partitions_def=fifteen_minutes_partitions_def,
+    partitions_def=monthly_partitions_def,
+    code_version="1",
+)
+def gdelt(context) -> pd.DataFrame:
+    df = pd.DataFrame()
+    context.log_event(
[…]

### 2023-01-17 src/foresight/dags/ingestion/resources.py

@@ -1,0 +1,33 @@
+
+from dagster import resource
+import fsspec
+
+@resource
+def gcp_container(init_context):
+    class GCPContainer:
+        def __init__(self, resource_config):
+            token = resource_config["token"]
+            project = resource_config["project"]
+            self.bucket = resource_config["bucket"]
+            # storage_options can be passed to dask dataframe's `.to_parquet`
+            self.storage_options = {"token": token,
+                                    "project": project}
+            # fspec-compatible filesystem interface to GCP storage
+            self.fs = fsspec.filesystem('gcs', **self.storage_options)
+            self.prefix = "gs://"
+    return GCPContainer(init_context.resource_config)
+
+@resource
+def local_storage(init_context):
+    class LocalStorage:
+        def __init__(self, resource_config):
+            token = None
+            project = resource_config["project"]
+            self.bucket = resource_config["bucket"]
+            # storage_options can be passed to dask dataframe's `.to_parquet`
+            self.storage_options = {"token": None,
+                                    "project": project}
+            # fspec-compatible filesystem interface to local disk
+            self.fs = fsspec.filesystem('file')
+            self.prefix = "file://"
+    return LocalStorage(init_context.resource_config)

### 2023-01-17 src/foresight/dags/repositories.py

@@ -1,0 +1,23 @@
+
+from dagster import repository
+from dagster import graph
+
+from dags.ingestion.resources import gcp_container
+from dags.ingestion.resources import local_storage
+from dags.ingestion.assets import gdelt
+from dags.ingestion.assets import acled 
+from dags.ingestion.assets import training
+
+@graph
+def run_ingestion():
+    return training(gdelt(), acled())
+
+@repository
+def local_ingestion_repository():
+    #return [partitioned_asset_job.to_job(resource_defs={"foresight_gdelt": local_storage})]
+    return [run_ingestion.to_job(resource_defs={"foresight_gdelt": local_storage})]
+
+@repository
+def cloud_ingestion_repository():
+    #return [partitioned_asset_job.to_job(resource_defs={"foresight_gdelt": gcp_container})]
+    return [run_ingestion.to_job(resource_defs={"foresight_gdelt": gcp_container})]

### 2023-01-17 src/foresight/dags/ingestion/resources.py
add comment

@@ -2,6 +2,8 @@
 from dagster import resource
 import fsspec
 
+# TODO implement as IO manager
+# https://docs.dagster.io/concepts/io-management/io-managers#defining-an-io-manager
 @resource
 def gcp_container(init_context):
     class GCPContainer:
@@ -17,6 +19,8 @@ def gcp_container(init_context):
             self.prefix = "gs://"
     return GCPContainer(init_context.resource_config)
 
+# TODO implement as IO manager
+# https://docs.dagster.io/concepts/io-management/io-managers#defining-an-io-manager
 @resource
 def local_storage(init_context):
     class LocalStorage:

### 2023-01-17 src/foresight/dags/repositories.py
remove stale comments

@@ -14,10 +14,8 @@ def run_ingestion():
 
 @repository
 def local_ingestion_repository():
-    #return [partitioned_asset_job.to_job(resource_defs={"foresight_gdelt": local_storage})]
     return [run_ingestion.to_job(resource_defs={"foresight_gdelt": local_storage})]
 
 @repository
 def cloud_ingestion_repository():
-    #return [partitioned_asset_job.to_job(resource_defs={"foresight_gdelt": gcp_container})]
     return [run_ingestion.to_job(resource_defs={"foresight_gdelt": gcp_container})]

### 2023-01-17 docs/chapter_03.qmd
add data lineage diagram

@@
 ---
+## Trustworthiness of data
+
+```{mermaid}
+graph TD
+    A[Conflict event] -->|reporter| B(Online news article)
+    B --> C
+
+    subgraph GDELT
+        C --> |Google Chrome Compact Language Detector - CLD2|D(Language detection)
+        D --> |Conditional random field and various lang-specific|E(Word segmentation)
+        E --> F(Morphological analysis)
+        F --> G(Sentiment and emotional assessment)
+        G --> H(Sentence and clause segmentation)
+        H --> I(Low pass translation)
+        H --> J(Mid pass translation)
+        H --> K(Toponymic translation)
+        H --> L(Well known entity translation)
+        I --> M(Candidate translation ranking)
+        M --> O(Chosen translation)
+        J --> M
+        M --> |some langs|N(High pass translation)
+        N --> O
+        O --> |Moses and other grammer models|Q(User adaptation - dynamic language models)
+        Q --> R(High pass UNK substitution)
+        R --> S(Recapitalization models)
+        S --> T(English translation, events, entities, geocoding, and scores)
+
+        T --> |Universal Sentence Encoder| U(English article embeddings)
+
+        K --> T
+        L --> T
+
+    end
+
+    B --> |human and machine processes|CC(Judgement against ACLED event inclusion criteria)
+    subgraph ACLED
+        CC --> |human and machine processes|DD[Coded events, categories, event actors, geolocations, reported fatalities]
+
+        DD --> |three rounds of human review|DD
+        DD --> EE(Weekly aggregations of events and fatalities)
+    end
+
+    EE --> V(Timeseries dataset)
+    U --> V
+
+    V --> |Our model| W(Our conflict predictions)
+
+```
+
+## Capstone project timeline
 

### 2023-01-18 src/foresight/dags/ingestion.py
reorganize ingestion pipeline

@@ -1,3 +1,9 @@
+import json
+import math
+import calendar
+import itertools
+import urllib.request
+
 import pandas as pd
 from dagster import Definitions
 from dagster import AssetSelection
@@ -11,13 +17,49 @@ from dagster import AssetIn
 from dagster import Output
 from dagster import define_asset_job
 from dagster import asset
+from dagster import AssetMaterialization
 from dagster import ExpectationResult
 from dagster import MetadataValue
+
 
 def calculate_bytes(df):
     return None
 
+# map of month name to month number
+month_map = dict([(m, n) for n, m in enumerate(calendar.month_name[1:], 1)])
+
+# return quarter for given month number
+quarter = lambda m: math.ceil(float(m) / 3)
+
+class SerializableGenerator(list):
+    """Generator that is serializable by JSON"""
+
+    def __init__(self, iterable):
+        tmp_body = iter(iterable)
+        try:
+            self._head = iter([next(tmp_body)])
+            self.append(tmp_body)
+        except StopIteration:
+            self._head = []
+
+    def __iter__(self):
+        return itertools.chain(self._head, *self[:1])
+
+def json_decode_many(s):
+    # https://stackoverflow.com/a/68942444
+    decoder = json.JSONDecoder()
+    _w = json.decoder.WHITESPACE.match
+    idx = 0
+    while True:
+        idx = _w(s, idx).end() # skip leading whitespace
+        if idx >= len(s):
+            break
+        obj, idx = decoder.raw_decode(s, idx=idx)
+        yield obj
+
+
 """
+# TODO try multipartition
 fifteen_minutes_partitions_def = MultiPartitionsDefinition({
     "quarterhour": TimeWindowPartitionsDefinition(start="20200101010101", fmt='%Y%m%d%H%M%S', cron_schedule="*/15 * * * *"),
     "location": StaticPartitionsDefinition(["ETH", "MAM", "NGR"]),})
@@ -31,9 +73,23 @@ fifteen_minutes_partitions_def = TimeWindowPartitionsDefinition(start="202001010
                                                                 cron_schedule="*/15 * * * *")
 monthly_partitions_def = MonthlyPartitionsDefinition(start_date="2020-01-01")
 
-@asset(partitions_def=monthly_partitions_def, code_version="1")
+@asset(
+	partitions_def=monthly_partitions_def,
+	code_version="1",
+	io_manager_key="io_manager"
+)
[…]

### 2023-01-18 src/foresight/dags/ingestion/resources.py

@@ -1,37 +1,0 @@
-
-from dagster import resource
-import fsspec
-
-# TODO implement as IO manager
-# https://docs.dagster.io/concepts/io-management/io-managers#defining-an-io-manager
-@resource
-def gcp_container(init_context):
-    class GCPContainer:
-        def __init__(self, resource_config):
-            token = resource_config["token"]
-            project = resource_config["project"]
-            self.bucket = resource_config["bucket"]
-            # storage_options can be passed to dask dataframe's `.to_parquet`
-            self.storage_options = {"token": token,
-                                    "project": project}
-            # fspec-compatible filesystem interface to GCP storage
-            self.fs = fsspec.filesystem('gcs', **self.storage_options)
-            self.prefix = "gs://"
-    return GCPContainer(init_context.resource_config)
-
-# TODO implement as IO manager
-# https://docs.dagster.io/concepts/io-management/io-managers#defining-an-io-manager
-@resource
-def local_storage(init_context):
-    class LocalStorage:
-        def __init__(self, resource_config):
-            token = None
-            project = resource_config["project"]
-            self.bucket = resource_config["bucket"]
-            # storage_options can be passed to dask dataframe's `.to_parquet`
-            self.storage_options = {"token": None,
-                                    "project": project}
-            # fspec-compatible filesystem interface to local disk
-            self.fs = fsspec.filesystem('file')
-            self.prefix = "file://"
-    return LocalStorage(init_context.resource_config)

### 2023-01-18 src/foresight/dags/parquet_io_manager.py

@@ -1,0 +1,77 @@
+import os
+from typing import Union
+
+import pandas
+from dagster import (
+    Field,
+    InputContext,
+    IOManager,
+    OutputContext,
+    _check as check,
+    io_manager,
+)
+from dagster._seven.temp_dir import get_system_temp_directory
+
+
+class PartitionedParquetIOManager(IOManager):
+    """
+    This IOManager will take in a pandas dataframe and store it in parquet at the
+    specified path.
+
+    It stores outputs for different partitions in different filepaths.
+
+    Downstream ops can either load this dataframe into a spark session or simply retrieve a path
+    to where the data is stored.
+    """
+
+    def __init__(self, base_path):
+        self._base_path = base_path
+
+    def handle_output(
+        self, context: OutputContext, obj: pandas.DataFrame
+    ):
+        path = self._get_path(context)
+        if "://" not in self._base_path:
+            os.makedirs(os.path.dirname(path), exist_ok=True)
+
+        if isinstance(obj, pandas.DataFrame):
+            row_count = len(obj)
+            context.log.info(f"Row count: {row_count}")
+            obj.to_parquet(path=path, index=False)
+        else:
+            raise Exception(f"Outputs of type {type(obj)} not supported.")
+
+        context.add_output_metadata({"row_count": row_count, "path": path})
+
+    def load_input(self, context) -> str:
+        path = self._get_path(context)
+
+        return check.failed(
+            f"Inputs of type {context.dagster_type} not supported. Please specify a valid type "
+            "for this input either on the argument of the @asset-decorated function."
+        )
+
+    def _get_path(self, context: Union[InputContext, OutputContext]):
+        key = context.asset_key.path[-1]  # type: ignore
+
+        if context.has_asset_partitions:
+            start, end = context.asset_partitions_time_window
+            dt_format = "%Y%m%d%H%M%S"
+            partition_str = start.strftime(dt_format) + "_" + end.strftime(dt_format)
+            return os.path.join(self._base_path, key, f"{partition_str}.pq")
+        else:
+            return os.path.join(self._base_path, f"{key}.pq")
+
+
+@io_manager(
+    config_schema={"base_path": Field(str, is_required=False)},
+)
+def local_partitioned_parquet_io_manager(init_context):
+    return PartitionedParquetIOManager(
+        base_path=init_context.resource_config.get("base_path", get_system_temp_directory())
+    )
+
+
[…]

### 2023-01-18 src/foresight/dags/repositories.py

@@ -1,21 +1,0 @@
-
-from dagster import repository
-from dagster import graph
-
-from dags.ingestion.resources import gcp_container
-from dags.ingestion.resources import local_storage
-from dags.ingestion.assets import gdelt
-from dags.ingestion.assets import acled 
-from dags.ingestion.assets import training
-
-@graph
-def run_ingestion():
-    return training(gdelt(), acled())
-
-@repository
-def local_ingestion_repository():
-    return [run_ingestion.to_job(resource_defs={"foresight_gdelt": local_storage})]
-
-@repository
-def cloud_ingestion_repository():
-    return [run_ingestion.to_job(resource_defs={"foresight_gdelt": gcp_container})]

### 2023-01-18 docs/chapter_03.qmd
fix data lineage diagram
update project timeline

@@
     A[Conflict event] -->|reporter| B(Online news article)
-    B --> C
+    B --> C(Scraped webpage)
 
@@
     V --> |Our model| W(Our conflict predictions)
-
 ```
@@
     dateFormat  YYYY-MM-DD
-    title       Capstone implementation timeline
+    title      Capstone project timeline
 
-    section Pre-capstone proposal
+    section Project proposal
     Chapter 01            :done,    pre1, 2023-01-10,2023-01-12
@@
     Data ingestion pipelines                :datapipe, after pre3, 2023-01-23
+    Training set and feature preparation    :after datapipe, 2023-02-05
 
     section Modelling
-    Establish set of models             :active, m1, after pre3, 1w
-    Establish model performance evaluation metrics :active, m2, after a1, 1w
+    Establish models and approaches to evaluate     :active, m1, after pre3, 1w
+    Establish performance evaluation metrics        :active, m2, after a1, 1w
     Implement model training pipelines              :modpipe, after pre3  , 2w
-    Train model A                       :after modpipe, 1w
-    Train model B                       :after modpipe, 1w
-    Train model C                       :after modpipe, 1w
-    Evaluate models                     :eval, after modpipe, 4w
+    Train modeling approach A                       :train, after modpipe, 4w
+    Train modeling approach B                       :train, after modpipe, 4w
+    Train modeling approach C                       :train, after modpipe, 4w
+    Evaluate models                                 :eval, after modpipe, 6w
+    Train selected model(s)                         :model, after train, 3w
 
     section Capstone paper
-    Draft findings                      :draft, after eval, 2w
-    Create figures                      :figs, after draft, 1w
-    Copyediting and revision            :final, after figs, 2w
+    Draft findings                      :draft, after train, 2023-04-10
+    Create figures                      :figs, after model, 2023-04-17
+    Copyediting and revision            :final, after draft, 2023-05-04
+    Presentation                        :pres, after final, 2023-05-11
 ```

### 2023-01-29 src/foresight/EDA/ACLED_EDA.py
Initial implementation of adding target candidates

@@ -31,11 +31,6 @@ ACLED['Month_Num'] = ACLED['Month'].map(month_num)
 
 ACLED['Date'] = pd.to_datetime(ACLED['Month'] + ' ' + ACLED['Year'].astype(str))
 
-#simple boolean: Did people die or not?
-ACLED['Fatalities_Bool'] = ACLED['Fatalities']>1
-
-#ACLED['More_Than_Average'] = ACLED['Fatalities']>
-#ACLED['More_Than_Average_Since_2021'] = ACLED['Fatalities']>
 
 
 def get_months (start_month, year, n_months, include_start_month = False):
@@ -73,12 +68,12 @@ def get_months (start_month, year, n_months, include_start_month = False):
 
     return date_range
 
-def create_base_df(start_month, start_year, country, n_months):
+def create_base_df(country, *args, **kwargs):
     """
     Creates a base df of n months from the start date for a given country
     This can be used for further trend analysis
     """
-    daterange = get_months(start_month,start_year, n_months)
+    daterange = get_months(*args, **kwargs)
 
     df = ACLED[
              (ACLED['Country'].values == country)
@@ -88,21 +83,49 @@ def create_base_df(start_month, start_year, country, n_months):
     return df
 
 
-def fatalities_previous_month(start_month, start_year, country):
-    base_df = create_base_df(start_month, start_year, country, 1)
+def fatalities_previous_month(country, start_month, start_year):
+    base_df = create_base_df(country, start_month, start_year, 1)
+    if len(base_df) == 0:
+        return None
     return base_df['Fatalities'].iloc[0]
 
 
-def n_month_mean(start_month, start_year, country, n_months):
-    base_df = create_base_df(start_month, start_year, country, n_months)
+def n_month_mean(country, *args, **kwargs):
+    base_df = create_base_df(country, *args, **kwargs)
     return base_df['Fatalities'].mean()
 
 
-def n_month_trend(start_month, start_year, country, n_months):
-    base_df = create_base_df(start_month, start_year, country, n_months, include_start_month = True)
+def n_month_trend(country, *args, **kwargs):
+    kwargs.update({'include_start_month':True})
+    base_df = create_base_df(country, *args, **kwargs)
[…]

### 2023-01-31 src/foresight/dags/ingestion_pandas.py
fix yearmonth format
add columns for first three countries
add metadata entries
update notebook with first three countries example code

@@ -132,13 +132,22 @@ def fetch_gkg(timestamp):
     _matches = matches.unstack()
     _matches.columns = _matches.columns.droplevel()
     # drop duplicate mentions of country
-    _df = pd.DataFrame(enumerate(list(map(set,_matches.values))), index=_matches.index)
+    # list(dict.fromkeys()) trick gives a set that preserves order
+    _df = pd.DataFrame(enumerate(list(map(lambda x: list(dict.fromkeys(x)),
+                                          _matches.values))), index=_matches.index)
+
     _df = _df.rename(columns={1: 'countries'})
     # remove nans introduced by droplevel
     _df['countries'] = _df['countries'].map(lambda x: list(filter(lambda y: not pd.isna(y), x)))
     _df['num_countries'] = _df['countries'].map(len)
     df = df.join(_df[['countries', 'num_countries']])
     df['num_countries'] = df['num_countries'].fillna(0).astype('int')
+    # add three columns with first three countries
+    for n in range(1, 4):
+        df[f"country-{n}"] = df['countries'].map(lambda c: c[n-1] if isinstance(c, list) and len(c) > n-1 else [])
+    # replace empty lists in new country columns with empty strings
+    for n in range(1, 4):
+        df[f"country-{n}"] = df[f"country-{n}"].map(lambda c: '' if len(c)==0 else c)
     return df
 
 @asset(
@@ -256,7 +265,7 @@ def gdelt(context, gkg, gsg) -> pd.DataFrame:
                                                  right_on='url')
     df_all['year'] = df_all['date'].dt.year
     df_all['month'] = df_all['date'].dt.month
-    df_all['yearmonth'] = df_all['date'].dt.strftime('%Y%d')
+    df_all['yearmonth'] = df_all['date'].dt.strftime('%Y%m')
 
     context.log_event(
         ExpectationResult(

### 2023-01-31 src/foresight/dags/io_pandas.py

@@ -1,3 +1,4 @@
+import hashlib
 import datetime
 from upath import UPath
 from typing import Dict
@@ -40,6 +41,9 @@ class PyArrowParquetIOManager(UPathIOManager):
         return dict()
 
     def dump_to_path(self, context: OutputContext, obj: pd.DataFrame, path: UPath):
+        # make key for dataset schema based on column names
+        dataset_key = hashlib.md5(''.join(obj.columns.values).encode()).hexdigest()[:6]
+        dataset_created = datetime.datetime.utcnow().strftime('%Y%m%d%H%M%S')
         dataset_path = UPath(self._base_path) / path.parts[-2]
 
         write_options = ds.ParquetFileFormat().make_write_options(compression='snappy')
@@ -53,7 +57,7 @@ class PyArrowParquetIOManager(UPathIOManager):
         # for writing and reading--*if* partitioning/appending
         # works the same way (and use pa.write_dataset instead
         # of pq.write_to_dataset)
-        basename_template = "part-{:%Y%m%d}-{{i}}.parquet".format(day_start)
+        basename_template = "part-{:%Y%m%d}-{{i}}_{}.parquet".format(day_start, dataset_key)
 
         written_paths = []
         written_metadata = []
@@ -86,6 +90,8 @@ class PyArrowParquetIOManager(UPathIOManager):
                                                   '%Y%m%d%H%M%S').strftime('%Y/%m')
         asset_obs_kwargs = {'asset_key': path.parts[-2],
                             'metadata': {"num_rows": len(obj),
+                                         "schema_key": dataset_key,
+                                         "dataset_created": dataset_created,
                                          "path": str(dataset_path),
                                          "pq_partition" : pq_partition,
                                          "pq_path": str(dataset_path / pq_partition),

### 2023-01-31 src/foresight/notebooks/reading_parquet_extracting_locations.py

@@ -104,8 +104,8 @@ gkg_ds.schema
 # https://arrow.apache.org/docs/python/dataset.html#projecting-columns
 # https://arrow.apache.org/docs/python/compute.html
 
-#pqds = pq.ParquetDataset('gcs://frsght/datasets/gdelt',
-pqds = pq.ParquetDataset('../datasets/gdelt',
+pqds = pq.ParquetDataset('gcs://frsght/datasets/gdelt',
+#pqds = pq.ParquetDataset('../datasets/gdelt',
                          use_legacy_dataset=False,
                          filters=[('num_countries', '>', 0),
                                   ('lang', '=', 'ENGLISH')])
@@ -274,7 +274,9 @@ _matches
 # create temporary dataframe with unique countries seen in each row
 # as a column containing lists while preserving index that 
 # references original dataframe indices
-_df = pd.DataFrame(enumerate(list(map(set,_matches.values))), index=_matches.index)
+# list(dict.fromkeys()) trick gives a set that preserves order
+_df = pd.DataFrame(enumerate(list(map(lambda x: list(dict.fromkeys(x)),_matches.values))), index=_matches.index)
+print(_df)
 _df = _df.rename(columns={1: 'countries'})
 # lists contain NaNs since the number of extracted countries varies per record,
 # so map a crazy double lambda expression to remove them
@@ -288,6 +290,14 @@ _df
 # finally, join the new columns to the original dataframe
 new_df = df.head().join(_df[['countries', 'num_countries']])
 new_df['num_countries'] = new_df['num_countries'].fillna(0).astype('int')
+
+print(new_df)
+
+for n in range(1, 4):
+    new_df[f"country-{n}"] = new_df['countries'].map(lambda c: c[n-1] if isinstance(c, list) and len(c) > n-1 else [])
+
+for n in range(1, 4):
+    new_df[f"country-{n}"] = new_df[f"country-{n}"].map(lambda c: '' if len(c)==0 else c)
 
 new_df
 

### 2023-01-31 src/foresight/__init__.py
add foresight package to pyproject.toml
bump dependencies

@@ -1,0 +1,1 @@
+__all__ = ["dags", "notebooks"]

### 2023-01-31 src/foresight/EDA/ACLED_EDA.py
Reimplemented all analysis functions as LabelMaker class. This reduces number of times base_df needs to be computed per row

@@ -31,101 +31,124 @@ ACLED['Month_Num'] = ACLED['Month'].map(month_num)
 
 ACLED['Date'] = pd.to_datetime(ACLED['Month'] + ' ' + ACLED['Year'].astype(str))
 
+class LabelMaker():
+    def __init__(self, n_months):
+        self.df = None
+        self.n_months = n_months
+    def get_months(self, start_month, year):
+        n_months = self.n_months
+        if type(start_month) == str:
+            start_month = month_num[start_month]
 
+        month_years = [(year, start_month)]
+        month = start_month
 
-def get_months (start_month, year, n_months, include_start_month = False):
-    """
-    Returns a DF with years/months for the time period begining
-    1 month before start_month and ending n months before start_month
-    """
+        for i in range(n_months):
+            if month <= 1:
+                month = 12
+                year = year - 1
+            else:
+                month = month - 1
 
-    if type(start_month) == str:
-        start_month = month_num[start_month]
+            month_years.append((year, month))
 
-    month_years = []
-    month = start_month
+        start_date = month_years[0]
+        end_date = month_years[-1]
 
-    if include_start_month:
-        month_years.append((year, start_month))
-        n_months = n_months-1
+        date_range = (
+            datetime.datetime(start_date[0], start_date[1], 1)
+            , datetime.datetime(end_date[0], end_date[1], 1)
+        )
 
-    for i in range(n_months):
-        if month <= 1:
-            month = 12
-            year = year - 1
-        else:
-            month = month - 1
+        self.date_range = date_range
+        return date_range
 
-        month_years.append((year, month))
+    def create_base_df(self, country, *args, **kwargs):
+        """
+        Creates a base df of n months from the start date for a given country
+        This can be used for further trend analysis
+        """
+        daterange = self.get_months(*args, **kwargs)
+        df = ACLED[
+            (ACLED['Country'].values == country)
+            & (ACLED['Date'].values >= np.datetime64(daterange[1]))
+            & (ACLED['Date'].values <= np.datetime64(daterange[0]))
+            ].reset_index()
+        self.base_df = df
 
[…]

### 2023-01-31 cloud/__main__.py
add pulumi stack for gcp

@@ -1,0 +1,55 @@
+
+import urllib.request
+
+import pulumi
+from pulumi import ResourceOptions
+from pulumi_gcp import compute
+
+compute_network = compute.Network(
+    "network",
+    auto_create_subnetworks=True,
+)
+
+my_ip = urllib.request.urlopen("http://wtfismyip.com/text").readlines()[0].decode().strip()
+
+compute_firewall = compute.Firewall(
+    "firewall",
+    network=compute_network.self_link,
+    allows=[compute.FirewallAllowArgs(
+        protocol="tcp",
+        ports=["22", "80", "3000"],
+    )],
+    source_ranges=[f"{my_ip}/32"]
+)
+
+# A simple bash script that will run when the webserver is initalized
+startup_script = """#!/bin/bash
+sudo useradd frsght -m -d /home/frsght
+git clone -b cloud https://github.com/ewheeler/foresight /home/frsght/foresight
+"""
+
+instance_addr = compute.address.Address("address")
+compute_instance = compute.Instance(
+    "instance",
+    machine_type="n2-standard-8",
+    metadata_startup_script=startup_script,
+    boot_disk=compute.InstanceBootDiskArgs(
+        initialize_params=compute.InstanceBootDiskInitializeParamsArgs(
+            image="deeplearning-platform-release/pytorch-1-13-cpu",
+            size=100
+        )
+    ),
+    network_interfaces=[compute.InstanceNetworkInterfaceArgs(
+            network=compute_network.id,
+            access_configs=[compute.InstanceNetworkInterfaceAccessConfigArgs(
+                nat_ip=instance_addr.address
+            )],
+    )],
+    service_account=compute.InstanceServiceAccountArgs(
+        scopes=["https://www.googleapis.com/auth/cloud-platform"],
+    ),
+    opts=ResourceOptions(depends_on=[compute_firewall]),
+)
+
+pulumi.export("instanceName", compute_instance.name)
+pulumi.export("instanceIP", instance_addr.address)

### 2023-02-01 docs/cpa.qmd
add cpa to docs

---
title: "Critical path analysis"
format: html
---

```{mermaid}
graph TD
    H[Cloud infra \n\nEST: 1 EFT: 2 \nLST: 3 LFT: 4\n TR: 1 SL: 2] --> A
    A[GDELT \n\nEST: 2 EFT: 4 \nLST: 4 LFT: 6\n TR: 2 SL: 2] -->B(Training dataset \n\nEST: 4 EFT: 5 \nLST: 6 LFT: 7\n TR: 1 SL: 2)
    M[GDELT sample \n\nEST: 1 EFT: 2 \nLST: 2 LFT: 3\n TR: 1 SL: 1] --> I(Sample dataset \n\nEST: 2 EFT: 3 \nLST: 3 LFT: 4\n TR: 1 SL: 1)
    G[ACLED \n\nEST: 1 EFT: 2 \nLST: 2 LFT: 3\n TR: 1 SL: 1] --> B
    G --> I
    B --> C{Working model  \n\nEST: 11 EFT: 13 \nLST: 12 LFT: 14\n TR: 2 SL: 1}
    B --> J{Model evaluation \n\nEST: 6 EFT: 11 \nLST: 7 LFT: 12\n TR: 5 SL: 1}
    C --> |Optional|L{Deployed model \n\nEST: 13 EFT: 14 \nLST: 14 LFT: 15\n TR: 1 SL: 1}
    I --> K{Model prototyping \n\nEST: 3 EFT: 6 \nLST: 4 LFT: 7\n TR: 3 SL: 1}
    K --> J
    J --> N(Draft report \n\nEST: 11 EFT: 13 \nLST: 12 LFT: 14\n TR: 2 SL: 1)
    N --> F
    J --> C
    K -->|Optional| D(API or UI \n\nEST: 6 EFT: 8 \nLST: 13 LFT: 15\n TR: 2 SL: 7)
    F --> E(Final presentation \n\nEST: 14 EFT: 15 \nLST: 15 LFT: 16\n TR: 1 SL: 1)
    C --> F(Final report \n\nEST: 13 EFT: 14 \nLST: 14 LFT: 15\n TR: 1 SL: 1)
```

### 2023-02-01 src/foresight/dags/__init__.py
add dask implementations of ingestion and io

@@ -1,7 +1,7 @@
 import os
 
-from .io_pandas import local_parquet_io_manager
-from .io_pandas import gcp_parquet_io_manager
+from .io_dask import local_parquet_io_manager
+from .io_dask import gcp_parquet_io_manager
 
 # generally not a fan of anything in this file, but dagster docs said to put this in __init__.py
 # https://docs.dagster.io/guides/dagster/using-environment-variables-and-secrets#example-1-per-environment-configuration

### 2023-02-01 src/foresight/dags/ingestion_dask.py

@@ -1,0 +1,305 @@
+import json
+import math
+import gzip
+import shutil
+import calendar
+import datetime
+import pathlib
+import itertools
+import functools
+import urllib.request
+
+import numpy as np
+import pandas as pd
+import dask.dataframe as dd
+from dagster import Definitions
+from dagster import AssetSelection
+from dagster import MonthlyPartitionsDefinition
+from dagster import DailyPartitionsDefinition
+from dagster import TimeWindowPartitionMapping
+from dagster import AssetIn
+from dagster import Output
+from dagster import define_asset_job
+from dagster import asset
+from dagster import ExpectationResult
+
+from dags import deployment_name
+from dags import resources
+
+
+# map of month name to month number
+month_map = dict([(m, n) for n, m in enumerate(calendar.month_name[1:], 1)])
+
+# return quarter for given month number
+quarter = lambda m: math.ceil(float(m) / 3)
+
+def datetime_range(start, end, delta):
+    current = start
+    while current < end:
+        yield current
+        current += delta
+
+def every_n_mins_between(start, end, minutes=15):
+    return (dt.strftime('%Y%m%d%H%M%S') for dt in datetime_range(start, end, datetime.timedelta(minutes=minutes)))
+
+class SerializableGenerator(list):
+    """Generator that is serializable by JSON"""
+
+    def __init__(self, iterable):
+        tmp_body = iter(iterable)
+        try:
+            self._head = iter([next(tmp_body)])
+            self.append(tmp_body)
+        except StopIteration:
+            self._head = []
+
+    def __iter__(self):
+        return itertools.chain(self._head, *self[:1])
+
+def json_decode_many(s):
+    # https://stackoverflow.com/a/68942444
+    decoder = json.JSONDecoder()
+    _w = json.decoder.WHITESPACE.match
+    idx = 0
+    while True:
+        idx = _w(s, idx).end() # skip leading whitespace
+        if idx >= len(s):
+            break
+        obj, idx = decoder.raw_decode(s, idx=idx)
+        yield obj
+
+daily_partitions_def = DailyPartitionsDefinition(start_date="20200101010101", fmt='%Y%m%d%H%M%S')
+
+monthly_partitions_def = MonthlyPartitionsDefinition(start_date="202001", fmt='%Y%m')
+
+@asset(
+	code_version="1",
[…]

### 2023-02-01 src/foresight/dags/io_dask.py

@@ -1,0 +1,116 @@
+import hashlib
+import datetime
+from upath import UPath
+
+import gcsfs
+import pyarrow as pa
+import dask.dataframe as dd
+from dagster import (
+    Field,
+    InitResourceContext,
+    InputContext,
+    OutputContext,
+    StringSource,
+    UPathIOManager,
+    AssetObservation,
+    io_manager,
+)
+
+
+class DaskParquetIOManager(UPathIOManager):
+    extension: str = ".parquet"
+
+    def _rewrite_path_day_to_month(self, path: UPath):
+        day_start = datetime.datetime.strptime(str(path.stem), '%Y%m%d%H%M%S')
+        # rather than write each day's asset partition as a separate 
+        # parquet partition on disk, we will write one parquet per month
+        month_start = day_start.replace(day=1)
+        month_path = UPath(self._base_path) / path.parts[-2]/  f"{month_start.strftime('%Y%m%d%H%M%S')}{self.extension}"
+        return month_path
+
+    def dump_to_path(self, context: OutputContext, obj: dd, path: UPath):
+        dataset_key = hashlib.md5(''.join(obj.columns.values).encode()).hexdigest()[:6]
+        dataset_created = datetime.datetime.utcnow().strftime('%Y%m%d%H%M%S')
+        dataset_path = UPath(self._base_path) / path.parts[-2]
+
+        # object dtypes (like a list of embeddings), must be specified
+        # so pyarrow can write the parquet file correctly
+        schema = "infer"
+        if 'docembed' in obj.columns:
+            schema = {"docembed": pa.list_(pa.float32())}
+
+        day_start = datetime.datetime.strptime(str(path.stem), '%Y%m%d%H%M%S')
+        part_namer = lambda n: "part-{:%Y%m%d}-{}_{}.parquet".format(day_start, n, dataset_key)
+
+        pq_args = {'path': dataset_path, 'schema': schema, 'ignore_divisions': True,
+                   'append': dataset_path.exists(), 'engine': 'pyarrow',
+                   'write_metadata_file': True, 'write_index': False,
+                   'name_function': part_namer}
+        if path.parts[-2] == 'gdelt':
+            pq_args.update({'partition_on': ['year', 'month']})
+
+        obj.to_parquet(**pq_args)
+
+        pq_partition = datetime.datetime.strptime(str(path.stem),
+                                                  '%Y%m%d%H%M%S').strftime('%Y/%m')
[…]

### 2023-02-01 src/foresight/notebooks/reading_parquet_extracting_locations.py

@@ -104,14 +104,13 @@ gkg_ds.schema
 # https://arrow.apache.org/docs/python/dataset.html#projecting-columns
 # https://arrow.apache.org/docs/python/compute.html
 
-pqds = pq.ParquetDataset('gcs://frsght/datasets/gdelt',
-#pqds = pq.ParquetDataset('../datasets/gdelt',
+#pqds = pq.ParquetDataset('gcs://frsght/datasets/gdelt',
+pqds = pq.ParquetDataset('../datasets/gdelt',
                          use_legacy_dataset=False,
-                         filters=[('num_countries', '>', 0),
-                                  ('lang', '=', 'ENGLISH')])
+                         filters=[('lang', '=', 'ENGLISH')])
 # -
 
-pqds.files
+pqds.files[:5]
 
 pqds.schema
 
@@ -125,6 +124,8 @@ print("file size: {} MB".format(pqds_size >> 20))
 
 emb_cols = [f"docembed-{n}" for n in range(512)]
 
+country_cols = [f"country-{n}" for n in range(1, 4)]
+
 # %%time
 # read specific subset of columns from the dataset into memory as a pyarrow.Table
 # (`read_pandas` param means it will read any pandas-specific metatdata that
@@ -133,7 +134,7 @@ emb_cols = [f"docembed-{n}" for n in range(512)]
 # NOTE that the pyarrow.Table API has some useful operations that
 # may be more computationally efficient than pandas
 # https://arrow.apache.org/docs/python/generated/pyarrow.Table.html#pyarrow.Table
-table = pqds.read_pandas(columns=['countries', 'date'] + emb_cols)
+table = pqds.read_pandas(columns=['date'] + emb_cols + country_cols)
 
 # lets see how big our filtered subset is in memory
 print("RSS (RAM): {} MB".format(pa.total_allocated_bytes() >> 20))
@@ -179,12 +180,23 @@ pbar.register()
 
 import dask.dataframe as dd
 
-filters=[('num_countries', '>', 0)]
-ddf = dd.read_parquet('gcs://frsght/datasets/gdelt', columns=['countries', 'date'] + emb_cols,
-                      split_row_groups=True, infer_divisions=True, filters=filters, engine='pyarrow')
-#ddf = dd.read_parquet('../datasets/gdelt', columns=['countries', 'date'] + emb_cols)
+filters = [('year', '=', 2020), ('month', '=', 7)]
[…]

### 2023-02-01 src/foresight/notebooks/reading_parquet_extracting_locations.py
add checks to confirm deduplication attempts

@@ -192,9 +192,15 @@ ddf.npartitions
 
 ddf['url'].nunique().compute()
 
+sum(ddf['url'].isnull().compute())
+
 ddf['GKGRECORDID'].nunique().compute()
 
+sum(ddf['GKGRECORDID'].isnull().compute())
+
 ddf['DocumentIdentifier'].nunique().compute()
+
+sum(ddf['DocumentIdentifier'].isnull().compute())
 
 len(ddf)
 
@@ -218,7 +224,7 @@ with warnings.catch_warnings():
     for n in range(1, 4):
         # emits _lots_ of warnings about dropping timezone info
         country_record_counts = ddf[f"country-{n}"].value_counts().compute()
-        print(n)
+        print(n, sum(ddf[f"country-{n}"].isnull().compute()) / len(ddf[f"country-{n}"]))
         print(country_record_counts)
 
 # %%time

### 2023-02-01 src/foresight/dags/ingestion_dask.py
fix tmp file deletion
improve deduplication attempts

@@ -151,7 +151,6 @@ def fetch_gkg(timestamp):
     ddf = ddf.join(_df[[f"country-{n}" for n in range(1, 4)]])
     # drop records that don't have any country
     ddf = ddf.dropna(subset=['country-1'])
-    ddf = ddf.drop_duplicates(subset=['GKGRECORDID', 'DocumentIdentifier'])
     return ddf
 
 @asset(
@@ -168,6 +167,8 @@ def gkg(context) -> dd.DataFrame:
                                                          partition_start + datetime.timedelta(days=1)))
 
     ddf = dd.concat(list(map(fetch_gkg, every_fifteen_timestamps)))
+    ddf = ddf.drop_duplicates(subset=['DocumentIdentifier'])
+    ddf = ddf.drop_duplicates(subset=['GKGRECORDID'])
     ddf = ddf.repartition(partition_size='100MB')
 
     context.log_event(
@@ -219,7 +220,6 @@ def fetch_gsg(timestamp, tmp_dir):
     # dask reads ndjson in chunks
     ddf = dd.read_json(json_tmp, blocksize=2**28)
     ddf['gsg_file'] = timestamp
-    ddf = ddf.drop_duplicates(subset=['url'])
     return ddf
 
 @asset(
@@ -229,16 +229,21 @@ def fetch_gsg(timestamp, tmp_dir):
 )
 def gsg(context) -> pd.DataFrame:
     partition_key = context.asset_partition_key_for_output()
-    # TODO make tmp_dir configurable
-    tmp_dir='/tmp/foresight'
+    # TODO make tmp_base configurable
+    tmp_base ='/tmp/foresight'
+    tmp_dir = f"{tmp_base}/{context.run.run_id}"
     partition_date_str = partition_key.split('|')[0]
     partition_start = datetime.datetime.strptime(partition_date_str, '%Y%m%d%H%M%S')
 
     every_fifteen_timestamps = list(every_n_mins_between(partition_start,
                                                          partition_start + datetime.timedelta(days=1)))
+    # `fetch_gsg` is not an op, so it doesn't have access to `context`.
+    # create a partial function that provides `tmp_dir`
+    # to `fetch_gsg` that we can still use with `map`
     gsg_fetcher = functools.partial(fetch_gsg,
                                     tmp_dir=tmp_dir)
     ddf = dd.concat(list(map(gsg_fetcher, every_fifteen_timestamps)))
[…]

### 2023-02-01 src/foresight/dags/io_dask.py

@@ -42,11 +42,12 @@ class DaskParquetIOManager(UPathIOManager):
         day_start = datetime.datetime.strptime(str(path.stem), '%Y%m%d%H%M%S')
         part_namer = lambda n: "part-{:%Y%m%d}-{}_{}.parquet".format(day_start, n, dataset_key)
 
+        # `to_parquet` arguments for all datasets
         pq_args = {'path': dataset_path, 'schema': schema, 'ignore_divisions': True,
                    'append': dataset_path.exists(), 'engine': 'pyarrow',
                    'write_metadata_file': True, 'write_index': False,
                    'name_function': part_namer}
-        if path.parts[-2] == 'gdelt':
+        if 'year' in obj.columns and 'month' in obj.columns:
             pq_args.update({'partition_on': ['year', 'month']})
 
         obj.to_parquet(**pq_args)

### 2023-02-01 src/foresight/dags/io_dask.py
add comment

@@ -50,6 +50,12 @@ class DaskParquetIOManager(UPathIOManager):
         if 'year' in obj.columns and 'month' in obj.columns:
             pq_args.update({'partition_on': ['year', 'month']})
 
+        # TODO writes can fail if multiple runs try to append to a parquet
+        # at the same time. there doesn't seem to be a great way to
+        # handle this (on aws at least) https://stackoverflow.com/q/38964736
+        # doesn't seem to lead to corrupt metadata as retrying the run works
+        # for now we'll retry at the job/run level, but if it
+        # happens a lot we can catch pyarrow.lib.ArrowInvalid and retry here
         obj.to_parquet(**pq_args)
 
         pq_partition = datetime.datetime.strptime(str(path.stem),

### 2023-02-01 src/foresight/dags/ingestion_dask.py
use different date field for partitioning
add comments

@@ -163,10 +163,16 @@ def gkg(context) -> dd.DataFrame:
     partition_date_str = partition_key.split('|')[0]
     partition_start = datetime.datetime.strptime(partition_date_str, '%Y%m%d%H%M%S')
 
+    # NOTE hardcoded partition length
     every_fifteen_timestamps = list(every_n_mins_between(partition_start,
                                                          partition_start + datetime.timedelta(days=1)))
 
     ddf = dd.concat(list(map(fetch_gkg, every_fifteen_timestamps)))
+    ddf['DATE'] = ddf['DATE'].map(str).map(lambda x: pd.to_datetime(x, format='%Y%m%d%H%M%S', errors='coerce'))
+    ddf = ddf.dropna(subset=['DATE'])
+    # gdelt's gkg docs say that these are both unique identifiers
+    # http://data.gdeltproject.org/documentation/GDELT-Global_Knowledge_Graph_Codebook-V2.1.pdf
+    # so discard any duplicates
     ddf = ddf.drop_duplicates(subset=['DocumentIdentifier'])
     ddf = ddf.drop_duplicates(subset=['GKGRECORDID'])
     ddf = ddf.repartition(partition_size='100MB')
@@ -235,6 +241,7 @@ def gsg(context) -> pd.DataFrame:
     partition_date_str = partition_key.split('|')[0]
     partition_start = datetime.datetime.strptime(partition_date_str, '%Y%m%d%H%M%S')
 
+    # NOTE hardcoded partition length
     every_fifteen_timestamps = list(every_n_mins_between(partition_start,
                                                          partition_start + datetime.timedelta(days=1)))
     # `fetch_gsg` is not an op, so it doesn't have access to `context`.
@@ -243,6 +250,8 @@ def gsg(context) -> pd.DataFrame:
     gsg_fetcher = functools.partial(fetch_gsg,
                                     tmp_dir=tmp_dir)
     ddf = dd.concat(list(map(gsg_fetcher, every_fifteen_timestamps)))
+    # can't find any field-level docs for gsg
+    # but assume these are meant to be unique like gkg's 'DocumentIdentifier'
     ddf = ddf.drop_duplicates(subset=['url'])
     ddf = ddf.repartition(partition_size='100MB')
 
@@ -274,9 +283,9 @@ def gsg(context) -> pd.DataFrame:
[…]

### 2023-02-01 src/foresight/dags/io_dask.py

@@ -47,6 +47,10 @@ class DaskParquetIOManager(UPathIOManager):
                    'append': dataset_path.exists(), 'engine': 'pyarrow',
                    'write_metadata_file': True, 'write_index': False,
                    'name_function': part_namer}
+        # NOTE these may not be the same as the dagster partition
+        # (which is reflected in pq part filenames `day_start`)
+        # as these columns in gdelt dataset are derived from
+        # GKG's 'DATE' field, which is the article's date of publication
         if 'year' in obj.columns and 'month' in obj.columns:
             pq_args.update({'partition_on': ['year', 'month']})
 

### 2023-02-01 src/foresight/EDA/ACLED_EDA.py
Remapped data source to use environment variable rather than hardcoded path

@@ -8,9 +8,7 @@ import numpy as np
 pd.options.display.expand_frame_repr = False
 pd.options.display.max_rows = 100
 
-#humdata_path = os.path.normpath(os.environ['humdata_path'])
-
-humdata_path = os.path.normpath(r'..\..\..\..\Data\ACLED_via_humdata_012623.xlsx')
+humdata_path = os.path.normpath(os.environ['humdata_path'])
 
 ACLED = pd.read_excel(humdata_path, sheet_name='Data')
 

### 2023-02-02 cloud/bootstrap_instance.py
add script to bootstrap VMs
resolve notebook conflict
add dependency

@@ -1,0 +1,66 @@
+import subprocess
+
+import paramiko
+
+_hostname = subprocess.run(["pulumi", "stack", "output", "instanceIP"], capture_output=True)
+if not _hostname.stderr:
+    hostname = _hostname.stdout.decode().strip()
+print(hostname)
+
+# TODO take cli args
+username = "ewheeler"
+pkey = paramiko.RSAKey.from_private_key_file("/Users/ewheeler/.ssh/google_compute_engine")
+
+client = paramiko.SSHClient()
+client.set_missing_host_key_policy(paramiko.AutoAddPolicy())
+client.load_system_host_keys()
+c = client.connect(hostname, port=22, username=username, pkey=pkey)
+
+_stdin, _stdout, _stderr = client.exec_command("conda info --base")
+if not _stderr:
+    conda_base = _stdout.read().decode()
+    print(conda_base)
+else:
+    conda_base = "/opt/conda"
+    print(_stderr.read().decode())
+
+foresight_dir = "/home/frsght/foresight"
+conda_source = f"source {conda_base}/etc/profile.d/conda.sh"
+conda_activate = f"source {conda_base}/etc/profile.d/conda.sh ; {conda_base}/bin/conda activate ; {conda_base}/bin/conda activate"
+conda_lock_file = "conda_platform_locks/conda-linux-64.lock"
+conda_create = f"{conda_source} && {conda_base}/bin/conda create --name foresight --file {foresight_dir}/{conda_lock_file} -y"
+# eureka! https://plainenglish.io/blog/eureka-i-finally-managed-to-activate-my-conda-environment-in-my-ci-pipeline-a8b578c21f03
+poetry_install = f"cd {foresight_dir} && {conda_base}/bin/conda run -n foresight {conda_base}/envs/foresight/bin/poetry install"
+
+def execute(cmd):
+    # pretend to be a login shell
+    command = f'/bin/bash -lc "{cmd}"'
+    stdin, stdout, stderr = client.exec_command(command)
+    while True:
+        line = stdout.readline()
+        if not line:
+            break
+        print(line, end="")
+    if stderr:
+        print(stderr.read().decode())
+        return None
+
+cmds = ["sudo chown -R ewheeler:ewheeler /home/frsght/foresight",
+        'PS1="${PS1:-}" conda shell.posix activate', # https://github.com/conda/conda/issues/11885#issuecomment-1259508672
[…]

### 2023-02-02 src/foresight/notebooks/reading_parquet_extracting_locations.py

@@ -63,6 +63,7 @@ if IN_COLAB:
 import logging
 import warnings
 import operator
+import datetime
 
 import numpy as np
 import pandas as pd
@@ -169,7 +170,7 @@ df_means
 
 from dask.distributed import Client
 client = Client(n_workers=4, threads_per_worker=2, processes=True,
-                memory_limit='12GB', silence_logs=logging.ERROR)
+                memory_limit='4GB', silence_logs=logging.ERROR)
 
 client
 
@@ -183,6 +184,20 @@ filters=[('num_countries', '>', 0)]
 ddf = dd.read_parquet('gcs://frsght/datasets/gdelt', columns=['countries', 'date'] + emb_cols,
                       split_row_groups=True, infer_divisions=True, filters=filters, engine='pyarrow')
 #ddf = dd.read_parquet('../datasets/gdelt', columns=['countries', 'date'] + emb_cols)
+
+_dates = ddf[['DATE', 'date']]
+_dates.compute()
+
+_dates['DATE'] = _dates['DATE'].map(str).map(lambda x: pd.to_datetime(x, format='%Y%m%d%H%M%S', errors='coerce'))
+
+_dates['delta'] = _dates.apply(lambda r: r['DATE'] - r['date'], axis=1)
+
+dates = _dates.compute()
+dates
+
+sum(dates['agree'])
+
+dates[dates['delta'] > datetime.timedelta(minutes=5)]
 
 ddf.npartitions
 

### 2023-02-02 cloud/bootstrap_instance.py
add supervisor to boostrap_instance script

@@ -1,21 +1,37 @@
+import time
+import pathlib
+import argparse
 import subprocess
 
 import paramiko
 
+parser = argparse.ArgumentParser(description="set up new VM")
+parser.add_argument("username", type=str, default="ewheeler",
+                    help="username on VM", nargs='?')
+parser.add_argument("pkey", type=str, default="/Users/ewheeler/.ssh/google_compute_engine",
+                    help="local path to private key", nargs='?')
+args = parser.parse_args()
+
+username = args.username
+pkey_path = args.pkey
+
+# find this script's directory
+dir_path = pathlib.Path(__file__).resolve().parent
+
+# get IP address of new VM
 _hostname = subprocess.run(["pulumi", "stack", "output", "instanceIP"], capture_output=True)
 if not _hostname.stderr:
     hostname = _hostname.stdout.decode().strip()
 print(hostname)
 
-# TODO take cli args
-username = "ewheeler"
-pkey = paramiko.RSAKey.from_private_key_file("/Users/ewheeler/.ssh/google_compute_engine")
+pkey = paramiko.RSAKey.from_private_key_file(pkey_path)
 
 client = paramiko.SSHClient()
 client.set_missing_host_key_policy(paramiko.AutoAddPolicy())
 client.load_system_host_keys()
 c = client.connect(hostname, port=22, username=username, pkey=pkey)
 
+# find location of conda binary
 _stdin, _stdout, _stderr = client.exec_command("conda info --base")
 if not _stderr:
     conda_base = _stdout.read().decode()
@@ -24,18 +40,23 @@ else:
     conda_base = "/opt/conda"
     print(_stderr.read().decode())
 
+# prepare commands
 foresight_dir = "/home/frsght/foresight"
 conda_source = f"source {conda_base}/etc/profile.d/conda.sh"
 conda_activate = f"source {conda_base}/etc/profile.d/conda.sh ; {conda_base}/bin/conda activate ; {conda_base}/bin/conda activate"
 conda_lock_file = "conda_platform_locks/conda-linux-64.lock"
 conda_create = f"{conda_source} && {conda_base}/bin/conda create --name foresight --file {foresight_dir}/{conda_lock_file} -y"
 # eureka! https://plainenglish.io/blog/eureka-i-finally-managed-to-activate-my-conda-environment-in-my-ci-pipeline-a8b578c21f03
[…]

### 2023-02-02 cloud/__main__.py
clone main branch on vm

@@ -25,7 +25,7 @@ compute_firewall = compute.Firewall(
 # A simple bash script that will run when the webserver is initalized
 startup_script = """#!/bin/bash
 sudo useradd frsght -m -d /home/frsght
-git clone -b cloud https://github.com/ewheeler/foresight /home/frsght/foresight
+git clone https://github.com/ewheeler/foresight /home/frsght/foresight
 """
 
 instance_addr = compute.address.Address("address")

### 2023-02-02 src/foresight/dags/__init__.py
read GCP credentials from env var

@@ -27,7 +27,8 @@ resources = {
             {
                 "base_path": "gs://frsght/datasets",
                 "project": "foresight-375620",
-                "token": "foresight-375620-01e36cd13a77.json"
+                "token": os.getenv("GOOGLE_APPLICATION_CREDENTIALS",
+                                   "foresight-375620-01e36cd13a77.json")
             }
         ),
     },

### 2023-02-02 docs/cpa.qmd
update cpa doc

@@
 ---
+
+```{mermaid}
+gantt
+    dateFormat  YYYY-MM-DD
+    title      Group Four: Predicting Conflict Escalation
+
+    section Precapstone proposal
+    Chapter 01                  :done, pre1, 2023-01-10, 2023-01-12
+    Chapter 02                  :done, pre2, 2023-01-17, 3d
+    Chapter 03                  :done, pre3, 2023-01-19, 2d
+
+    section Compute and Data
+    Cloud infra                 :active, cloud, after pre3, 1w
+    Data ingestion pipelines    :done, datapipe, after pre3, 2w
+    GDELT full dataset          :active, gdelt, after cloud, 2w
+    GDELT sample dataset        :done, gsample, after pre3, 1w
+    ACLED dataset               :done, acled, after pre3, 1w
+    Sample training dataset     :active, sample, after acled, 1w
+    Full training dataset       :dataset, after gdelt, 1w
+
+    section Modelling
+    Establish models and approaches to evaluate :done,  m1, after pre3, 1w
+    Establish performance evaluation metrics    :active, m2, after a1, 1w
+    Model prototyping                           :proto, after dataset, 3w
+    Model evaluation                            :eval, after proto, 5w
+    Working final model                         :model, after eval, 2w
+
+    section Stretch goals
+    API or UI                                   :api, after proto, 2w
+    Deployed model                              :prod, after model, 1w
+
+    section Capstone paper
+    Draft report                                :draft, after eval, 2w
+    Final report                                :report, after draft, 1w
+    Presentation                                :pres, after report, 1w
+```
 
@@
     G[ACLED \n\nEST: 1 EFT: 2 \nLST: 2 LFT: 3\n TR: 1 SL: 1] --> B
+
     G --> I
-    B --> C{Working model  \n\nEST: 11 EFT: 13 \nLST: 12 LFT: 14\n TR: 2 SL: 1}
+    B --> C{Working final model    \n\nEST: 11 EFT: 13 \nLST: 12 LFT: 14\n TR: 2 SL: 1}
     B --> J{Model evaluation \n\nEST: 6 EFT: 11 \nLST: 7 LFT: 12\n TR: 5 SL: 1}
+    I --> K{Model prototyping \n\nEST: 3 EFT: 6 \nLST: 4 LFT: 7\n TR: 3 SL: 1}
+
     C --> |Optional|L{Deployed model \n\nEST: 13 EFT: 14 \nLST: 14 LFT: 15\n TR: 1 SL: 1}
-    I --> K{Model prototyping \n\nEST: 3 EFT: 6 \nLST: 4 LFT: 7\n TR: 3 SL: 1}
+    K -->|Optional| D(API or UI \n\nEST: 6 EFT: 8 \nLST: 13 LFT: 15\n TR: 2 SL: 7)
+
     K --> J
@@
     J --> C
-    K -->|Optional| D(API or UI \n\nEST: 6 EFT: 8 \nLST: 13 LFT: 15\n TR: 2 SL: 7)
     F --> E(Final presentation \n\nEST: 14 EFT: 15 \nLST: 15 LFT: 16\n TR: 1 SL: 1)

### 2023-02-03 src/foresight/dags/__init__.py
optionally read dask scheduler from env

@@ -1,5 +1,8 @@
 import os
 
+from dask.distributed import Client
+client = Client(os.getenv('DASK_SCHEDULER', None))
+
 from .io_dask import local_parquet_io_manager
 from .io_dask import gcp_parquet_io_manager
 

### 2023-02-04 cloud/__main__.py
use dask distributed
bump a few dependencies

@@ -31,7 +31,7 @@ git clone https://github.com/ewheeler/foresight /home/frsght/foresight
 instance_addr = compute.address.Address("address")
 compute_instance = compute.Instance(
     "instance",
-    machine_type="n2-standard-8",
+    machine_type="n2-highmem-8",
     metadata_startup_script=startup_script,
     boot_disk=compute.InstanceBootDiskArgs(
         initialize_params=compute.InstanceBootDiskInitializeParamsArgs(

### 2023-02-04 cloud/bootstrap_instance.py

@@ -10,6 +10,8 @@ parser.add_argument("username", type=str, default="ewheeler",
                     help="username on VM", nargs='?')
 parser.add_argument("pkey", type=str, default="/Users/ewheeler/.ssh/google_compute_engine",
                     help="local path to private key", nargs='?')
+#parser.add_argument("ip", type=str, default="34.123.171.183",
+#                    help="ip", nargs='?')
 args = parser.parse_args()
 
 username = args.username
@@ -18,10 +20,16 @@ pkey_path = args.pkey
 # find this script's directory
 dir_path = pathlib.Path(__file__).resolve().parent
 
+#if args.ip:
+#    hostname = args.ip
+#else:
+if True:
 # get IP address of new VM
-_hostname = subprocess.run(["pulumi", "stack", "output", "instanceIP"], capture_output=True)
-if not _hostname.stderr:
-    hostname = _hostname.stdout.decode().strip()
+    _hostname = subprocess.run(["pulumi", "stack", "output", "instanceIP"], capture_output=True)
+    if not _hostname.stderr:
+        hostname = _hostname.stdout.decode().strip()
+    else:
+        print(_hostname.stderr)
 print(hostname)
 
 pkey = paramiko.RSAKey.from_private_key_file(pkey_path)

### 2023-02-04 src/foresight/dags/__init__.py

@@ -1,8 +1,5 @@
 import os
 
-from dask.distributed import Client
-client = Client(os.getenv('DASK_SCHEDULER', None))
-
 from .io_dask import local_parquet_io_manager
 from .io_dask import gcp_parquet_io_manager
 

### 2023-02-04 src/foresight/dags/ingestion_dask.py

@@ -11,7 +11,6 @@ import urllib.request
 
 import numpy as np
 import pandas as pd
-import dask.dataframe as dd
 from dagster import Definitions
 from dagster import AssetSelection
 from dagster import MonthlyPartitionsDefinition
@@ -25,6 +24,13 @@ from dagster import ExpectationResult
 
 from dags import deployment_name
 from dags import resources
+
+import dask
+import dask.dataframe as dd
+from dask.distributed import Client
+
+client = Client('tcp://127.0.0.1:8786')
+dask.config.set(scheduler='distributed')
 
 
 # map of month name to month number

### 2023-04-21 docs/hes_final/article.qmd
update hes dos

---
title: "Predicting Conflict Escalation"
author:
  - name: Diana Kolusheva
    email: email@h
    affiliation: Harvard University
  - name: Chad Stoughton
    email: email@h
    affiliation: Harvard University
  - name: Evan Wheeler
    email: email@h
    affiliation: Harvard University
format: pdf
title-block-banner: true
abstract: |
  In 2021, UNICEF responded to 483 new and ongoing humanitarian situations in 153 countries (natural disasters, socio-political crises, health emergencies, nutrition crises, etc.) in addition to COVID-19 response efforts globally. Innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children. Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need. Predicting conflict has long been of interest to researchers and practitioners of international relations, and statistical and computational methods to do so can be found in the literature as early as 1944. Today, much of the work in this area has continued to focus on country or region-specific models to predict conflict.

---

## Introduction

## Conflict forecasting

Some more text.

## Methodology

## News media and conflict data
{{< embed figs.ipynb#fig-umap-3d >}}

## Results

### XGB

### Transformer

### Comparison with ...

## Conclusions

## Future work

### 2023-04-21 docs/hes_final/article.qmd
clean up hes docs

@@
 ## News media and conflict data
-{{< embed figs.ipynb#fig-umap-3d >}}
+
+### GDELT
+
+```{mermaid}
+%%| fig-width: 12
+graph TD
+    A[Conflict event] -->|reporter| B(Online news article)
+    B --> C(Scraped webpage)
+
+    subgraph GDELT
+        C --> |Google Chrome Compact Language Detector - CLD2|D(Language detection)
+        D --> |Conditional random field and various lang-specific|E(Word segmentation)
+        E --> F(Morphological analysis)
+        F --> G(Sentiment and emotional assessment)
+        G --> H(Sentence and clause segmentation)
+        H --> I(Low pass translation)
+        H --> J(Mid pass translation)
+        H --> K(Toponymic translation)
+        H --> L(Well known entity translation)
+        I --> M(Candidate translation ranking)
+        M --> O(Chosen translation)
+        J --> M
+        M --> |some langs|N(High pass translation)
+        N --> O
+        O --> |Moses and other grammer models|Q(User adaptation - dynamic language models)
+        Q --> R(High pass UNK substitution)
+        R --> S(Recapitalization models)
+        S --> T(English translation, events, entities, geocoding, and scores)
+        T --> |Universal Sentence Encoder| U(English article embeddings)
+        K --> T
+        L --> T
+    end
+
+    B --> |human and machine processes|CC(Judgement against ACLED event inclusion criteria)
+    subgraph ACLED
+        CC --> |human and machine processes|DD[Coded events, categories, event actors, geolocations, reported fatalities]
+        DD --> |three rounds of human review|DD
+        DD --> EE(Weekly aggregations of events and fatalities)
+    end
+
+    EE --> V(Timeseries dataset)
+    U --> V
+    V --> |Our model| W(Our conflict predictions)
+```
+
+### ACLED
 
@@
 
+![Precision-recall curve](imgs/prc.png){width=33%, fig-alt="Precision-recall curve."}
+
+```{mermaid}
+flowchart LR
+  A[Hard edge] --> B(Round edge)
+  B --> C{Decision}
+  C --> D[Result one]
+  C --> E[Result two]
+```
+
 ### Transformer
+
+![ROC](imgs/roc.png){width=33%, fig-alt="Reciever operating characteristic curve."}
 

### 2023-04-22 docs/hes_final/article.qmd

@@
 
+In 2021, UNICEF responded to 483 new and ongoing humanitarian situations in 153 countries (natural disasters, socio-political crises, health emergencies, nutrition crises, etc.) in addition to COVID-19 response efforts globally.
+This is a sharp increase compared to 2018, when UNICEF responded to 285 new and ongoing humanitarian emergencies in 90 countries.
+With the rising threat of the climate emergency and shifts toward authoritarianism, children around the world will be increasingly vulnerable to the impacts of catastrophe and conflict.
+Recent trends and global economic uncertainty call into question whether Individual, corporate, and government donor contributions will grow in step with these increased needs.
+
+According to UNICEF’s 2023 Humanitarian Action for Children publication,
+> “Conflict drives 80 per cent of all humanitarian needs.
+The number of countries experiencing violent conflict is at a thirty-year high.
+And, around the world, attacks on children continue unabated.
+From Ukraine to Yemen and from Ethiopia to Nigeria and beyond, warring parties are flouting one of the most basic rules of war: protect the children.
+> …
+> Altogether, nearly 37 million children worldwide are displaced due to conflict and violence, a number of displaced children not seen since the Second World War.
+And this number does not even include those children displaced by poverty or climate
+> change or by the search for a better life.
+Nor does it include those children displaced in 2022 due to the war in Ukraine, which itself has caused the fastest-growing refugee crisis in Europe since the Second World War.
+> …
+> lack of rigorous data and analysis of human-induced hazards remains challenging”
+
+Innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children.
+Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need.
+Specifically, such an automated approach will reduce efforts spent monitoring world events and allow for the initiation of emergency preparedness protocols sooner.
+
+Our approach to international conflict draws from both neorealism and constructivism.
+Neorealism assumes that state actors are the primary drivers of international conflict, and emphasizes the role of power and competition as drivers of conflict.
+This informs our approach in that we will be modeling conflict at the level of states.
+Constructivism emphasizes the role of behavioral norms, ideological factors, and identity in shaping international relations.
+Conflicts between peoples and groups are shaped by these factors.
+Journalists both respond to and play a role in shaping these beliefs and norms, implying that their work should correlate with rising or falling conflict.
+
 ## Conflict forecasting
@@
 ## News media and conflict data
+
+The [Global Database of Events Language and Tone (GDELT) Project](https://www.gdeltproject.org/) provides public APIs of current events derived from news articles scraped from sources around the world.
+With updates every fifteen minutes, GDELT’s Event Database documents physical events such as diplomatic interactions, social unrest, and military action.
+Approximately 60 attributes are assigned to each event from GDELT’s processing and analysis, including named entities involved, geolocation details, categorization among 300 event types, as well as metrics related to average ‘tone’ of articles about the event and potential severity of event impact.
+In 2021, GDELT introduced its Global Similarity Graph Document Embeddings dataset, which precomputes 512 dimension news-article-level embeddings using the Universal Sentence Encoder.
+
+The Armed Conflict Location & Event Data Project (ACLED)’s team of researchers monitor thousands of sources in 20 languages to generate a dataset with monthly totals of conflict events and their related fatalities.
+ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, or protests involving more than three people.
+ACLED’s monthly, per-country totals are generally published every two weeks.
+United Nations Office for the Coordination of Humanitarian Affairs’ (OCHA) Humanitarian Data Exchange platform hosts these regular updates.
+
+Our training data was derived from this GDELT and ACLED data.
+GDELT news article embeddings, averaged per-country and per-month, serve as our independent variable.
+Target variables for training were derived from ACLED’s conflict metrics.
+We derived a binary response variable representing conflict escalation three months into the future with respect to each month’s average article embeddings.
+Specifically, the binary response variable is calculated by TK.
+Our goal was to use GDELT data from a given time window to predict the ACLED metrics from a later time window.
 
@@
 flowchart LR
-  A[Hard edge] --> B(Round edge)
-  B --> C{Decision}
-  C --> D[Result one]
-  C --> E[Result two]
+    GDELT[GDELT Embeddings] -->|average by country by month| AGG(Monthly news representation)
+    Countries[Country labels] -->|one-hot encoding| AGG
+
+    AGG -->|PCA| PCA(169 dimension representation)
+    PCA --> KNN{KNN}
+    PCA --> XGD{XGBoost}
+
+    XGD --> AVG{Mean}
+    KNN --> AVG
+    AVG --> Pred(Prediction of conflict escalation in 3 months)
+
+    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
+    Spike(Conflict response binary variable) --> XGD
+    Spike --> KNN
 ```

### 2023-04-22 docs/hes_final/article.qmd
update draft

--- 2023-04-21 docs/hes_final/article.qmd
@@
     affiliation: Harvard University
+lang: en
 format: pdf
+classoption: twocolumn
+bibliography: bibliography.bibtex
+embed-resources: true
 title-block-banner: true
@@
 
+### Motivation
+In 2021, UNICEF responded to 483 new and ongoing humanitarian situations in 153 countries (natural disasters, socio-political crises, health emergencies, nutrition crises, etc.) in addition to COVID-19 response efforts globally.
+This is a sharp increase compared to 2018, when UNICEF responded to 285 new and ongoing humanitarian emergencies in 90 countries.
+With the rising threat of the climate emergency and shifts toward authoritarianism, children around the world will be increasingly vulnerable to the impacts of catastrophe and conflict.
+Recent trends and global economic uncertainty call into question whether Individual, corporate, and government donor contributions will grow in step with these increased needs.
+
+According to UNICEF’s 2023 Humanitarian Action for Children publication,
+
+> “Conflict drives 80 per cent of all humanitarian needs.
+The number of countries experiencing violent conflict is at a thirty-year high.
+And, around the world, attacks on children continue unabated.
+From Ukraine to Yemen and from Ethiopia to Nigeria and beyond, warring parties are flouting one of the most basic rules of war: protect the children.
+… Altogether, nearly 37 million children worldwide are displaced due to conflict and violence, a number of displaced children not seen since the Second World War.
+And this number does not even include those children displaced by poverty or climate change or by the search for a better life.
+Nor does it include those children displaced in 2022 due to the war in Ukraine, which itself has caused the fastest-growing refugee crisis in Europe since the Second World War.
+… lack of rigorous data and analysis of human-induced hazards remains challenging”[@unicef]
+
+Innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children.
+Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need.
+Specifically, such an automated approach will reduce efforts spent monitoring world events and allow for the initiation of emergency preparedness protocols sooner.
+
+### Objective
+
+Our practical objective was to develop a tool to support UNICEF in conflict identification and response.
+Our research objective was to demonstrate that a generalizable, global conflict prediction model is viable, and that such a model can be built using primarily public text data.
+
 ## Conflict forecasting
 
-Some more text.
+Predicting conflict has long been of interest to researchers and practitioners of international relations, and statistical and computational methods to do so can be found in the literature as early as 1944.
+Modern methodologies that resemble those we outline here began to appear in the literature in the early 2010s, when researchers began using holdout datasets for model validation [@10.2307/40730711].
+This shift followed work by Ward et al that sharply criticized the use of statistical significance tests on in-sample prediction to validate conflict models [@doi:10.1177/0022343309356491].
+Our methodology closely mirrors that which emerged in the following decade, including that described in more contemporary work, such as Mueller and Ruah [@RePEc:cam:camdae:2220].
+In essence, this methodology involves using time-lagged input data to predict some conflict label that corresponds to a date some time after the input data would have become available.
+The model is then validated using an unseen holdout set, typically at the end of the date range available in the data [@doi:10.1080/03050629.2022.2070745].
+This overcomes the problem of overfitting observed in earlier methods, and provides a principled starting point from which different models can be compared.
+
+Today, much of the work in this area has continued to focus on country or region-specific models to predict conflict.
+This introduces a “small-data problem” where relatively few observations of conflict are available for a given country [@RePEc:cam:camdae:2220].
+Some efforts have been made to overcome this by building generalizable models that can theoretically be used to make predictions on any country [@wen2023transformers].
+We followed the latter approach in developing our model, hypothesizing that the larger number of observations available to a global model will allow it to extract more generalizable signals of conflict, and that such a model will be better able to predict unforeseen risks in countries with little available training data.
+
+There are two principal ways in which our approach differs from previous efforts described in the literature.
+The first is our choice of model.
+Many earlier approaches, including those in O'Brien, used more traditional statistical methods, including logistic regression and actor simulations, to predict conflict.
+Contemporaneous work demonstrated that neural networks can outperform these traditional approaches [@10.2307/4145319].
+Despite this, neural networks do not seem to have been widely adopted in conflict prediction modeling – though there are some examples, such as in Chen et al [@DBLP:conf/sac/ChenJY20].
+Our work explores both approaches, and we find similar performance in some metrics, with major trade offs in other areas.
+
+The second principal difference in our approach is our choice of input data.
+While newspaper and media content has been used in conflict prediction models before, it has only been included as a supplement to models that primarily rely on other features.
+Despite this, models that include text features have been shown to outperform equivalent models that do not [@RePEc:cam:camdae:2220].
+In contrast, our approach istext-first.
+In other words, we made predictions primarily from text before adding other features.
+To achieve this, we relied on text embeddings of the news content.
+To the best of our knowledge, no conflict prediction model described in the literature makes use of text embeddings.
+Previous efforts have included statistical analysis of text, and unsupervised Latent Dirichlet Allocation (LDA) for topic extraction [@RePEc:cam:camdae:2220].
+These comparatively crude approaches to Natural Language Processing (NLP) have relegated text data to a peripheral role in conflict prediction.
+Our method uses text as the primary data source to make meaningful predictions.
+Such a model removes the need for the extensive expert surveys described in O’Brien, and overcomes the weaknesses of the strongly mean-regressing time-series models described in Mueller and Rauh.
+
+## News media and conflict data
+
+The [Global Database of Events Language and Tone (GDELT) Project](https://www.gdeltproject.org/) provides public APIs of current events derived from news articles scraped from sources around the world [@leetaru2013gdelt].
+With updates every fifteen minutes, GDELT’s Event Database documents physical events such as diplomatic interactions, social unrest, and military action.
+Approximately 60 attributes are assigned to each event from GDELT’s processing and analysis, including named entities involved, geolocation details, categorization among 300 event types, as well as metrics related to average ‘tone’ of articles about the event and potential severity of event impact.
+In 2021, GDELT introduced its Global Similarity Graph Document Embeddings dataset, which precomputes 512 dimension news-article-level embeddings using the Universal Sentence Encoder [@gdeltgsg].
+
+The GDELT Dataset has been used for a wide variety of research purposes, including for conflict prediction [@DBLP:conf/sac/ChenJY20].
+While we believed this to be the best dataset available for our purposes, prior research has revealed some limitations that we must consider.
+GDELT’s coverage overrepresents the western world, particularly the United States and Europe.
+GDELT does represent Russia, India and China better than other event datasets, but it continues to underrepresent other parts of the world, particularly the global south.
+English is the most common language in the GDELT dataset, comprising more than 40% of the articles in its corpus [@Kwak_An_2021].
+
+The [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/)’s team of researchers monitor thousands of sources in 20 languages to generate a dataset with monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914].
+ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and  protests.
+
+While many of GDELT’s databases reflect news articles from 1 January 1979 onward, GDELT’s news article embeddings exist only from 1 January 2020, limiting the time period for which training data can be drawn.
+Availability of ACLED’s monthly event and fatality totals varies by country, but event and fatality totals are available for all countries since at least 1 January 2020.
+
+Our training data was derived from these GDELT and ACLED data.
+GDELT news article embeddings serve as our independent variable.
+Target variables for training were derived from ACLED’s conflict metrics.
+Hard dates were set for train, test, and validation splits, and none of our models saw data published after October 2022 during training, with the remaining months used for testing and validation.
+
+```{mermaid}
+%%| fig-width: 3
+%%| fig-height: 5
+%%| fig-cap: Fig 1. Overview of data generation process for our conflict dataset 
+flowchart TD
+    GSG[GDELT Global Similarty Graph] -->|sample 20% of daily files| Merge{Merge on article URL,\n select relevant features}
+    GKG[GDELT Global Knowledge Graph] -->|sample 20% of daily files| Merge
+
+    Merge --> GDELT(GDELT dataset)
+
+    ACLED[ACLED monthly conflict totals] --> Spike{Calculate our lagged\n 'spike' binary variable}
+
+    GDELT --> Data(Time-partitioned train, val, & test datasets)
+    Spike --> Data
+```
+
+## Defining Conflict Escalation
+
+Our task of modeling conflict escalation required a precise definition of what conflict escalation is.
+The raw event and fatality counts provided by ACLED served as the basis for our conflict metric, however, to achieve our objective of creating a generalizable conflict model, we required a metric that was consistent across varying local conditions.
+Small changes in the absolute number of monthly fatalities can have wildly different implications depending on the country and time period.
+
+To address this, we derived a metric designed to highlight relative “spikes” in political violence.
+Calculating whether a spike occurred in a given month is a two-step process.
+First we calculated the trend in violence for a given month, defined by the slope of an ordinary least squares regression of the number of ACLED-documented fatalities that occurred in the target month, and the preceding two months.
+This trend allowed us to estimate whether violence was increasing or decreasing in the target country at any given time.
+Next, we calculated whether that slope was in the 75th percentile of such slopes in the target country for the preceding two years.
+This produced a binary metric that is sensitive to small increases in political violence in peaceful countries, while also filtering out small fluctuations in long-standing conflicts.
 
@@
 
-## News media and conflict data
-{{< embed figs.ipynb#fig-umap-3d >}}
+In order to push beyond prior attempts at conflict prediction, our approach to modeling drew from prior work in other fields.
+Our training data was structured such that each document is represented as an embedding vector, and multiple documents are combined with a label to create an aggregate picture of recent events.
+This is unique in that it is both a Natural Language Processing (NLP) task, and a time-series task, with each vector representing both semantic meaning, and an event or observation at a point in time.
+This lends itself to a variety of different approaches, and we have paid particular attention to the Transformer architecture.
+We reviewed recent work on Transformers for time-series data, and we hypothesized that the architecture’s ability to extract long-term dependencies, as well as its ability to find complex relationships between vector embeddings made it uniquely well-suited to this task [@wen2023transformers].
+
+We explored two approaches for getting conflict spike signals from the embeddings.
+In the first approach we used a fit-in-memory dataset of monthly embedding averages per country and applied statistical classification models such as K Nearest Neighbors and XGBoost.
+In the second approach we sampled article level embeddings and applied Transformer-based models.
+See the figures (X, Y) for details of both methods.
+
+### Statistical model using monthly embedding averages
+
+In this approach we prepared the dataset by averaging the embeddings per country per month.
+We also included one-hot encoded columns for countries.
+Since we only had access to embeddings starting in January 2020, the total length of this dataset was just a little over 6,000 samples (training, validation, and test sets combined).
+Given the high dimensionality of the data (512-dimensional embedding vectors + 700 dummy variables for countries), this dataset suffered from “curse of dimensionality”.
+This was handled by performing Principal Component Analysis – we found that reducing the number of dimensions to represent 80% of variance (~169 components with our selected train-test cutoff) improved the overall performance.
+Another issue was a class imbalance – conflict spike was present in only 14% of the training samples.
+We used random over-sampling technique to upsample the positive examples.
+
+The model used in this approach is a stacked model that averages probabilities independently predicted by a K Nearest Neighbors model and an XGBoost model.
+We used standard open-source versions of the underlying models.
+The intention behind averaging the predicted probabilities was to reduce the impact of individual model weaknesses.
+
+```{mermaid}
+%%| fig-width: 3
+%%| fig-height: 5
+%%| fig-cap: Fig 2. Statistical model architecture
+flowchart TD
+    GDELT[GDELT Embeddings] -->|average by country by month| AGG(Monthly news representation)
+    Countries[Country labels] -->|one-hot encoding| AGG
+
+    AGG -->|PCA| PCA(169 dimension representation)
+    PCA --> KNN{KNN}
+    PCA --> XGD{XGBoost}
+
+    XGD --> AVG{Mean}
+    KNN --> AVG
+    AVG --> Pred(Prediction of conflict escalation in 3 months)
+
+    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
+    Spike(Conflict response\n binary variable) --> XGD
+    Spike --> KNN
+```
+
+### Transformer Model using Resampled Embeddings
+
+Our second approach used a pair of transformer models to extract information directly from the GDELT embedding corpus.[@vaswani2017attention] Architecturally, both models are encoder-only transformers, but for simplicity we will refer to the first of these models as “the encoder” and the second as the “the classifier”.
+
+Prior to training, the GDELT dataset was split into one 512xn matrix for each country and month where n is the number of articles in the dataset published in that month related to that country.
+Each of these matrices was then split into several random subsamples of 512x50.
+Samples for countries in which n<50 were padded, and multiple samples were taken for countries in which n >50.
+This resampling dramatically increased the number of training samples by acting as though the same month had occurred multiple times while still maintaining uniqueness of each sample.
+Simultaneously, limiting the size of each sample to 50 articles reduced the dimensionality of each sample, reducing the risk of overfitting.
+
+The encoder was then trained using these samples, with the time-shifted spike metric as its target.
+Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons, and its weights were frozen.
+The encoder was then used to scan each of the original 512xn matrices, resulting in a length-16 vector for each 512x50 subsample.
+This reduced the dimensions of each sample to 16x(n/50).
+
+The classifier was then trained on this new, distilled dataset without resampling.
+The same time-shifted spike metric was again used as the training target.
+The reduced dimensionality of this dataset was sufficient to offset the risk of overfitting.
+
+```{mermaid}
+%%| fig-width: 3
+%%| fig-height: 5
+%%| fig-cap: Fig 3. Transformer model architecture
+flowchart TD
+    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)
+
+    Articles --> Trans1{Encoder\n transformer}
+    Trans1 -->|classifier head present during training| Compact(16 dimension representations)
+
+    Compact --> Trans2{Binary\n classifier\n transformer}
+    Trans2 --> Pred(Prediction of conflict escalation in 3 months)
+
+    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
+    Spike(Conflict response\n binary variable) --> Trans2
+
+    Spike -->|included during model training| Trans1
+```
 
@@
 
-### XGB
+We looked at the area under the ROC curve, F1 score, precision, recall and confusion matrix to design a holistic evaluation method at both fine-tuning/validation stage and final test set prediction stage.
+Despite conceptual differences in dataset creation and modeling techniques, both of the described above approaches produced comparative results.
 
-### Transformer
+TODO: Table (for numerical metrics) and plots (for roc_auc) for 2 models
+
+Further investigation showed that while overall numeric metrics look similar, the Transformer-based model generalized better across the globe.
+
+TODO: map comparison images
+
+While performance is often the main objective to optimize for when developing prediction models, one of our goals was to develop a model that can assist UNICEF in their decision making.
+With that in mind, interpretability and cost of deployment are important factors for model evaluation.
 
@@
 ## Future work
+
+## References
+
+::: {#refs}
+:::

### 2023-04-25 docs/hes_final/article.qmd
get tables to work in two column article

@@
 lang: en
-format: pdf
-classoption: twocolumn
+format:
+  pdf:
+    classoption:
+    - twocolumn
 bibliography: bibliography.bibtex
+include-in-header:
+  - text: |
+      \usepackage{float}
+      \makeatletter
+      \let\oldlt\longtable
+      \let\endoldlt\endlongtable
+      \def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
+      \def\longtable@i[#1]{\begin{figure}[H]
+      \onecolumn
+      \begin{minipage}{0.5\textwidth}
+      \oldlt[#1]
+      }
+      \def\longtable@ii{\begin{figure}[H]
+      \onecolumn
+      \begin{minipage}{0.5\textwidth}
+      \oldlt
+      }
+      \def\endlongtable{\endoldlt
+      \end{minipage}
+      \twocolumn
+      \end{figure}}
+      \makeatother
 embed-resources: true
@@
 
-Prior to training, the GDELT dataset was split into one 512xn matrix for each country and month where n is the number of articles in the dataset published in that month related to that country.
-Each of these matrices was then split into several random subsamples of 512x50.
-Samples for countries in which n<50 were padded, and multiple samples were taken for countries in which n >50.
+Prior to training, the GDELT dataset was split into one $512 x n$ matrix for each country and month where n is the number of articles in the dataset published in that month related to that country.
+Each of these matrices was then split into several random subsamples of $512 x 50$.
+Samples for countries in which $n < 50$ were padded, and multiple samples were taken for countries in which $n > 50$.
 This resampling dramatically increased the number of training samples by acting as though the same month had occurred multiple times while still maintaining uniqueness of each sample.
@@
 Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons, and its weights were frozen.
-The encoder was then used to scan each of the original 512xn matrices, resulting in a length-16 vector for each 512x50 subsample.
-This reduced the dimensions of each sample to 16x(n/50).
+The encoder was then used to scan each of the original $512 x n$ matrices, resulting in a length-16 vector for each $512 x 50$ subsample.
+This reduced the dimensions of each sample to $16 x (n/50)$.
 
@@
 TODO: Table (for numerical metrics) and plots (for roc_auc) for 2 models
+```{python}
+#| echo: false
+#| label: tbl-planets
+#| tbl-cap: Planets
+
+from IPython.display import Markdown
+from tabulate import tabulate
+table = [["Sun",696000,1989100000],
+         ["Earth",6371,5973.6],
+         ["Moon",1737,73.5],
+         ["Mars",3390,641.85]]
+Markdown(tabulate(
+  table,
+  headers=["Planet","R (km)", "mass (x 10^29 kg)"]
+))
+```
 

### 2023-04-25 docs/hes_final/article.qmd
transformer section rewrite

--- 2023-04-22 docs/hes_final/article.qmd
@@
 
-Our second approach used a pair of transformer models to extract information directly from the GDELT embedding corpus.[@vaswani2017attention] Architecturally, both models are encoder-only transformers, but for simplicity we will refer to the first of these models as “the encoder” and the second as the “the classifier”.
+Our second approach used a pair of transformer models to extract information directly from the GDELT embedding corpus.[@vaswani2017attention] Architecturally, both models are encoder-only transformers, but for simplicity, we will refer to the first of these models as “the encoder” and the second as the “classifier”.
 
-Prior to training, the GDELT dataset was split into one 512xn matrix for each country and month where n is the number of articles in the dataset published in that month related to that country.
-Each of these matrices was then split into several random subsamples of 512x50.
-Samples for countries in which n<50 were padded, and multiple samples were taken for countries in which n >50.
-This resampling dramatically increased the number of training samples by acting as though the same month had occurred multiple times while still maintaining uniqueness of each sample.
-Simultaneously, limiting the size of each sample to 50 articles reduced the dimensionality of each sample, reducing the risk of overfitting.
+The purpose of the encoder was to reduce the dimensionality of an embedding corpus in a manner analogous to a convolutional layer in a convolutional neural network.
+The classifier could then be used to make final predictions on a distilled dataset.
 
-The encoder was then trained using these samples, with the time-shifted spike metric as its target.
-Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons, and its weights were frozen.
-The encoder was then used to scan each of the original 512xn matrices, resulting in a length-16 vector for each 512x50 subsample.
-This reduced the dimensions of each sample to 16x(n/50).
+Prior to training, a base dataset was created by splitting the GDELT dataset into one $512xn$ matrix for each country and month where $n$ is the number of articles in the dataset published in that month related to that country.
+Each of these matrices was then split into several random subsamples of $512x50$.
+Samples for countries in which $n<50$ were padded.
+This process had two effects on training.
+First, by limiting the size of each sample to 50 articles, we reduced the dimensionality of each sample.
+Simultaneously, by reusing the same training label for multiple unique training samples, we were able to treat each month as though it had occurred multiple times.
+This greatly increased the size of the training dataset and helped overcome the limitations imposed by the limited time window available in the GDELT embedding corpus.
 
-The classifier was then trained on this new, distilled dataset without resampling.
-The same time-shifted spike metric was again used as the training target.
-The reduced dimensionality of this dataset was sufficient to offset the risk of overfitting.
+This new dataset represents a distorted view of the world, which overrepresents countries that receive more media coverage and underrepresents those that receive less.
+Despite this limitation, this dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target.
+Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons.
+
+With the encoder's training complete, we returned to the original base dataset.
+Each country-month matrix was again split into $512x50$ matrices, this time without resampling.
+Each of these samples could then be passed into the encoder, which distilled it into a $length-16$ vector.
+Each vector was then stacked, resulting in a single matrix for each country and month of size $16x\frac{n}{50}$.
+This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target.
 

### 2023-04-26 docs/hes_final/article.qmd
add config to article for less aggressive line wrap hyphenation

@@
 bibliography: bibliography.bibtex
+# https://github.com/quarto-dev/quarto-cli/discussions/2786
 include-in-header:
@@
       \makeatother
+# https://tex.stackexchange.com/questions/31301/how-to-reduce-the-number-of-hyphenation
+include-before-body:
+  - text: |
+      \pretolerance=5000
+      \tolerance=9000
+      \emergencystretch=0pt
+      \exhyphenpenalty=100
+      \righthyphenmin=4
+      \lefthyphenmin=5
 embed-resources: true

### 2023-04-26 docs/hes_final/article.qmd
add notebook for figures and examples of cell embeds

@@
 embed-resources: true
+notebook-links: false
 title-block-banner: true
@@
 
+```{python}
+#| echo: false
+#| label: tbl-planets
+#| tbl-cap: Planets
+
+from IPython.display import Markdown
+from tabulate import tabulate
+table = [["Sun",696000,1989100000],
+         ["Earth",6371,5973.6],
+         ["Moon",1737,73.5],
+         ["Mars",3390,641.85]]
+Markdown(tabulate(
+  table,
+  headers=["Planet","R (km)", "mass (x 10^29 kg)"]
+))
+```
 ## Defining Conflict Escalation
@@
 TODO: Table (for numerical metrics) and plots (for roc_auc) for 2 models
-```{python}
-#| echo: false
-#| label: tbl-planets
-#| tbl-cap: Planets
 
-from IPython.display import Markdown
-from tabulate import tabulate
-table = [["Sun",696000,1989100000],
-         ["Earth",6371,5973.6],
-         ["Moon",1737,73.5],
-         ["Mars",3390,641.85]]
-Markdown(tabulate(
-  table,
-  headers=["Planet","R (km)", "mass (x 10^29 kg)"]
-))
-```
+{{< embed figs.ipynb#tbl-metrics >}}
 
 Further investigation showed that while overall numeric metrics look similar, the Transformer-based model generalized better across the globe.
+
+{{< embed figs.ipynb#fig-roc >}}
 

### 2023-04-28 docs/hes_final/article.qmd
add precision recall curve plot to article
separate metrics where max is best vs min
Extend results section

@@
 
-{{< embed figs.ipynb#tbl-metrics >}}
+{{< embed figs.ipynb#tbl-best-max-metrics >}}
+
+{{< embed figs.ipynb#tbl-best-min-metrics >}}
 
 Further investigation showed that while overall numeric metrics look similar, the Transformer-based model generalized better across the globe.
+Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the
+generalizability is a significant reason to prefer Transformer architecture and resampled dataset.
 
 {{< embed figs.ipynb#fig-roc >}}
+
+{{< embed figs.ipynb#fig-prc >}}
 
@@
 While performance is often the main objective to optimize for when developing prediction models, one of our goals was to develop a model that can assist UNICEF in their decision making.
-With that in mind, interpretability and cost of deployment are important factors for model evaluation.
+With that in mind, cost of deployment is also an important factor for model evaluation.
+Average embedding dataset fits in memory while resampling article level embeddings requires additional data engineering
+steps to be processed in batches. Both dataset size and corresponding model size make average embeddings based statistical
+model more attractive from this perspective.
+
+Overall, the results suggest that it is possible to predict upcoming conflict escalation from text embeddings of news articles. 
+Different data sampling and modeling approaches produce comparable results based on common metrics.
+Transformer-based model generalizes better across geographies while statistical model provides a size/cost of deployment advantage.
 

### 2023-04-28 docs/hes_final/article.qmd
Update results and add conclusions

@@
 
+We evaluated our models with both a research and a practical objectives in mind.
+
+From a research perspective we were interested in feasibility of making accurate and generalizable predictions of
+conflict escalation using publicly available text embeddings.
 We looked at the area under the ROC curve, F1 score, precision, recall and confusion matrix to design a holistic evaluation method at both fine-tuning/validation stage and final test set prediction stage.
@@
 
-While performance is often the main objective to optimize for when developing prediction models, one of our goals was to develop a model that can assist UNICEF in their decision making.
+While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making.
 With that in mind, cost of deployment is also an important factor for model evaluation.
@@
 
-Overall, the results suggest that it is possible to predict upcoming conflict escalation from text embeddings of news articles. 
-Different data sampling and modeling approaches produce comparable results based on common metrics.
-Transformer-based model generalizes better across geographies while statistical model provides a size/cost of deployment advantage.
+We think that both approaches produced promising results and the model choice depends on the developers' specific goals and available resources.
 
@@
 ## Conclusions
+
+In this work we explored the possibility of predicting future geo-political conflict escalation at a country level
+using a global collection of news publications. We explored different ways of aggregating the data and
+applied both statistical machine learning and deep learning transformer architecture models.
+The results suggest that text embeddings provide a strong enough signal to make such predictions.
+Different data sampling and modeling approaches produce comparable results based on common metrics. 
+Transformer-based model generalizes better across geographies while statistical model provides a size/cost of deployment advantage.
 

### 2023-04-28 docs/hes_final/article.qmd
Add future directions

@@
 
+We think there are many ways to extend this work. We can group them to align with our initial research and
+practical objectives.
+
+### Future research directions
+
+One of the potential research extensions of this work is to continue experimenting
+with data sampling and aggregation approaches. We used either monthly averages 
+to create $512x1$ vectors or article level samples to create $512x50$ matrices
+in our dataset preparation. While we received some promising results, it is possible 
+that the first approach makes the data too coarse and the second too noisy. Another
+alternative could be generating daily averages and combining them into $512x31$ (largest number of days in a month)
+matrices. This could potentially be more sensitive to the signal than the monthly averages
+and less noisy than article level samples.
+
+Another direction is dimensionality reduction technique. We observed a clear separation of clusters using a
+UMAP technique supervised by the Spike target variable. Using the dataset reduced with this technique
+could be a better alternative to a principal component analysis our statistical model uses.
+
+On the transormers side, the output of the encoder also represented a reduced dimension dataset
+further fed into the second transformer, the classifier. Using the output of the encoder as input to other
+classifiers could be another direction of research.
+
+Finally, due to time and resource constraints in this project we used 20% subset
+of daily news articles. Increasing that percentage could potentially lead to better results.
+
+### Future practical directions
+
+Building an application with a user interface could be a natural extension of this work.
+We envision a dashboard where a user can view a world map where countries are heatmap colored
+according to the estimated probability of escalating conflict. Countries with highest risk
+can be flagged or shown to a user in a separate view. In addition to the visual part
+of this application it is important to ensure that the model is deployed and
+can receive new data input and make inference.
+
+Building a more robust data processing pipeline is another important direction. This would involve
+designing either a batch or streaming system that can reliably collect data, preprocess
+and aggregate it into required format, train the model and make predictions. The system should handle 
+computational resources and cost constraints.
+
 ## References

### 2023-04-28 src/foresight/Analysis/Model_Anlysis.py
More Model Analysis Code

@@ -2,6 +2,7 @@ import pandas as pd
 from google.cloud import storage
 from sklearn.linear_model import LogisticRegression
 import numpy as np
+from sklearn.metrics import f1_socre, roc_curve, roc_auc_score
 import matplotlib.pyplot as plt
 import matplotlib.patches as mpatches
 from matplotlib.lines import Line2D
@@ -14,18 +15,33 @@ metadata = pd.read_csv('gcs://frsght/datasets_stacked/metadata.csv')
 ACLED = pd.read_csv('gcs://frsght/acled_labels/ACLED_Labels_032723.csv')
 transformer_preds = pd.read_csv('gcs://frsght/model_predictions/chad_preds.csv')
 xgb_preds = pd.read_csv('gcs://frsght/model_predictions/diana_preds.csv')
+iso_table = pd.read_html('https://www.worlddata.info/countrycodes.php')[0]
+
+iso2 = pd.Series(iso_table['ISO 3166-1 alpha3'].values, index=iso_table['Fips 10']).to_dict()
+iso3 = pd.Series(iso_table['ISO 3166-1 alpha2'].values, index=iso_table['Fips 10']).to_dict()
 
 metadata = metadata.drop_duplicates()
 transformer_preds = transformer_preds.drop_duplicates()
 metadata['country_ym'] =  metadata['yearmonth'].astype(str) + '_' + metadata['country']
 
 xgb_preds['country_ym'] = xgb_preds['target_window'].str[0:4]+xgb_preds['target_window'].str[5:]+"_"+xgb_preds['fips']
+xgb_preds['xgb_correct'] = xgb_preds['lagged_spike'].astype(int) == xgb_preds['prediction']
 
 transformer_preds['tf_correct'] = transformer_preds['Spike'] == transformer_preds['class_preds']
 
 preds = transformer_preds.merge(metadata, how = 'left', on = 'country_ym')
 
 preds = preds.merge(xgb_preds, on = 'country_ym', how = 'inner')
+
+tf_metrics = transformer_preds.groupby('Code').mean()
+tf_metrics['iso2'] = tf_metrics.index.map(iso2)
+tf_metrics['iso3'] = tf_metrics.index.map(iso3)
+tf_metrics = tf_metrics[['tf_correct', 'iso2', 'iso3']]
+xgb_metrics = xgb_preds.groupby('fips').mean()
+xgb_metrics['iso2'] = xgb_metrics.index.map(iso2)
[…]

### 2023-04-28 docs/hes_final/article.qmd
Defining Conflict Escalation Minor Rewrite to acknowledge prior work

@@
 Our task of modeling conflict escalation required a precise definition of what conflict escalation is.
-The raw event and fatality counts provided by ACLED served as the basis for our conflict metric, however, to achieve our objective of creating a generalizable conflict model, we required a metric that was consistent across varying local conditions.
-Small changes in the absolute number of monthly fatalities can have wildly different implications depending on the country and time period.
+Prior work has treated this as a binary classification task, asking whether or not conflict escalated in a given month, and we adopt this approach as well.
+[@10.1093/jeea/jvac025]
+
+In contrast to prior work, however, we do not treat the simple presence of a fatality due to conflict as sufficient to indicate conflict escalation.
+To achieve our objective of creating a generalizable conflict model, we required a metric that was consistent across varying local conditions.
+Small changes in the absolute number of monthly fatalities can have widely different implications depending on the country and time period.
 
@@
 Calculating whether a spike occurred in a given month is a two-step process.
-First we calculated the trend in violence for a given month, defined by the slope of an ordinary least squares regression of the number of ACLED-documented fatalities that occurred in the target month, and the preceding two months.
+First, we calculated the trend in violence for a given month, defined by the slope of an ordinary least squares regression of the number of ACLED-documented fatalities that occurred in the target month, and the preceding two months.
 This trend allowed us to estimate whether violence was increasing or decreasing in the target country at any given time.

### 2023-04-29 docs/hes_final/article.qmd
Language Pass and Methods Rework

@@
 
-There are two principal ways in which our approach differs from previous efforts described in the literature.
-The first is our choice of model.
-Many earlier approaches, including those in O'Brien, used more traditional statistical methods, including logistic regression and actor simulations, to predict conflict.
-Contemporaneous work demonstrated that neural networks can outperform these traditional approaches [@10.2307/4145319].
-Despite this, neural networks do not seem to have been widely adopted in conflict prediction modeling – though there are some examples, such as in Chen et al [@DBLP:conf/sac/ChenJY20].
-Our work explores both approaches, and we find similar performance in some metrics, with major trade offs in other areas.
-
-The second principal difference in our approach is our choice of input data.
-While newspaper and media content has been used in conflict prediction models before, it has only been included as a supplement to models that primarily rely on other features.
+## Methodology
+The principal difference between our approach and that of prior work is in our choice of predictive features, which we extract directly from text embeddings of publicly available news coverage.
+While newspaper and media content has been used in conflict prediction models before, it is generally only included as a supplement to models that primarily rely on other features.
 Despite this, models that include text features have been shown to outperform equivalent models that do not [@RePEc:cam:camdae:2220].
-In contrast, our approach istext-first.
-In other words, we made predictions primarily from text before adding other features.
+In contrast, our approach is text-first, making predictions primarily or exclusively from text, with few or no other features added.
 To achieve this, we relied on text embeddings of the news content.
@@
 Previous efforts have included statistical analysis of text, and unsupervised Latent Dirichlet Allocation (LDA) for topic extraction [@RePEc:cam:camdae:2220].
-These comparatively crude approaches to Natural Language Processing (NLP) have relegated text data to a peripheral role in conflict prediction.
+These approaches to Natural Language Processing (NLP) have relegated text data to a peripheral role in conflict prediction.
 Our method uses text as the primary data source to make meaningful predictions.
@@
 
-## News media and conflict data
+In order to elevate text data to the center-stage of our prediction efforts, we drew from prior work in other fields.
+Our training data was structured such that each document is represented as an embedding vector, and multiple documents are combined with a label to create an aggregate picture of recent events.
+We explored two approaches for extracting conflict signals from text embeddings.
+In the first approach we used a fit-in-memory dataset of monthly embedding averages per country per month and applied statistical classification models such as K Nearest Neighbors and XGBoost.
+In the second approach we stacked multiple article-level embeddings and applied Transformer-based models to extract a signal directly from the text corpus.
+See the figures (X, Y) for details of both methods.
+
+### Defining Conflict Escalation
+Our task of modeling conflict escalation required a precise definition of what conflict escalation is.
+We began with raw event and casualty counts from The [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/) , whose team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914].
+ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and protests.
+With these totals in-hand, we needed to create a target to label our training data.
+Prior work has treated this as a binary classification task, asking whether or not conflict escalated in a given month, and we adopt this approach as well.
+[@10.1093/jeea/jvac025]
+In contrast to prior work, however, we do not treat the simple presence of a fatality due to conflict as sufficient to indicate conflict escalation.
+To achieve our objective of creating a generalizable conflict model, we required a metric that was consistent across varying local conditions.
+Small changes in the absolute number of monthly fatalities can have widely different implications depending on the country and time period.
+To address this, we derived a metric designed to highlight relative “spikes” in political violence.
+Calculating whether a spike occurred in a given month is a two-step process.
+First, we calculated the trend in violence for a given month, defined by the slope of an ordinary least squares regression of the number of ACLED-documented fatalities that occurred in the target month, and the preceding two months.
+This trend allowed us to estimate whether violence was increasing or decreasing in the target country at any given time.
+Next, we calculated whether that slope was in the $75^{th}$ percentile of such slopes in the target country for the preceding two years.
+This produced a binary metric that is sensitive to small increases in political violence in peaceful countries, while also filtering out small fluctuations in long-standing conflicts.
+
+### News Content Data
 
@@
 
-The [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/)’s team of researchers monitor thousands of sources in 20 languages to generate a dataset with monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914].
-ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and  protests.
-
 While many of GDELT’s databases reflect news articles from 1 January 1979 onward, GDELT’s news article embeddings exist only from 1 January 2020, limiting the time period for which training data can be drawn.
 Availability of ACLED’s monthly event and fatality totals varies by country, but event and fatality totals are available for all countries since at least 1 January 2020.
-
 Our training data was derived from these GDELT and ACLED data.
-GDELT news article embeddings serve as our independent variable.
-Target variables for training were derived from ACLED’s conflict metrics.
+GDELT news article embeddings served as our independent variable, while the “spike” metric derived from ACLED’s conflict data served as our target variable.
 Hard dates were set for train, test, and validation splits, and none of our models saw data published after October 2022 during training, with the remaining months used for testing and validation.
@@
 ```
-
-```{python}
-#| echo: false
-#| label: tbl-planets
-#| tbl-cap: Planets
-
-from IPython.display import Markdown
-from tabulate import tabulate
-table = [["Sun",696000,1989100000],
-         ["Earth",6371,5973.6],
-         ["Moon",1737,73.5],
-         ["Mars",3390,641.85]]
-Markdown(tabulate(
-  table,
-  headers=["Planet","R (km)", "mass (x 10^29 kg)"]
-))
-```
-## Defining Conflict Escalation
-
-Our task of modeling conflict escalation required a precise definition of what conflict escalation is.
-Prior work has treated this as a binary classification task, asking whether or not conflict escalated in a given month, and we adopt this approach as well.
-[@10.1093/jeea/jvac025]
-
-In contrast to prior work, however, we do not treat the simple presence of a fatality due to conflict as sufficient to indicate conflict escalation.
-To achieve our objective of creating a generalizable conflict model, we required a metric that was consistent across varying local conditions.
-Small changes in the absolute number of monthly fatalities can have widely different implications depending on the country and time period.
-
-To address this, we derived a metric designed to highlight relative “spikes” in political violence.
-Calculating whether a spike occurred in a given month is a two-step process.
-First, we calculated the trend in violence for a given month, defined by the slope of an ordinary least squares regression of the number of ACLED-documented fatalities that occurred in the target month, and the preceding two months.
-This trend allowed us to estimate whether violence was increasing or decreasing in the target country at any given time.
-Next, we calculated whether that slope was in the 75th percentile of such slopes in the target country for the preceding two years.
-This produced a binary metric that is sensitive to small increases in political violence in peaceful countries, while also filtering out small fluctuations in long-standing conflicts.
-
-## Methodology
-
-In order to push beyond prior attempts at conflict prediction, our approach to modeling drew from prior work in other fields.
-Our training data was structured such that each document is represented as an embedding vector, and multiple documents are combined with a label to create an aggregate picture of recent events.
-This is unique in that it is both a Natural Language Processing (NLP) task, and a time-series task, with each vector representing both semantic meaning, and an event or observation at a point in time.
-This lends itself to a variety of different approaches, and we have paid particular attention to the Transformer architecture.
-We reviewed recent work on Transformers for time-series data, and we hypothesized that the architecture’s ability to extract long-term dependencies, as well as its ability to find complex relationships between vector embeddings made it uniquely well-suited to this task [@wen2023transformers].
-
-We explored two approaches for getting conflict spike signals from the embeddings.
-In the first approach we used a fit-in-memory dataset of monthly embedding averages per country and applied statistical classification models such as K Nearest Neighbors and XGBoost.
-In the second approach we sampled article level embeddings and applied Transformer-based models.
-See the figures (X, Y) for details of both methods.
 
@@
 
-The purpose of the encoder was to reduce the dimensionality of an embedding corpus in a manner analogous to a convolutional layer in a convolutional neural network.
-The classifier could then be used to make final predictions on a distilled dataset.
+The purpose of the encoder was to reduce the dimensionality of the embedding corpus by analyzing a subset of article embeddings and returning a single vector of substantially reduced size.
+This is roughly analogous to the role of a convolutional layer in a CNN.
+The classifier could then be used to make final predictions on the resulting distilled dataset.
 
-Prior to training, a base dataset was created by splitting the GDELT dataset into one $512xn$ matrix for each country and month where $n$ is the number of articles in the dataset published in that month related to that country.
-Each of these matrices was then split into several random subsamples of $512x50$.
-Samples for countries in which $n<50$ were padded.
+To prepare the training dataset, our full sample of the GDELT corpus was split into a single $512xn$ matrix for each country and month where $n$ is the number of articles published in that month related to that country.
+Each of these matrices was then split into several random subsamples of 50 articles each.
+Samples for countries in which $n<50$ were padded to $512x50$.
 This process had two effects on training.
 First, by limiting the size of each sample to 50 articles, we reduced the dimensionality of each sample.
-Simultaneously, by reusing the same training label for multiple unique training samples, we were able to treat each month as though it had occurred multiple times.
-This greatly increased the size of the training dataset and helped overcome the limitations imposed by the limited time window available in the GDELT embedding corpus.
+Simultaneously, by splitting the corpus into multiple unique training examples per country per month, we were able to substantially increase the size of the training dataset.
 
-This new dataset represents a distorted view of the world, which overrepresents countries that receive more media coverage and underrepresents those that receive less.
+This new dataset offered major advantages in combatting the curse of dimensionality, however, it also represented a distorted view of the world, which overrepresents countries that receive more media coverage and underrepresents those that receive les, amplifying the biases already present in the GDELT dataset.
 Despite this limitation, this dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target.
 Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons.
+The output of these 16 neurons would be used in the classification stage.
 
 With the encoder's training complete, we returned to the original base dataset.
-Each country-month matrix was again split into $512x50$ matrices, this time without resampling.
+Each country-month matrix was again split into $512x50$ matrices.
 Each of these samples could then be passed into the encoder, which distilled it into a $length-16$ vector.
-Each vector was then stacked, resulting in a single matrix for each country and month of size $16x\frac{n}{50}$.
+Each vector was then stacked, resulting in a single training example for each country and month of size $16x\frac{n}{50}$.
 This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target.
+In contrast to our statistical modeling approach, the transformer models are trained purely on text embeddings.
+The model was not conditioned by country, and no additional features were added beyond the text corpus itself.
 
@@
 
-{{< embed figs.ipynb#tbl-best-max-metrics >}}
-
-{{< embed figs.ipynb#tbl-best-min-metrics >}}
-
 Further investigation showed that while overall numeric metrics look similar, the Transformer-based model generalized better across the globe.
@@
 generalizability is a significant reason to prefer Transformer architecture and resampled dataset.
-
-{{< embed figs.ipynb#fig-roc >}}
-
-{{< embed figs.ipynb#fig-prc >}}
 
@@
 using a global collection of news publications. We explored different ways of aggregating the data and
-applied both statistical machine learning and deep learning transformer architecture models.
+applied both statistical machine learning and deep learning models.
 The results suggest that text embeddings provide a strong enough signal to make such predictions.
@@
 
-We think there are many ways to extend this work. We can group them to align with our initial research and
-practical objectives.
+We foresee several ways to extend this work, and there are exciting opportunities for both further research and operationalizations.
 
@@
 
-On the transormers side, the output of the encoder also represented a reduced dimension dataset
-further fed into the second transformer, the classifier. Using the output of the encoder as input to other
-classifiers could be another direction of research.
+The transformer models have a great deal of room for hyperparameter tuning.
+Further experimentation with the size of the encoder output, the number of articles per training sample, and implementing overlapping strides during distillation could all improve the downstream performance of the classifier.
+It may also be possible to use the output of the encoder as input to some other classification model architecture.
 
@@
 
-Building a more robust data processing pipeline is another important direction. This would involve
+Building a more robust and automated data processing pipeline is another important direction. This would involve
 designing either a batch or streaming system that can reliably collect data, preprocess

### 2023-04-29 docs/hes_final/article.qmd
embed images written by figs.ipynb

@@
 
-TODO: Table (for numerical metrics) and plots (for roc_auc) for 2 models
+![Model performance metrics](imgs/tbl-best-max-metrics.png){width=50%}
+
+![Model performance metrics](imgs/tbl-best-min-metrics.png){width=50%}
+
+![Receiver Operator Characteristics](imgs/fig-roc.png){width=50%}
+
+![Precision Recall Curve](imgs/fig-prc.png){width=50%}
 
@@
 
-TODO: map comparison images
+![Predictions for 2023-01 from transformer model](imgs/fig-y_pred_proba-transformer-202301.png){width=50%}
+
+![Predictions for 2023-01 from xgboost model](imgs/fig-y_pred_proba-xgboost-202301.png){width=50%}
+
+![Predictions for 2023-01 from ffnn model](imgs/fig-y_pred_proba-ffnn-202301.png){width=50%}
 

### 2023-04-29 docs/hes_final/article.qmd
add umap figures and ffnn description

@@
 
+## Feed Forward Neural Network
+
+Our third approach employed a feed forward neural network model.
+Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner.
+Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space.
+The fit UMAP model, was applied to the validation and test data without supervision.
+These reduced, 128 dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons.
+The model's 4577 parameters were trained for 3 epochs.
+
+![Distribution of supervised UMAP reduction to 2 dimensions](imgs/fig-umap2d-jointplot.png){width=50%}
+
+![Scatterplot of 2d UMAP reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){width=50%}
+
+Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total') and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges, 4 categories of gross national income per capita, and 3 categories of Fragile States Index ranges).
+
+```{mermaid}
+%%| fig-width: 3
+%%| fig-height: 5
+%%| fig-cap: Fig 6. Feed forward neural network model architecture
+flowchart TD
+    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)
+
+    Articles --> UMAP{Supervised UMAP}
+    UMAP --> Compact(128 dimension representations)
+
+    Compact --> FFNN{Feed Forward Neural Network}
+    FFNN --> Pred(Prediction of conflict escalation in 3 months)
+
+    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
+    Spike(Conflict response binary variable) --> |included during training|UMAP
+
+    Spike -->FFNN
+
+    Context[Country context indicators] --> FFNN
+```
+
 ## Results
@@
 
-Another direction is dimensionality reduction technique. We observed a clear separation of clusters using a
-UMAP technique supervised by the Spike target variable. Using the dataset reduced with this technique
-could be a better alternative to a principal component analysis our statistical model uses.
-
 The transformer models have a great deal of room for hyperparameter tuning.

### 2023-04-30 docs/hes_final/article.qmd
Formatting and language edits

@@
 Despite this, models that include text features have been shown to outperform equivalent models that do not [@RePEc:cam:camdae:2220].
+
 In contrast, our approach is text-first, making predictions primarily or exclusively from text, with few or no other features added.
-To achieve this, we relied on text embeddings of the news content.
 To the best of our knowledge, no conflict prediction model described in the literature makes use of text embeddings.
@@
 These approaches to Natural Language Processing (NLP) have relegated text data to a peripheral role in conflict prediction.
-Our method uses text as the primary data source to make meaningful predictions.
-Such a model removes the need for the extensive expert surveys described in O’Brien, and overcomes the weaknesses of the strongly mean-regressing time-series models described in Mueller and Rauh.
 
@@
 Our training data was structured such that each document is represented as an embedding vector, and multiple documents are combined with a label to create an aggregate picture of recent events.
-We explored two approaches for extracting conflict signals from text embeddings.
+
+We developed two approaches for extracting conflict signals from text embeddings.
 In the first approach we used a fit-in-memory dataset of monthly embedding averages per country per month and applied statistical classification models such as K Nearest Neighbors and XGBoost.
@@
 [@10.1093/jeea/jvac025]
+
 In contrast to prior work, however, we do not treat the simple presence of a fatality due to conflict as sufficient to indicate conflict escalation.
@@
 Small changes in the absolute number of monthly fatalities can have widely different implications depending on the country and time period.
+
 To address this, we derived a metric designed to highlight relative “spikes” in political violence.

### 2023-04-30 docs/hes_final/article.qmd
update test prediction map figures

--- 2023-04-29 docs/hes_final/article.qmd
@@
 
-![Predictions for 2023-01 from transformer model](imgs/fig-y_pred_proba-transformer-202301.png){width=50%}
+![Transformer model Predictions on test data](imgs/fig-transformer-test-predictions.png){width=55%}
 
-![Predictions for 2023-01 from xgboost model](imgs/fig-y_pred_proba-xgboost-202301.png){width=50%}
+![XGBoost model predictions on test data](imgs/fig-xgboost-test-predictions.png){width=55%}
 
-![Predictions for 2023-01 from ffnn model](imgs/fig-y_pred_proba-ffnn-202301.png){width=50%}
+![FFNN model predictions on test data](imgs/fig-ffnn-test-predictions.png){width=55%}
 

### 2023-04-30 docs/hes_final/article.qmd
add citation for UMAP

--- 2023-04-30 docs/hes_final/article.qmd
@@
 Our task of modeling conflict escalation required a precise definition of what conflict escalation is.
-We began with raw event and casualty counts from The [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/) , whose team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914].
+We began with raw event and casualty counts from The [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/), whose team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914].
 ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and protests.
@@
 Our third approach employed a feed forward neural network model.
-Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner.
+Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner [@sainburg2021parametric].
 Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space.

### 2023-04-30 docs/hes_final/article.qmd
Formatting Fixes and new spike fig

@@
 
+![The Spike metric applied to Ukraine](imgs/Ukraine_Spikes.png)
+
 ### News Content Data
@@
 
-To prepare the training dataset, our full sample of the GDELT corpus was split into a single $512xn$ matrix for each country and month where $n$ is the number of articles published in that month related to that country.
+To prepare the training dataset, our full sample of the GDELT corpus was split into a single $512\times n$ matrix for each country and month where $n$ is the number of articles published in that month related to that country.
 Each of these matrices was then split into several random subsamples of 50 articles each.
-Samples for countries in which $n<50$ were padded to $512x50$.
+Samples for countries in which $n<50$ were padded to $512\times50$.
 This process had two effects on training.
@@
 With the encoder's training complete, we returned to the original base dataset.
-Each country-month matrix was again split into $512x50$ matrices.
+Each country-month matrix was again split into $512\times50$ matrices.
 Each of these samples could then be passed into the encoder, which distilled it into a $length-16$ vector.
-Each vector was then stacked, resulting in a single training example for each country and month of size $16x\frac{n}{50}$.
+Each vector was then stacked, resulting in a single training example for each country and month of size $16\times\frac{n}{50}$.
 This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target.
@@
 with data sampling and aggregation approaches. We used either monthly averages 
-to create $512x1$ vectors or article level samples to create $512x50$ matrices
+to create $512\times1$ vectors or article level samples to create $512\times50$ matrices
 in our dataset preparation. While we received some promising results, it is possible 
 that the first approach makes the data too coarse and the second too noisy. Another
-alternative could be generating daily averages and combining them into $512x31$ (largest number of days in a month)
+alternative could be generating daily averages and combining them into $512\times31$ (largest number of days in a month)
 matrices. This could potentially be more sensitive to the signal than the monthly averages

### 2023-05-02 docs/hes_final/article.qmd
New Abstract and Intro Rewrite

@@
 abstract: |
-  In 2021, UNICEF responded to 483 new and ongoing humanitarian situations in 153 countries (natural disasters, socio-political crises, health emergencies, nutrition crises, etc.) in addition to COVID-19 response efforts globally.
-Innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children.
-Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need.
-Predicting conflict has long been of interest to researchers and practitioners of international relations, and statistical and computational methods to do so can be found in the literature as early as 1944.
-Today, much of the work in this area has continued to focus on country or region-specific models to predict conflict.
+The task of forecasting conflict escalation is of great interest to policy makers, executives, and social scientists.
+In this article, we propose a novel, text-first approach to computational conflict prediction using contemporary natural language processing and machine learning techniques.
+We outline three methods to extract generalizable conflict signals directly from a corpus of document embeddings derived from publicly available news sources.
+The first uses text data enriched by data about the target country, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features.
+We demonstrate that usable conflict signals can be extracted directly from text sources, and we outline potential avenues for further research that we believe may yield models that can provide early warnings to decision makers to facilitate strategic planning, resource allocation, and disaster mitigation.
 
@@
 ### Motivation
-In 2021, UNICEF responded to 483 new and ongoing humanitarian situations in 153 countries (natural disasters, socio-political crises, health emergencies, nutrition crises, etc.) in addition to COVID-19 response efforts globally.
-This is a sharp increase compared to 2018, when UNICEF responded to 285 new and ongoing humanitarian emergencies in 90 countries.
-With the rising threat of the climate emergency and shifts toward authoritarianism, children around the world will be increasingly vulnerable to the impacts of catastrophe and conflict.
-Recent trends and global economic uncertainty call into question whether Individual, corporate, and government donor contributions will grow in step with these increased needs.
-
-According to UNICEF’s 2023 Humanitarian Action for Children publication,
-
-> “Conflict drives 80 per cent of all humanitarian needs.
-The number of countries experiencing violent conflict is at a thirty-year high.
-And, around the world, attacks on children continue unabated.
-From Ukraine to Yemen and from Ethiopia to Nigeria and beyond, warring parties are flouting one of the most basic rules of war: protect the children.
-… Altogether, nearly 37 million children worldwide are displaced due to conflict and violence, a number of displaced children not seen since the Second World War.
-And this number does not even include those children displaced by poverty or climate change or by the search for a better life.
-Nor does it include those children displaced in 2022 due to the war in Ukraine, which itself has caused the fastest-growing refugee crisis in Europe since the Second World War.
-… lack of rigorous data and analysis of human-induced hazards remains challenging”[@unicef]
-
 Innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children.
 Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need.
-Specifically, such an automated approach will reduce efforts spent monitoring world events and allow for the initiation of emergency preparedness protocols sooner.
+Predicting conflict has long been of interest to researchers and practitioners of international relations, and statistical and computational methods to do so can be found in the literature as early as 1944.
+Today, much of the work in this area has continued to focus on country or regionspecific models to predict conflict.
+
+This project was undertaken in response to the acute needs of the UNICEF Emergency Operations Team.
+In 2021, UNICEF responded to 483 new and ongoing humanitarian situations in 153 countries (natural disasters, socio-political crises, health emergencies, nutrition crises, etc.) in addition to its global COVID-19 response efforts.
+This represents a sharp increase since 2018, and the number of disasters continues to climb.
+With the rising threat of the climate emergency, global shifts toward authoritarianism, and increasing levels of global conflict, innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children.
+Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need.
 

### 2023-05-02 docs/hes_final/article.qmd
Swapped FFNN and Transformer Sections

@@
 
+## Feed Forward Neural Network
+
+Our second approach employed a feed forward neural network model.
+Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner [@sainburg2021parametric].
+Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space.
+The fit UMAP model, was applied to the validation and test data without supervision.
+These reduced, 128 dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons.
+The model's 4577 parameters were trained for 3 epochs.
+
+![Distribution of supervised UMAP reduction to 2 dimensions](imgs/fig-umap2d-jointplot.png){width=50%}
+
+![Scatterplot of 2d UMAP reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){width=50%}
+
+Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total') and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges, 4 categories of gross national income per capita, and 3 categories of Fragile States Index ranges).
+
+```{mermaid}
+%%| fig-width: 3
+%%| fig-height: 5
+%%| fig-cap: Fig 6. Feed forward neural network model architecture
+flowchart TD
+    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)
+
+    Articles --> UMAP{Supervised UMAP}
+    UMAP --> Compact(128 dimension representations)
+
+    Compact --> FFNN{Feed Forward Neural Network}
+    FFNN --> Pred(Prediction of conflict escalation in 3 months)
+
+    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
+    Spike(Conflict response binary variable) --> |included during training|UMAP
+
+    Spike -->FFNN
+
+    Context[Country context indicators] --> FFNN
+```
+
 ### Transformer Model using Resampled Embeddings
 
-Our second approach used a pair of transformer models to extract information directly from the GDELT embedding corpus.[@vaswani2017attention] Architecturally, both models are encoder-only transformers, but for simplicity, we will refer to the first of these models as “the encoder” and the second as the “classifier”.
+Our third approach used a pair of transformer models to extract information directly from the GDELT embedding corpus.[@vaswani2017attention] Architecturally, both models are encoder-only transformers, but for simplicity, we will refer to the first of these models as “the encoder” and the second as the “classifier”.
 
@@
 ```
-
-## Feed Forward Neural Network
-
-Our third approach employed a feed forward neural network model.
-Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner [@sainburg2021parametric].
-Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space.
-The fit UMAP model, was applied to the validation and test data without supervision.
-These reduced, 128 dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons.
-The model's 4577 parameters were trained for 3 epochs.
-
-![Distribution of supervised UMAP reduction to 2 dimensions](imgs/fig-umap2d-jointplot.png){width=50%}
-
-![Scatterplot of 2d UMAP reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){width=50%}
-
-Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total') and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges, 4 categories of gross national income per capita, and 3 categories of Fragile States Index ranges).
-
-```{mermaid}
-%%| fig-width: 3
-%%| fig-height: 5
-%%| fig-cap: Fig 6. Feed forward neural network model architecture
-flowchart TD
-    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)
-
-    Articles --> UMAP{Supervised UMAP}
-    UMAP --> Compact(128 dimension representations)
-
-    Compact --> FFNN{Feed Forward Neural Network}
-    FFNN --> Pred(Prediction of conflict escalation in 3 months)
-
-    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
-    Spike(Conflict response binary variable) --> |included during training|UMAP
-
-    Spike -->FFNN
-
-    Context[Country context indicators] --> FFNN
-```
-
 ## Results

### 2023-05-02 docs/hes_final/article.qmd
Formatting Fixes
New Figures and explanations

@@
 In this article, we propose a novel, text-first approach to computational conflict prediction using contemporary natural language processing and machine learning techniques.
-We outline three methods to extract generalizable conflict signals directly from a corpus of document embeddings derived from publicly available news sources.
-The first uses text data enriched by data about the target country, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features.
+We outline three methods to extract generalizable conflict signals directly from a corpus of ofer 2.5 million document embeddings derived from publicly available news sources.
+The first uses text data enriched by country context features, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features.
 We demonstrate that usable conflict signals can be extracted directly from text sources, and we outline potential avenues for further research that we believe may yield models that can provide early warnings to decision makers to facilitate strategic planning, resource allocation, and disaster mitigation.
@@
 ### Motivation
-Innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children.
-Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need.
-Predicting conflict has long been of interest to researchers and practitioners of international relations, and statistical and computational methods to do so can be found in the literature as early as 1944.
-Today, much of the work in this area has continued to focus on country or regionspecific models to predict conflict.
-
 This project was undertaken in response to the acute needs of the UNICEF Emergency Operations Team.
@@
 ![The Spike metric applied to Ukraine](imgs/Ukraine_Spikes.png)
+
+Figure 1 shows how this metric captures conflict fluctuations in Ukraine.
+Prior to the Russian invasion of February 2022, small fluctuations were able to trigger a spike.
+However, much larger fluctuations in the period following the invasion are filtered out, as the spike formula adjusts to the higher level of baseline conflict.
+The increase in violence during the Ukrainian counter-offensive in the late summer again triggers a spike, and this reflects significant changes in the situation on the ground.
 
@@
 
+![Reducing Embedding Dimensionality with Transformer Encoder](imgs/Transformer_Viz.png){width=50%}
+
 With the encoder's training complete, we returned to the original base dataset.

### 2023-05-02 docs/hes_final/article.qmd
Rearanged and renamed model sections

@@
 
+### Training Dataset
+
 While many of GDELT’s databases reflect news articles from 1 January 1979 onward, GDELT’s news article embeddings exist only from 1 January 2020, limiting the time period for which training data can be drawn.
 Availability of ACLED’s monthly event and fatality totals varies by country, but event and fatality totals are available for all countries since at least 1 January 2020.
+
 Our training data was derived from these GDELT and ACLED data.
 GDELT news article embeddings served as our independent variable, while the “spike” metric derived from ACLED’s conflict data served as our target variable.
+The spike metric was used to label article data on a three-month time lag, allowing us to make predictions of future events.
 Hard dates were set for train, test, and validation splits, and none of our models saw data published after October 2022 during training, with the remaining months used for testing and validation.
@@
 ```
+## Conflict Prediction Using Text and Country Features
 
-### Statistical model using monthly embedding averages
+Our first approach employed a feed forward neural network model.
+Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner [@sainburg2021parametric].
+Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space.
+The fit UMAP model, was applied to the validation and test data without supervision.
+These reduced, 128 dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons.
+The model's 4577 parameters were trained for 3 epochs.
+
+![Distribution of supervised UMAP reduction to 2 dimensions](imgs/fig-umap2d-jointplot.png){width=50%}
+
+![Scatterplot of 2d UMAP reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){width=50%}
+
+Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total') and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges, 4 categories of gross national income per capita, and 3 categories of Fragile States Index ranges).
+
+```{mermaid}
+%%| fig-width: 3
+%%| fig-height: 5
+%%| fig-cap: Fig 6. Feed forward neural network model architecture
+flowchart TD
+    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)
+
+    Articles --> UMAP{Supervised UMAP}
+    UMAP --> Compact(128 dimension representations)
+
+    Compact --> FFNN{Feed Forward Neural Network}
+    FFNN --> Pred(Prediction of conflict escalation in 3 months)
+
+    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
+    Spike(Conflict response binary variable) --> |included during training|UMAP
+
+    Spike -->FFNN
+
+    Context[Country context indicators] --> FFNN
+```
+
+### Conflict Prediction Using Text and Country Labels
 
@@
 
-## Feed Forward Neural Network
-
-Our second approach employed a feed forward neural network model.
-Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner [@sainburg2021parametric].
-Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space.
-The fit UMAP model, was applied to the validation and test data without supervision.
-These reduced, 128 dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons.
-The model's 4577 parameters were trained for 3 epochs.
-
-![Distribution of supervised UMAP reduction to 2 dimensions](imgs/fig-umap2d-jointplot.png){width=50%}
-
-![Scatterplot of 2d UMAP reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){width=50%}
-
-Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total') and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges, 4 categories of gross national income per capita, and 3 categories of Fragile States Index ranges).
-
-```{mermaid}
-%%| fig-width: 3
-%%| fig-height: 5
-%%| fig-cap: Fig 6. Feed forward neural network model architecture
-flowchart TD
-    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)
-
-    Articles --> UMAP{Supervised UMAP}
-    UMAP --> Compact(128 dimension representations)
-
-    Compact --> FFNN{Feed Forward Neural Network}
-    FFNN --> Pred(Prediction of conflict escalation in 3 months)
-
-    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
-    Spike(Conflict response binary variable) --> |included during training|UMAP
-
-    Spike -->FFNN
-
-    Context[Country context indicators] --> FFNN
-```
-
-### Transformer Model using Resampled Embeddings
+### Conflict Prediction Using Text Alone
 

### 2023-05-02 docs/hes_final/article.qmd
Language Tweaks

@@
 ```
-## Conflict Prediction Using Text and Country Features
+### Conflict Prediction Using Text and Country Features
 
@@
 In this approach we prepared the dataset by averaging the embeddings per country per month.
-We also included one-hot encoded columns for countries.
+We also included one-hot encoded columns for countries, but no other country features were added.
 Since we only had access to embeddings starting in January 2020, the total length of this dataset was just a little over 6,000 samples (training, validation, and test sets combined).
@@
 
-This new dataset offered major advantages in combatting the curse of dimensionality, however, it also represented a distorted view of the world, which overrepresents countries that receive more media coverage and underrepresents those that receive les, amplifying the biases already present in the GDELT dataset.
+This new dataset offered major advantages in combatting the curse of dimensionality, however, it also represented a distorted view of the world, which overrepresents countries that receive more media coverage and underrepresents those that receive less, amplifying the biases already present in the GDELT dataset.
 Despite this limitation, this dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target.
@@
 This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target.
-In contrast to our statistical modeling approach, the transformer models are trained purely on text embeddings.
+In contrast to our earlier approaches, the dual-transformer model was trained purely on text embeddings.
 The model was not conditioned by country, and no additional features were added beyond the text corpus itself.

### 2023-05-02 docs/hes_final/article.qmd
Margin, spacing and figure size adjustments

@@
     - twocolumn
+    margin-left: 25mm
+    margin-right: 25mm
+    margin-top: 25mm
+    margin-bottom: 25mm
+    whitespace: medium
 bibliography: bibliography.bibtex
@@
       \lefthyphenmin=5
+      \setlength{\belowcaptionskip}{-5pt}
 embed-resources: true
@@
 ---
-
 ## Introduction
@@
 ## Methodology
+
 The principal difference between our approach and that of prior work is in our choice of predictive features, which we extract directly from text embeddings of publicly available news coverage.
@@
 %%| fig-width: 3
-%%| fig-height: 5
+%%| fig-height: 3
 %%| fig-cap: Fig 6. Feed forward neural network model architecture
@@
 ```
-
 ### Conflict Prediction Using Text and Country Labels
@@
 %%| fig-width: 3
-%%| fig-height: 5
+%%| fig-height: 3
 %%| fig-cap: Fig 2. Statistical model architecture
@@
 ```
-
 ### Conflict Prediction Using Text Alone
@@
 Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons.
-The output of these 16 neurons would be used in the classification stage.
+The output of these 16 neurons would be used in the classification stage
 
@@
 %%| fig-width: 3
-%%| fig-height: 5
+%%| fig-height: 3
 %%| fig-cap: Fig 3. Transformer model architecture

### 2023-05-02 docs/hes_final/article.qmd
Extend metrics figure title for interpretation

@@
 
-![Model performance metrics](imgs/tbl-best-max-metrics.png){width=50%}
+![Model performance metrics (higher values are better)](imgs/tbl-best-max-metrics.png){width=50%}
 
-![Model performance metrics](imgs/tbl-best-min-metrics.png){width=50%}
+![Model performance metrics (lower values are better)](imgs/tbl-best-min-metrics.png){width=50%}
 

### 2023-05-03 docs/hes_final/article.qmd
Rewrite the results section.

@@
 
-We evaluated our models with both a research and a practical objectives in mind.
+We evaluated the models against a set of common performance metrics (see Figures 4-9).
+Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results.
+Each of the models outperformed the other two
+in a subset of these metrics. This suggests that the choice of the model architecture for this task would depend on specific application goals. We assume that UNICEF decision makers would
+appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives. At the same time, if the false positive rate is too high the signal would become too
+noisy for the tool to be useful. Given these assumptions and the class imbalance of the original
+dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for
+class imbalance) seem to be most relevant metrics. Two models using country features seem to perform
+better on these two metrics.
 
-From a research perspective we were interested in feasibility of making accurate and generalizable predictions of
-conflict escalation using publicly available text embeddings.
-We looked at the area under the ROC curve, F1 score, precision, recall and confusion matrix to design a holistic evaluation method at both fine-tuning/validation stage and final test set prediction stage.
-Despite conceptual differences in dataset creation and modeling techniques, both of the described above approaches produced comparative results.
+In this work we defined the target in a way that was not used in prior literature, we cannot directly
+compare our models' performance to others. The most similar work used a presense or absense
+of any armed conflict related fatality as their binary target  [@10.1093/jeea/jvac025].
+They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in peaceful countries) with their text based (LDA topic retrieval) model.
+Our best ROC_AUC score is 69% but the comparison is not very meaningful due to difference in targets.
 
@@
 
-Further investigation showed that while overall numeric metrics look similar, the Transformer-based model generalized better across the globe.
+Further investigation showed that while overall numeric metrics look similar and there seems to be a slight performance advantage in using country data in addition to text features, the Transformer-based text-only model generalized better across the globe.
 Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the
@@
 With that in mind, cost of deployment is also an important factor for model evaluation.
-Average embedding dataset fits in memory while resampling article level embeddings requires additional data engineering
-steps to be processed in batches. Both dataset size and corresponding model size make average embeddings based statistical
-model more attractive from this perspective.
+Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings in two other proposed methods requires additional data engineering
+steps to be processed in batches.
 
-We think that both approaches produced promising results and the model choice depends on the developers' specific goals and available resources.
-
-### Comparison with ...
+We think that all three approaches produced promising results and the model choice depends on the developers' goals, performance metrics and available resources.
 

### 2023-05-03 docs/hes_final/article.qmd
Reordering mermaid plots

--- 2023-05-02 docs/hes_final/article.qmd
@@
 ### Defining Conflict Escalation
+
 Our task of modeling conflict escalation required a precise definition of what conflict escalation is.
@@
 
-While many of GDELT’s databases reflect news articles from 1 January 1979 onward, GDELT’s news article embeddings exist only from 1 January 2020, limiting the time period for which training data can be drawn.
-Availability of ACLED’s monthly event and fatality totals varies by country, but event and fatality totals are available for all countries since at least 1 January 2020.
-
-Our training data was derived from these GDELT and ACLED data.
+Our training data was derived from these GDELT and ACLED datasets.
 GDELT news article embeddings served as our independent variable, while the “spike” metric derived from ACLED’s conflict data served as our target variable.
 The spike metric was used to label article data on a three-month time lag, allowing us to make predictions of future events.
-Hard dates were set for train, test, and validation splits, and none of our models saw data published after October 2022 during training, with the remaining months used for testing and validation.
+
+ GDELT’s news article embeddings exist only from 1 January 2020, limiting the time period for which training data can be drawn.
+Availability of ACLED’s monthly event and fatality totals varies by country, but event and fatality totals are available for all countries since at least 1 January 2020.
+Our training dataset thus ran from January 2020 through October 2022.
+Data from November 2022 through March 2023 was used for model validation and testing.
 
@@
 
+```{mermaid}
+%%| fig-width: 5
+%%| fig-height: 3
+%%| fig-cap: Fig 3. Transformer model architecture
+flowchart TD
+    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)
+
+    Articles --> Trans1{Encoder\n transformer}
+    Trans1 -->|classifier head present during training| Compact(16 dimension representations)
+
+    Compact --> Trans2{Binary\n classifier\n transformer}
+    Trans2 --> Pred(Prediction of conflict escalation in 3 months)
+
+    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
+    Spike(Conflict response\n binary variable) --> Trans2
+
+    Spike -->|included during model training| Trans1
+```
+
 To prepare the training dataset, our full sample of the GDELT corpus was split into a single $512\times n$ matrix for each country and month where $n$ is the number of articles published in that month related to that country.
@@
 
-```{mermaid}
-%%| fig-width: 3
-%%| fig-height: 3
-%%| fig-cap: Fig 3. Transformer model architecture
-flowchart TD
-    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)
-
-    Articles --> Trans1{Encoder\n transformer}
-    Trans1 -->|classifier head present during training| Compact(16 dimension representations)
-
-    Compact --> Trans2{Binary\n classifier\n transformer}
-    Trans2 --> Pred(Prediction of conflict escalation in 3 months)
-
-    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
-    Spike(Conflict response\n binary variable) --> Trans2
-
-    Spike -->|included during model training| Trans1
-```
 ## Results
@@
 
+{{< pagebreak >}}
+
 ## References

### 2023-05-03 docs/hes_final/article.qmd
add references, fix figure captions, rm line breaks

--- 2023-05-03 docs/hes_final/article.qmd
@@
     whitespace: medium
+    mermaid:
+      theme: neutral
 bibliography: bibliography.bibtex
@@
 
-We developed two approaches for extracting conflict signals from text embeddings.
-In the first approach we used a fit-in-memory dataset of monthly embedding averages per country per month and applied statistical classification models such as K Nearest Neighbors and XGBoost.
-In the second approach we stacked multiple article-level embeddings and applied Transformer-based models to extract a signal directly from the text corpus.
-See the figures (X, Y) for details of both methods.
+We developed three approaches for extracting conflict signals from text embeddings.
+The first approach used supervised dimensionality reduction and a feed-forward neural network conditioned with country context characteristics.
+In the second approach we used a fit-in-memory dataset of monthly embedding averages per country per month and applied statistical classification models such as K Nearest Neighbors and XGBoost.
+In the third approach we stacked multiple article-level embeddings and applied Transformer-based models to extract a signal directly from the text corpus.
+See the @fig-ffnn, @fig-xgboost, and @fig-transformer for details of modeling approaches.
 
@@
 
-![The Spike metric applied to Ukraine](imgs/Ukraine_Spikes.png)
+![The Spike metric applied to Ukraine](imgs/Ukraine_Spikes.png){#fig-spike}
 
-Figure 1 shows how this metric captures conflict fluctuations in Ukraine.
+@fig-spike shows how this metric captures conflict fluctuations in Ukraine.
 Prior to the Russian invasion of February 2022, small fluctuations were able to trigger a spike.
@@
 
+::: {#fig-dgp layout-ncol=1}
 ```{mermaid}
@@
 %%| fig-height: 5
-%%| fig-cap: Fig 1. Overview of data generation process for our conflict dataset 
 flowchart TD
@@
 ```
+Overview of data generation process for our conflict dataset 
+:::
+
 ### Conflict Prediction Using Text and Country Features
@@
 
-![Distribution of supervised UMAP reduction to 2 dimensions](imgs/fig-umap2d-jointplot.png){width=50%}
+![Density of embeddings resulting from supervised UMAP reduction to 2 dimensions](imgs/fig-umap2d-jointplot.png){#fig-umap2d-jointplot}
 
-![Scatterplot of 2d UMAP reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){width=50%}
+![Scatterplot of 2-dimension UMAP-reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){#fig-umap2d-ds-targets}
 
-Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total') and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges, 4 categories of gross national income per capita, and 3 categories of Fragile States Index ranges).
+Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total' from UNICEF's State of the World's Children report) and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges [@hdi], 4 categories of gross national income per capita [@wb], and 3 categories of Fragile States Index ranges[@fsi]).
 
+::: {#fig-ffnn layout-ncol=1}
 ```{mermaid}
@@
 %%| fig-height: 3
-%%| fig-cap: Fig 6. Feed forward neural network model architecture
 flowchart TD
@@
 ```
+Feed forward neural network model architecture
+:::
 ### Conflict Prediction Using Text and Country Labels
@@
 
+::: {#fig-xgboost layout-ncol=1}
 ```{mermaid}
@@
 %%| fig-height: 3
-%%| fig-cap: Fig 2. Statistical model architecture
 flowchart TD
@@
 ```
+Statistical model architecture
+:::
 ### Conflict Prediction Using Text Alone
@@
 
-![Reducing Embedding Dimensionality with Transformer Encoder](imgs/Transformer_Viz.png){width=50%}
+![Reducing Embedding Dimensionality with Transformer Encoder](imgs/Transformer_Viz.png){#fig-transformer-viz}
 
@@
 
+::: {#fig-transformer layout-ncol=1}
 ```{mermaid}
@@
 %%| fig-height: 3
-%%| fig-cap: Fig 3. Transformer model architecture
 flowchart TD
@@
 ```
+Transformer model architecture
+:::
 ## Results
@@
 Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results.
-Each of the models outperformed the other two
-in a subset of these metrics. This suggests that the choice of the model architecture for this task would depend on specific application goals. We assume that UNICEF decision makers would
-appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives. At the same time, if the false positive rate is too high the signal would become too
-noisy for the tool to be useful. Given these assumptions and the class imbalance of the original
-dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for
-class imbalance) seem to be most relevant metrics. Two models using country features seem to perform
+Each of the models outperformed the other two in a subset of these metrics.
+This suggests that the choice of the model architecture for this task would depend on specific application goals.
+We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives.
+At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful.
+Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics.
+Two models using country features seem to perform
 better on these two metrics.
 
-In this work we defined the target in a way that was not used in prior literature, we cannot directly
-compare our models' performance to others. The most similar work used a presense or absense
-of any armed conflict related fatality as their binary target  [@10.1093/jeea/jvac025].
+In this work we defined the target in a way that was not used in prior literature, we cannot directly compare our models' performance to others.
+The most similar work used a presense or absense of any armed conflict related fatality as their binary target  [@10.1093/jeea/jvac025].
 They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in peaceful countries) with their text based (LDA topic retrieval) model.
@@
 
-![Model performance metrics (higher values are better)](imgs/tbl-best-max-metrics.png){width=50%}
+![Model performance metrics (higher values are better)](imgs/tbl-best-max-metrics.png){#fig-best-max-metrics}
 
-![Model performance metrics (lower values are better)](imgs/tbl-best-min-metrics.png){width=50%}
+![Model performance metrics (lower values are better)](imgs/tbl-best-min-metrics.png){#fig-best-min-metrics}
 
-![Receiver Operator Characteristics](imgs/fig-roc.png){width=50%}
+![Receiver Operator Characteristics](imgs/fig-roc.png){#fig-roc}
 
-![Precision Recall Curve](imgs/fig-prc.png){width=50%}
+![Precision Recall Curve](imgs/fig-prc.png){#fig-prc}
 
 Further investigation showed that while overall numeric metrics look similar and there seems to be a slight performance advantage in using country data in addition to text features, the Transformer-based text-only model generalized better across the globe.
-Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the
-generalizability is a significant reason to prefer Transformer architecture and resampled dataset.
+Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the generalizability is a significant reason to prefer Transformer architecture and resampled dataset.
 
-![Transformer model Predictions on test data](imgs/fig-transformer-test-predictions.png){width=55%}
+![FFNN model predictions on test data](imgs/fig-ffnn-test-predictions.png){#fig-ffnn-test-predictions}
 
-![XGBoost model predictions on test data](imgs/fig-xgboost-test-predictions.png){width=55%}
+![XGBoost model predictions on test data](imgs/fig-xgboost-test-predictions.png){#fig-xgboost-test-predictions}
 
-![FFNN model predictions on test data](imgs/fig-ffnn-test-predictions.png){width=55%}
+![Transformer model Predictions on test data](imgs/fig-transformer-test-predictions.png){#fig-transformer-test-predictions}
 
@@
 With that in mind, cost of deployment is also an important factor for model evaluation.
-Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings in two other proposed methods requires additional data engineering
-steps to be processed in batches.
+Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings in two other proposed methods requires additional data engineering steps to be processed in batches.
 
@@
 
-In this work we explored the possibility of predicting future geo-political conflict escalation at a country level
-using a global collection of news publications. We explored different ways of aggregating the data and
-applied both statistical machine learning and deep learning models.
+In this work we explored the possibility of predicting future geo-political conflict escalation at a country level using a global collection of news publications.
+We explored different ways of aggregating the data and applied both statistical machine learning and deep learning models.
 The results suggest that text embeddings provide a strong enough signal to make such predictions.
@@
 
-One of the potential research extensions of this work is to continue experimenting
-with data sampling and aggregation approaches. We used either monthly averages 
-to create $512\times1$ vectors or article level samples to create $512\times50$ matrices
-in our dataset preparation. While we received some promising results, it is possible 
-that the first approach makes the data too coarse and the second too noisy. Another
-alternative could be generating daily averages and combining them into $512\times31$ (largest number of days in a month)
-matrices. This could potentially be more sensitive to the signal than the monthly averages
-and less noisy than article level samples.
+One of the potential research extensions of this work is to continue experimenting with data sampling and aggregation approaches.
+We used either monthly averages to create $512\times1$ vectors or article level samples to create $512\times50$ matrices in our dataset preparation.
+While we received some promising results, it is possible that the first approach makes the data too coarse and the second too noisy.
+Another alternative could be generating daily averages and combining them into $512\times31$ (largest number of days in a month) matrices.
+This could potentially be more sensitive to the signal than the monthly averages and less noisy than article level samples.
 
@@
 
-Finally, due to time and resource constraints in this project we used 20% subset
-of daily news articles. Increasing that percentage could potentially lead to better results.
+Finally, due to time and resource constraints in this project we used 20% subset of daily news articles. Increasing that percentage could potentially lead to better results.
 
@@
 Building an application with a user interface could be a natural extension of this work.
-We envision a dashboard where a user can view a world map where countries are heatmap colored
-according to the estimated probability of escalating conflict. Countries with highest risk
-can be flagged or shown to a user in a separate view. In addition to the visual part
-of this application it is important to ensure that the model is deployed and
-can receive new data input and make inference.
+We envision a dashboard where a user can view a world map where countries are heatmap colored according to the estimated probability of escalating conflict.
+Countries with highest risk can be flagged or shown to a user in a separate view.
+In addition to the visual part of this application it is important to ensure that the model is deployed and can receive new data input and make inference.
 
-Building a more robust and automated data processing pipeline is another important direction. This would involve
-designing either a batch or streaming system that can reliably collect data, preprocess
-and aggregate it into required format, train the model and make predictions. The system should handle 
-computational resources and cost constraints.
+Building a more robust and automated data processing pipeline is another important direction.
+This would involve designing either a batch or streaming system that can reliably collect data, preprocess and aggregate it into required format, train the model and make predictions.
+The system should handle computational resources and cost constraints.
 

### 2023-05-03 docs/hes_final/article.qmd
update text with figure references

@@
 ```{mermaid}
-%%| fig-width: 3
+%%| fig-width: 5
 %%| fig-height: 3
@@
 
-We evaluated the models against a set of common performance metrics (see Figures 4-9).
+We evaluated the models against a set of common performance metrics (see @fig-best-max-metrics, @fig-best-min-metrics, @fig-roc, @fig-prc).
 Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results.

### 2023-05-04 docs/hes_final/article.qmd
add test accuracy maps, use markdown tables

@@
       \makeatother
+       \usepackage[font=small,labelfont={bf,small}]{caption}
 # https://tex.stackexchange.com/questions/31301/how-to-reduce-the-number-of-hyphenation
@@
 In this article, we propose a novel, text-first approach to computational conflict prediction using contemporary natural language processing and machine learning techniques.
-We outline three methods to extract generalizable conflict signals directly from a corpus of ofer 2.5 million document embeddings derived from publicly available news sources.
+We outline three methods to extract generalizable conflict signals directly from a corpus of over 2.5 million document embeddings derived from publicly available news sources.
 The first uses text data enriched by country context features, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features.
@@
 
-Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total' from UNICEF's State of the World's Children report) and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges [@hdi], 4 categories of gross national income per capita [@wb], and 3 categories of Fragile States Index ranges[@fsi]).
+Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total' from UNICEF's State of the World's Children report [@sowc]) and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges [@hdi], 4 categories of gross national income per capita [@wb], and 3 categories of Fragile States Index ranges[@fsi]).
 
@@
 ```{mermaid}
-%%| fig-width: 5
+%%| fig-width: 3
 %%| fig-height: 3
@@
 :::
+
 ## Results
 
-We evaluated the models against a set of common performance metrics (see @fig-best-max-metrics, @fig-best-min-metrics, @fig-roc, @fig-prc).
+We evaluated the models against a set of common performance metrics (see @tbl-best-max-metrics, @tbl-best-min-metrics, @fig-roc, @fig-prc).
 Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results.
@@
 Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics.
-Two models using country features seem to perform
-better on these two metrics.
+Two models using country features seem to perform better on these two metrics.
 
@@
 
-![Model performance metrics (higher values are better)](imgs/tbl-best-max-metrics.png){#fig-best-max-metrics}
+\onecolumn
 
-![Model performance metrics (lower values are better)](imgs/tbl-best-min-metrics.png){#fig-best-min-metrics}
+::: {layout="[[1, 1]]"}
+| metric               | model       |   value |
+|:---------------------|:------------|--------:|
+| False Discovery Rate | Transformer |    0.80 |
+| False Discovery Rate | XGBoost     |    0.83 |
+| False Discovery Rate | FFNN        |    0.74 |
+| False Negative Rate  | Transformer |    0.53 |
+| False Negative Rate  | XGBoost     |    0.28 |
+| False Negative Rate  | FFNN        |    0.59 |
+| False Positive Rate  | Transformer |    0.30 |
+| False Positive Rate  | XGBoost     |    0.51 |
+| False Positive Rate  | FFNN        |    0.27 |
+
+: Model performance metrics (smaller values are better) {#tbl-best-min-metrics}
+
+| metric             | model       |   value |
+|:-------------------|:------------|--------:|
+| Cohen Kappa Score  | XGBoost     |    0.09 |
+| Cohen Kappa Score  | FFNN        |    0.12 |
+| Cohen Kappa Score  | Transformer |    0.11 |
+| F1 Score           | FFNN        |    0.32 |
+| F1 Score           | XGBoost     |    0.28 |
+| F1 Score           | Transformer |    0.28 |
+| Precision Score    | XGBoost     |    0.17 |
+| Precision Score    | FFNN        |    0.26 |
+| Precision Score    | Transformer |    0.20 |
+| Recall Score       | XGBoost     |    0.72 |
+| Recall Score       | Transformer |    0.47 |
+| Recall Score       | FFNN        |    0.41 |
+| Roc Auc Score      | Transformer |    0.64 |
+| Roc Auc Score      | XGBoost     |    0.69 |
+| Roc Auc Score      | FFNN        |    0.64 |
+| True Negative Rate | Transformer |    0.70 |
+| True Negative Rate | XGBoost     |    0.49 |
+| True Negative Rate | FFNN        |    0.73 |
+| True Positive Rate | FFNN        |    0.41 |
+| True Positive Rate | XGBoost     |    0.72 |
+| True Positive Rate | Transformer |    0.47 |
+: Model performance metrics (larger values are better) {#tbl-best-max-metrics}
+
+:::
+\twocolumn
+
+\onecolumn
+
+::: {layout="[[1, 1]]"}
 
@@
 
+:::
+::: {layout="[[1, 1]]"}
+
+![FFNN model accuracy on test data](imgs/fig-ffnn-test-accuracy.png){#fig-ffnn-test-accuracy}
+
+![XGBoost model accuracy on test data](imgs/fig-xgboost-test-accuracy.png){#fig-xgboost-test-accuracy}
+
+:::
+::: {layout="[[1, 1]]"}
+
+![Transformer model accuracy on test data](imgs/fig-transformer-test-accuracy.png){#fig-transformer-test-accuracy}
+
+:::
+\twocolumn
+
 Further investigation showed that while overall numeric metrics look similar and there seems to be a slight performance advantage in using country data in addition to text features, the Transformer-based text-only model generalized better across the globe.
 Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the generalizability is a significant reason to prefer Transformer architecture and resampled dataset.
-
-![FFNN model predictions on test data](imgs/fig-ffnn-test-predictions.png){#fig-ffnn-test-predictions}
-
-![XGBoost model predictions on test data](imgs/fig-xgboost-test-predictions.png){#fig-xgboost-test-predictions}
-
-![Transformer model Predictions on test data](imgs/fig-transformer-test-predictions.png){#fig-transformer-test-predictions}
 
@@
 
-We think that all three approaches produced promising results and the model choice depends on the developers' goals, performance metrics and available resources.
+We think that all three approaches produced promising results and the model choice depends on the user's goals, performance metrics and available resources.
 

### 2023-05-04 docs/hes_final/article.qmd
More changes to results

@@
 
-We evaluated the models against a set of common performance metrics (see @tbl-best-max-metrics, @tbl-best-min-metrics, @fig-roc, @fig-prc).
-Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results.
-Each of the models outperformed the other two in a subset of these metrics.
-This suggests that the choice of the model architecture for this task would depend on specific application goals.
-We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives.
-At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful.
-Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics.
-Two models using country features seem to perform better on these two metrics.
+We evaluated the models against a set of common performance metrics (see @tbl-best-max-metrics, @tbl-best-min-metrics, @fig-roc, 
+@fig-prc). Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches 
+produced comparative results. Each of the models outperformed the other two in a subset of these metrics.
 
-In this work we defined the target in a way that was not used in prior literature, we cannot directly compare our models' performance to others.
-The most similar work used a presense or absense of any armed conflict related fatality as their binary target  [@10.1093/jeea/jvac025].
-They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in peaceful countries) with their text based (LDA topic retrieval) model.
-Our best ROC_AUC score is 69% but the comparison is not very meaningful due to difference in targets.
+Our target was defined differently from related work which made direct comparison less meaningful.
+The closest approach used a presense or absense of any armed conflict related fatality as their binary target  [@10.1093/jeea/jvac025]. 
+They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in 
+peaceful countries) with their text based (LDA topic extraction) model. Our best ROC_AUC score is 69% (achieved by text + country labels 
+XGBoost/KNN model).
+
+We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false 
+positives. At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful. Given 
+these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total 
+number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem 
+to be most relevant metrics. The relative ranks of the models are the opposite for these two metrics with each of the two models
+with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
 
@@
 
-While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making.
-With that in mind, cost of deployment is also an important factor for model evaluation.
-Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings in two other proposed methods requires additional data engineering steps to be processed in batches.
+While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a 
+model that can assist UNICEF in their decision making. With that in mind, cost of deployment is also an important factor for model 
+evaluation. Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings 
+in two other proposed methods requires additional data engineering steps to be processed in batches.
 
@@
 
-In this work we explored the possibility of predicting future geo-political conflict escalation at a country level using a global collection of news publications.
-We explored different ways of aggregating the data and applied both statistical machine learning and deep learning models.
-The results suggest that text embeddings provide a strong enough signal to make such predictions.
-Different data sampling and modeling approaches produce comparable results based on common metrics.
-Transformer-based model generalizes better across geographies while statistical model provides a size/cost of deployment advantage.
+In this work we explored the possibility of predicting future geo-political conflict escalation at a country level using a global 
+collection of news publications. We explored different ways of aggregating the data and applied both statistical machine learning and 
+deep learning models. The results suggest that text embeddings provide a strong enough signal to make such predictions. Different data 
+sampling and modeling approaches produce comparable results based on common metrics. Transformer-based model generalizes better across 
+geographies while statistical model provides a size/cost of deployment advantage.
 

### 2023-05-04 docs/hes_final/article.qmd
Extend conclusions with "so what?"

@@
 
-Further investigation showed that while overall numeric metrics look similar and there seems to be a slight performance advantage in using country data in addition to text features, the Transformer-based text-only model generalized better across the globe.
-Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the generalizability is a significant reason to prefer Transformer architecture and resampled dataset.
+Further investigation showed that while overall numeric metrics look similar, the Transformer-based text-only model generalized better 
+across the globe. Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the 
+generalizability could be a reason to prefer Transformer architecture and resampled dataset.
 
@@
 
-We think that all three approaches produced promising results and the model choice depends on the user's goals, performance metrics and available resources.
+We think that all three approaches produced promising results and the model choice depends on the user's goals, selected performance 
+metrics and available resources.
 
@@
 In this work we explored the possibility of predicting future geo-political conflict escalation at a country level using a global 
-collection of news publications. We explored different ways of aggregating the data and applied both statistical machine learning and 
-deep learning models. The results suggest that text embeddings provide a strong enough signal to make such predictions. Different data 
-sampling and modeling approaches produce comparable results based on common metrics. Transformer-based model generalizes better across 
-geographies while statistical model provides a size/cost of deployment advantage.
+collection of news publications. We compared the usage of text-only data and data enriched with country labels and metadata. We 
+experimented with different ways of aggregating the data and applied both statistical machine learning and deep learning models. The 
+results suggest that text embeddings provide a strong enough signal to make such predictions. Different data 
+sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common
+numetic metrics. Transformer-based model generalized better across geographies while statistical model provided a size/cost of 
+deployment advantage.
+
+While this work did not produce a single best model for conflict escalation predictions, it demonstrated the possibility of such 
+predictions from primarily text data. We provided a clear definition of a binary conflict escalation target and described several 
+approaches for data aggregation and modeling. We hope this work will contribute to both UNICEF operations and political science 
+research. The performance results can serve as a benchmark for future research projects in this 
+field. We outline some ideas for future work in the next section.
 
@@
 
-One of the potential research extensions of this work is to continue experimenting with data sampling and aggregation approaches.
-We used either monthly averages to create $512\times1$ vectors or article level samples to create $512\times50$ matrices in our dataset preparation.
-While we received some promising results, it is possible that the first approach makes the data too coarse and the second too noisy.
-Another alternative could be generating daily averages and combining them into $512\times31$ (largest number of days in a month) matrices.
-This could potentially be more sensitive to the signal than the monthly averages and less noisy than article level samples.
+One of the potential research extensions of this work is to continue experimenting with data sampling and aggregation approaches. There 
+are many alternatives to the approaches described above, for instance generating daily averages and combining them into $512\times31$ 
+(largest number of days in a month) matrices. This could potentially be more sensitive to the signal than the monthly averages and less 
+noisy than article level samples.
 
-The transformer models have a great deal of room for hyperparameter tuning.
-Further experimentation with the size of the encoder output, the number of articles per training sample, and implementing overlapping strides during distillation could all improve the downstream performance of the classifier.
-It may also be possible to use the output of the encoder as input to some other classification model architecture.
+The transformer models have a great deal of room for hyperparameter tuning. Further experimentation with the size of the encoder output, 
+the number of articles per training sample, and implementing overlapping strides during distillation could all improve the downstream 
+performance of the classifier. It may also be possible to use the output of the encoder as input to some other classification model 
+architecture.
 
@@
 
-Building an application with a user interface could be a natural extension of this work.
-We envision a dashboard where a user can view a world map where countries are heatmap colored according to the estimated probability of escalating conflict.
-Countries with highest risk can be flagged or shown to a user in a separate view.
-In addition to the visual part of this application it is important to ensure that the model is deployed and can receive new data input and make inference.
+Building an application with a user interface could be a natural extension of this work. We envision a dashboard where a user can view a 
+world map where countries are heatmap colored according to the estimated probability of escalating conflict. Countries with highest risk 
+can be flagged or shown to a user in a separate view. In addition to the visual part of this application, model deployment for easy
+inference is an important project.
 
-Building a more robust and automated data processing pipeline is another important direction.
-This would involve designing either a batch or streaming system that can reliably collect data, preprocess and aggregate it into required format, train the model and make predictions.
-The system should handle computational resources and cost constraints.
+Building a more robust and automated data processing pipeline is another important direction. This would involve designing either a 
+batch or streaming system that can reliably collect data, preprocess and aggregate it into required format, train the model and make 
+predictions. The system should handle computational resources and cost constraints.
 

### 2023-05-04 docs/hes_final/article.qmd
Added Ethics Section

@@
 
+### Ethical Considerations
+This work carries significant ethical considerations, especially if it were to be operationalized.
+The first and most apparent is bias in the training data.
+All of our models showed varying performance by region, and overreliance on such models by decisionmakers could lead to resources being diverted away from regions that need them, but which were underrepresented in the training data and thus had less accurate results.
+If models like these were to become widely used, they may be vulnerable to adversarial attacks.
+Large volumes of fake news content could be used to overwhelm the true signal of the model, obscuring the intentions of malicious actors, or causing panic.
+Finally, and perhaps most concerningly, is the potential for conflict prediction models to deliver self-fulfilling prophesies.
+If policy makers become convinced of impending conflict in their own or neighboring countries because of the predictions of an AI model, they may take actions that make conflict more likely.
+Reducing these risks requires emphasizing that these are models, not oracles.
+They are measuring the temperature of the global media conversation, not gazing into the future.
+Decisionmakers need to understand that though models like these can be useful tools, they are not a substitute for careful deliberation and dialogue.
+At their best, these models can provide an early warning for conflict, and if the right actions are taken, then they can be made to be wrong.
+
 ## Future work

### 2023-05-04 docs/hes_final/article.qmd
Ethical Paragraph Breaks
Restructure metrics tables

@@
 ::: {layout="[[1, 1]]"}
-| metric               | model       |   value |
-|:---------------------|:------------|--------:|
-| False Discovery Rate | Transformer |    0.80 |
-| False Discovery Rate | XGBoost     |    0.83 |
-| False Discovery Rate | FFNN        |    0.74 |
-| False Negative Rate  | Transformer |    0.53 |
-| False Negative Rate  | XGBoost     |    0.28 |
-| False Negative Rate  | FFNN        |    0.59 |
-| False Positive Rate  | Transformer |    0.30 |
-| False Positive Rate  | XGBoost     |    0.51 |
-| False Positive Rate  | FFNN        |    0.27 |
+| metric               |   FFNN |   XGBoost |   Transformer |
+|:---------------------|-------:|----------:|--------------:|
+| False Discovery Rate |   0.74 |      0.83 |          0.80 |
+| False Negative Rate  |   0.59 |      0.28 |          0.53 |
+| False Positive Rate  |   0.27 |      0.51 |          0.30 |
 
@@
 
-| metric             | model       |   value |
-|:-------------------|:------------|--------:|
-| Cohen Kappa Score  | XGBoost     |    0.09 |
-| Cohen Kappa Score  | FFNN        |    0.12 |
-| Cohen Kappa Score  | Transformer |    0.11 |
-| F1 Score           | FFNN        |    0.32 |
-| F1 Score           | XGBoost     |    0.28 |
-| F1 Score           | Transformer |    0.28 |
-| Precision Score    | XGBoost     |    0.17 |
-| Precision Score    | FFNN        |    0.26 |
-| Precision Score    | Transformer |    0.20 |
-| Recall Score       | XGBoost     |    0.72 |
-| Recall Score       | Transformer |    0.47 |
-| Recall Score       | FFNN        |    0.41 |
-| Roc Auc Score      | Transformer |    0.64 |
-| Roc Auc Score      | XGBoost     |    0.69 |
-| Roc Auc Score      | FFNN        |    0.64 |
-| True Negative Rate | Transformer |    0.70 |
-| True Negative Rate | XGBoost     |    0.49 |
-| True Negative Rate | FFNN        |    0.73 |
-| True Positive Rate | FFNN        |    0.41 |
-| True Positive Rate | XGBoost     |    0.72 |
-| True Positive Rate | Transformer |    0.47 |
+| metric             |   FFNN |   XGBoost |   Transformer |
+|:-------------------|-------:|----------:|--------------:|
+| Cohen Kappa Score  |   0.12 |      0.09 |          0.11 |
+| F1 Score           |   0.32 |      0.28 |          0.28 |
+| Precision Score    |   0.26 |      0.17 |          0.20 |
+| Recall Score       |   0.41 |      0.72 |          0.47 |
+| Roc Auc Score      |   0.64 |      0.69 |          0.64 |
+| True Negative Rate |   0.73 |      0.49 |          0.70 |
 : Model performance metrics (larger values are better) {#tbl-best-max-metrics}
@@
 All of our models showed varying performance by region, and overreliance on such models by decisionmakers could lead to resources being diverted away from regions that need them, but which were underrepresented in the training data and thus had less accurate results.
+
 If models like these were to become widely used, they may be vulnerable to adversarial attacks.
@@
 If policy makers become convinced of impending conflict in their own or neighboring countries because of the predictions of an AI model, they may take actions that make conflict more likely.
+
 Reducing these risks requires emphasizing that these are models, not oracles.

### 2023-05-04 docs/hes_final/article.qmd
Reordering Future Work, Ethics, Conclusion, rewrote future research section

@@
 
-## Conclusions
-
-In this work we explored the possibility of predicting future geo-political conflict escalation at a country level using a global 
-collection of news publications. We compared the usage of text-only data and data enriched with country labels and metadata. We 
-experimented with different ways of aggregating the data and applied both statistical machine learning and deep learning models. The 
-results suggest that text embeddings provide a strong enough signal to make such predictions. Different data 
-sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common
-numetic metrics. Transformer-based model generalized better across geographies while statistical model provided a size/cost of 
-deployment advantage.
-
-While this work did not produce a single best model for conflict escalation predictions, it demonstrated the possibility of such 
-predictions from primarily text data. We provided a clear definition of a binary conflict escalation target and described several 
-approaches for data aggregation and modeling. We hope this work will contribute to both UNICEF operations and political science 
-research. The performance results can serve as a benchmark for future research projects in this 
-field. We outline some ideas for future work in the next section.
-
-### Ethical Considerations
+## Ethical Considerations
 This work carries significant ethical considerations, especially if it were to be operationalized.
@@
 
-One of the potential research extensions of this work is to continue experimenting with data sampling and aggregation approaches. There 
-are many alternatives to the approaches described above, for instance generating daily averages and combining them into $512\times31$ 
-(largest number of days in a month) matrices. This could potentially be more sensitive to the signal than the monthly averages and less 
-noisy than article level samples.
+Due to time and resource constraints, our work used only 20% of the daily embeddings available through  GDELT, and future research could focus on scaling the methods outline.
+We also believe that continued experimentation with data sampling and aggregation approaches, as well as model tuning, could yield models that outperform those outlined here without additional data.
 
-The transformer models have a great deal of room for hyperparameter tuning. Further experimentation with the size of the encoder output, 
-the number of articles per training sample, and implementing overlapping strides during distillation could all improve the downstream 
-performance of the classifier. It may also be possible to use the output of the encoder as input to some other classification model 
-architecture.
-
-Finally, due to time and resource constraints in this project we used 20% subset of daily news articles. Increasing that percentage could potentially lead to better results.
+Finally, we believe deeper research into regional disparities in model performance is needed.
+Understanding the limitations of global media coverage for conflict prediction will be critical to any effort to operationalize this work.
 
@@
 
+## Conclusions
+
+In this work we explored the possibility of predicting future geo-political conflict escalation at a country level using a global
+collection of news publications. We compared the usage of text-only data and data enriched with country labels and metadata. We
+experimented with different ways of aggregating the data and applied both statistical machine learning and deep learning models. The
+results suggest that text embeddings provide a strong enough signal to make such predictions. Different data
+sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common
+numetic metrics. Transformer-based model generalized better across geographies while statistical model provided a size/cost of
+deployment advantage.
+
+While this work did not produce a single best model for conflict escalation predictions, it demonstrated the possibility of such
+predictions from primarily text data. We provided a clear definition of a binary conflict escalation target and described several
+approaches for data aggregation and modeling. We hope this work will contribute to both UNICEF operations and political science
+research. The performance results can serve as a benchmark for future research projects in this
+field.
+
 {{< pagebreak >}}

### 2023-05-04 docs/hes_final/article.qmd
put tables in different rows to avoid overlaps

@@
 
-::: {layout="[[1, 1]]"}
+::: {layout="[[1], [1]]"}
 | metric               |   FFNN |   XGBoost |   Transformer |
@@
 | True Negative Rate |   0.73 |      0.49 |          0.70 |
+
 : Model performance metrics (larger values are better) {#tbl-best-max-metrics}

### 2023-05-04 docs/hes_final/article.qmd
add citation, copy edit tweaks

@@
 The first uses text data enriched by country context features, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features.
-We demonstrate that usable conflict signals can be extracted directly from text sources, and we outline potential avenues for further research that we believe may yield models that can provide early warnings to decision makers to facilitate strategic planning, resource allocation, and disaster mitigation.
+We demonstrate that signals suggesting impending conflict can be extracted directly from text sources, and we outline potential avenues for further research that we believe may yield models that can provide early warnings to decision makers to facilitate strategic planning, resource allocation, and disaster mitigation.
 
@@
 Our methodology closely mirrors that which emerged in the following decade, including that described in more contemporary work, such as Mueller and Ruah [@RePEc:cam:camdae:2220].
-In essence, this methodology involves using time-lagged input data to predict some conflict label that corresponds to a date some time after the input data would have become available.
+In essence, this methodology involves using time-lagged input data to predict a conflict label that corresponds to a date some time after the input data would have become available.
 The model is then validated using an unseen holdout set, typically at the end of the date range available in the data [@doi:10.1080/03050629.2022.2070745].
@@
 The first approach used supervised dimensionality reduction and a feed-forward neural network conditioned with country context characteristics.
-In the second approach we used a fit-in-memory dataset of monthly embedding averages per country per month and applied statistical classification models such as K Nearest Neighbors and XGBoost.
+In the second approach we used a compact dataset of monthly embedding averages per country per month and applied statistical classification models such as K-Nearest Neighbors and XGBoost.
 In the third approach we stacked multiple article-level embeddings and applied Transformer-based models to extract a signal directly from the text corpus.
@@
 ### Defining Conflict Escalation
-Our task of modeling conflict escalation required a precise definition of what conflict escalation is.
-We began with raw event and casualty counts from The [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/), whose team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914].
+Our task of modeling conflict escalation required a precise definition of what conflict escalation means.
+We began with raw event and casualty counts from the [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/), who's team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914].
 ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and protests.
@@
 Approximately 60 attributes are assigned to each event from GDELT’s processing and analysis, including named entities involved, geolocation details, categorization among 300 event types, as well as metrics related to average ‘tone’ of articles about the event and potential severity of event impact.
-In 2021, GDELT introduced its Global Similarity Graph Document Embeddings dataset, which precomputes 512 dimension news-article-level embeddings using the Universal Sentence Encoder [@gdeltgsg].
+In 2021, GDELT introduced its Global Similarity Graph Document Embeddings dataset, which precomputes 512-dimension document-level embeddings of their scraped news articles using the Universal Sentence Encoder [@gdeltgsg] [@cer2018universal].
 
@@
 Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space.
-The fit UMAP model, was applied to the validation and test data without supervision.
-These reduced, 128 dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons.
+The fit UMAP model was applied to the validation and test data without supervision.
+These reduced, 128-dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons.
 The model's 4577 parameters were trained for 3 epochs.
@@
 
-Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total' from UNICEF's State of the World's Children report [@sowc]) and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges [@hdi], 4 categories of gross national income per capita [@wb], and 3 categories of Fragile States Index ranges[@fsi]).
+Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total' from UNICEF's State of the World's Children report [@sowc]) and three one-hot-encoded categorical indicators (categories of Human Development Index ranges [@hdi], categories of gross national income per capita [@wb], and categories of Fragile States Index ranges[@fsi]).
 
@@
 
-In this approach we prepared the dataset by averaging the embeddings per country per month.
+In this approach we prepared the dataset by averaging news article embeddings per country, per month.
 We also included one-hot encoded columns for countries, but no other country features were added.
 Since we only had access to embeddings starting in January 2020, the total length of this dataset was just a little over 6,000 samples (training, validation, and test sets combined).
-Given the high dimensionality of the data (512-dimensional embedding vectors + 700 dummy variables for countries), this dataset suffered from “curse of dimensionality”.
-This was handled by performing Principal Component Analysis – we found that reducing the number of dimensions to represent 80% of variance (~169 components with our selected train-test cutoff) improved the overall performance.
-Another issue was a class imbalance – conflict spike was present in only 14% of the training samples.
+Given the high dimensionality of the data (512-dimensional embedding vectors + 700 dummy variables for countries), this dataset suffered from the “curse of dimensionality”.
+This was handled by performing Principal Component Analysis; we found that reducing the number of dimensions to represent 80% of variance (~169 components with our selected train-test cutoff) improved the overall performance.
+Another issue was a class imbalance: conflict spike was present in only 14% of the training samples.
 We used random over-sampling technique to upsample the positive examples.
 
-The model used in this approach is a stacked model that averages probabilities independently predicted by a K Nearest Neighbors model and an XGBoost model.
+The model used in this approach is a stacked model that averages probabilities independently predicted by a K-Nearest Neighbors model and an XGBoost model.
 We used standard open-source versions of the underlying models.
@@
 
-Our third approach used a pair of transformer models to extract information directly from the GDELT embedding corpus.[@vaswani2017attention] Architecturally, both models are encoder-only transformers, but for simplicity, we will refer to the first of these models as “the encoder” and the second as the “classifier”.
+Our third approach used a pair of transformer models to extract information directly from the GDELT embedding corpus[@vaswani2017attention].
+Architecturally, both models are encoder-only transformers, but for simplicity, we will refer to the first of these models as “the encoder” and the second as the “classifier”.
 
@@
 This is roughly analogous to the role of a convolutional layer in a CNN.
-The classifier could then be used to make final predictions on the resulting distilled dataset.
+The classifier could then be used to make final predictions on the resulting, distilled dataset.
 
@@
 
-This new dataset offered major advantages in combatting the curse of dimensionality, however, it also represented a distorted view of the world, which overrepresents countries that receive more media coverage and underrepresents those that receive less, amplifying the biases already present in the GDELT dataset.
-Despite this limitation, this dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target.
+This new dataset offered major advantages in combatting the curse of dimensionality, however, it also presented a distorted view of the world which overrepresents countries that receive more media coverage and underrepresents those that receive less, amplifying the biases already present in the GDELT dataset.
+Despite this limitation, the dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target.
 Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons.
-The output of these 16 neurons would be used in the classification stage
+The output of these 16 neurons would be used in the classification stage.
 
@@
 This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target.
-In contrast to our earlier approaches, the dual-transformer model was trained purely on text embeddings.
+In contrast to our earlier approaches, the dual-transformer model was trained purely on news article text embeddings.
 The model was not conditioned by country, and no additional features were added beyond the text corpus itself.
@@
 
-Further investigation showed that while overall numeric metrics look similar, the Transformer-based text-only model generalized better 
+Further investigation showed that while overall numeric metrics look similar, the Transformer-based text-only model appeared to generalized better 
 across the globe. Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the 
@@
 The first and most apparent is bias in the training data.
-All of our models showed varying performance by region, and overreliance on such models by decisionmakers could lead to resources being diverted away from regions that need them, but which were underrepresented in the training data and thus had less accurate results.
+All of our models showed varying performance by region, and overreliance on such models by decision makers could lead to resources being diverted away from regions that need them, but which were underrepresented in the training data and thus had less accurate results.
 
 If models like these were to become widely used, they may be vulnerable to adversarial attacks.
-Large volumes of fake news content could be used to overwhelm the true signal of the model, obscuring the intentions of malicious actors, or causing panic.
+Large volumes of fake news content could be used to overwhelm the true signal from news articles to obscure the intentions of malicious actors, or to cause panic.
 Finally, and perhaps most concerningly, is the potential for conflict prediction models to deliver self-fulfilling prophesies.
@@
 They are measuring the temperature of the global media conversation, not gazing into the future.
-Decisionmakers need to understand that though models like these can be useful tools, they are not a substitute for careful deliberation and dialogue.
+Decision makers need to understand that though models like these can be useful tools, they are not a substitute for careful deliberation and dialogue.
 At their best, these models can provide an early warning for conflict, and if the right actions are taken, then they can be made to be wrong.
@@
 
-In this work we explored the possibility of predicting future geo-political conflict escalation at a country level using a global
-collection of news publications. We compared the usage of text-only data and data enriched with country labels and metadata. We
+In this work we explored the possibility of predicting future geopolitical conflict escalation at a country level using a global
+collection of news publications. We compared the usage of text-only data and data enriched with country labels and social indicators. We
 experimented with different ways of aggregating the data and applied both statistical machine learning and deep learning models. The

### 2023-05-04 docs/hes_final/article.qmd
add tex partial to show author affiliations

@@
       theme: neutral
+    keep-tex: true
 bibliography: bibliography.bibtex
+template-partials: 
+- title.tex
 # https://github.com/quarto-dev/quarto-cli/discussions/2786
@@
   - text: |
+      \usepackage[noblocks]{authblk}
+      \renewcommand*{\Authsep}{, }
+      \renewcommand*{\Authand}{, }
+      \renewcommand*{\Authands}{, }
+      \renewcommand\Affilfont{\small}
       \usepackage{float}

### 2023-05-04 docs/hes_final/article.qmd
add comment with link to source of author affiliation tip

@@
 bibliography: bibliography.bibtex
+# https://stackoverflow.com/a/75043433
 template-partials: 

### 2023-05-04 docs/hes_final/article.qmd
remove carriage returns
change title, add subtitle

--- 2023-05-04 docs/hes_final/article.qmd
@@
 ---
-title: "Predicting Conflict Escalation"
+title: Text Is All You Need
+subtitle: Predicting conflict escalation with news article text content
 author:
@@
   - name: Evan Wheeler
-    email: email@h
+    email: evw217@g.harvard.edu
     affiliation: Harvard University
@@
   pdf:
+    documentclass: scrartcl
     classoption:
@@
 
-We evaluated the models against a set of common performance metrics (see @tbl-best-max-metrics, @tbl-best-min-metrics, @fig-roc, 
-@fig-prc). Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches 
-produced comparative results. Each of the models outperformed the other two in a subset of these metrics.
+We evaluated the models against a set of common performance metrics (see @tbl-best-max-metrics, @tbl-best-min-metrics, @fig-roc, @fig-prc).
+Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results.
+Each of the models outperformed the other two in a subset of these metrics.
 
@@
 The closest approach used a presense or absense of any armed conflict related fatality as their binary target  [@10.1093/jeea/jvac025].
-They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in 
-peaceful countries) with their text based (LDA topic extraction) model. Our best ROC_AUC score is 69% (achieved by text + country labels 
-XGBoost/KNN model).
+They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in peaceful countries) with their text based (LDA topic extraction) model.
+Our best ROC_AUC score is 69% (achieved by text + country labels XGBoost/KNN model).
 
-We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false 
-positives. At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful. Given 
-these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total 
-number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem 
-to be most relevant metrics. The relative ranks of the models are the opposite for these two metrics with each of the two models
-with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
+We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives.
+At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful.
+Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics.
+The relative ranks of the models are the opposite for these two metrics with each of the two models with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
 
@@
 
-Further investigation showed that while overall numeric metrics look similar, the Transformer-based text-only model appeared to generalized better 
-across the globe. Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the 
-generalizability could be a reason to prefer Transformer architecture and resampled dataset.
+Further investigation showed that while overall numeric metrics look similar, the Transformer-based text-only model appeared to generalized better across the globe.
+Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the generalizability could be a reason to prefer Transformer architecture and resampled dataset.
 
-While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a 
-model that can assist UNICEF in their decision making. With that in mind, cost of deployment is also an important factor for model 
-evaluation. Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings 
-in two other proposed methods requires additional data engineering steps to be processed in batches.
+While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making.
+With that in mind, cost of deployment is also an important factor for model evaluation.
+Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings in two other proposed methods requires additional data engineering steps to be processed in batches.
 
-We think that all three approaches produced promising results and the model choice depends on the user's goals, selected performance 
-metrics and available resources.
+We think that all three approaches produced promising results and the model choice depends on the user's goals, selected performance metrics and available resources.
 
@@
 
-Building an application with a user interface could be a natural extension of this work. We envision a dashboard where a user can view a 
-world map where countries are heatmap colored according to the estimated probability of escalating conflict. Countries with highest risk 
-can be flagged or shown to a user in a separate view. In addition to the visual part of this application, model deployment for easy
-inference is an important project.
+Building an application with a user interface could be a natural extension of this work.
+We envision a dashboard where a user can view a world map where countries are heatmap colored according to the estimated probability of escalating conflict.
+Countries with highest risk can be flagged or shown to a user in a separate view.
+In addition to the visual part of this application, model deployment for easy inference is an important project.
 
-Building a more robust and automated data processing pipeline is another important direction. This would involve designing either a 
-batch or streaming system that can reliably collect data, preprocess and aggregate it into required format, train the model and make 
-predictions. The system should handle computational resources and cost constraints.
+Building a more robust and automated data processing pipeline is another important direction.
+This would involve designing either a batch or streaming system that can reliably collect data, preprocess and aggregate it into required format, train the model and make predictions.
+The system should handle computational resources and cost constraints.
 
@@
 
-In this work we explored the possibility of predicting future geopolitical conflict escalation at a country level using a global
-collection of news publications. We compared the usage of text-only data and data enriched with country labels and social indicators. We
-experimented with different ways of aggregating the data and applied both statistical machine learning and deep learning models. The
-results suggest that text embeddings provide a strong enough signal to make such predictions. Different data
-sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common
-numetic metrics. Transformer-based model generalized better across geographies while statistical model provided a size/cost of
-deployment advantage.
+In this work we explored the possibility of predicting future geopolitical conflict escalation at a country level using a global collection of news publications.
+We compared the usage of text-only data and data enriched with country labels and social indicators.
+We experimented with different ways of aggregating the data and applied both statistical machine learning and deep learning models.
+The results suggest that text embeddings provide a strong enough signal to make such predictions.
+Different data sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common numetic metrics.
+Transformer-based model generalized better across geographies while statistical model provided a size/cost of deployment advantage.
 
-While this work did not produce a single best model for conflict escalation predictions, it demonstrated the possibility of such
-predictions from primarily text data. We provided a clear definition of a binary conflict escalation target and described several
-approaches for data aggregation and modeling. We hope this work will contribute to both UNICEF operations and political science
-research. The performance results can serve as a benchmark for future research projects in this
-field.
+While this work did not produce a single best model for conflict escalation predictions, it demonstrated the possibility of such predictions from primarily text data.
+We provided a clear definition of a binary conflict escalation target and described several approaches for data aggregation and modeling.
+We hope this work will contribute to both UNICEF operations and political science research.
+The performance results can serve as a benchmark for future research projects in this field.
 

### 2023-05-04 docs/hes_final/article.qmd
whitespace fixes

@@
 
-Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total' from UNICEF's State of the World's Children report [@sowc]) and three one-hot-encoded categorical indicators (categories of Human Development Index ranges [@hdi], categories of gross national income per capita [@wb], and categories of Fragile States Index ranges[@fsi]).
+Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total' from UNICEF's State of the World's Children report [@sowc]) and three one-hot-encoded categorical indicators (categories of Human Development Index ranges [@hdi], categories of gross national income per capita [@wb], and categories of Fragile States Index ranges [@fsi]).
 
@@
 
-Our third approach used a pair of transformer models to extract information directly from the GDELT embedding corpus[@vaswani2017attention].
+Our third approach used a pair of transformer models to extract information directly from the GDELT embedding corpus [@vaswani2017attention].
 Architecturally, both models are encoder-only transformers, but for simplicity, we will refer to the first of these models as “the encoder” and the second as the “classifier”.

### 2023-05-04 docs/hes_final/article.qmd
update metric tables based on updated ffnn preds

@@
 ::: {layout="[[1], [1]]"}
+
 | metric               |   FFNN |   XGBoost |   Transformer |
 |:---------------------|-------:|----------:|--------------:|
-| False Discovery Rate |   0.74 |      0.83 |          0.80 |
+| False Discovery Rate |   0.73 |      0.83 |          0.80 |
 | False Negative Rate  |   0.59 |      0.28 |          0.53 |
-| False Positive Rate  |   0.27 |      0.51 |          0.30 |
+| False Positive Rate  |   0.19 |      0.51 |          0.30 |
 
@@
 |:-------------------|-------:|----------:|--------------:|
-| Cohen Kappa Score  |   0.12 |      0.09 |          0.11 |
-| F1 Score           |   0.32 |      0.28 |          0.28 |
-| Precision Score    |   0.26 |      0.17 |          0.20 |
+| Cohen Kappa Score  |   0.18 |      0.09 |          0.11 |
+| F1 Score           |   0.33 |      0.28 |          0.28 |
+| Precision Score    |   0.27 |      0.17 |          0.20 |
 | Recall Score       |   0.41 |      0.72 |          0.47 |
-| Roc Auc Score      |   0.64 |      0.69 |          0.64 |
-| True Negative Rate |   0.73 |      0.49 |          0.70 |
+| Roc Auc Score      |   0.67 |      0.69 |          0.64 |
+| True Negative Rate |   0.81 |      0.49 |          0.70 |
 

### 2023-05-04 docs/hes_final/article.qmd
Use images for tables and rearrange images

@@
 
-We evaluated the models against a set of common performance metrics (see @tbl-best-max-metrics, @tbl-best-min-metrics, @fig-roc, @fig-prc).
-Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results.
-Each of the models outperformed the other two in a subset of these metrics.
+We evaluated the models against a set of common performance metrics (see @fig-best-max-metrics, @fig-best-min-metrics, @fig-roc, 
+@fig-prc). Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches 
+produced comparative results. Each of the models outperformed the other two in a subset of these metrics.
 
@@
 
-::: {layout="[[1], [1]]"}
+::: {layout="[[1, 1]]"}
 
-| metric               |   FFNN |   XGBoost |   Transformer |
-|:---------------------|-------:|----------:|--------------:|
-| False Discovery Rate |   0.73 |      0.83 |          0.80 |
-| False Negative Rate  |   0.59 |      0.28 |          0.53 |
-| False Positive Rate  |   0.19 |      0.51 |          0.30 |
+![Model performance metrics (larger values are better)](imgs/fig-best-max-metrics.png){#fig-best-max-metrics}
 
-: Model performance metrics (smaller values are better) {#tbl-best-min-metrics}
-
-| metric             |   FFNN |   XGBoost |   Transformer |
-|:-------------------|-------:|----------:|--------------:|
-| Cohen Kappa Score  |   0.18 |      0.09 |          0.11 |
-| F1 Score           |   0.33 |      0.28 |          0.28 |
-| Precision Score    |   0.27 |      0.17 |          0.20 |
-| Recall Score       |   0.41 |      0.72 |          0.47 |
-| Roc Auc Score      |   0.67 |      0.69 |          0.64 |
-| True Negative Rate |   0.81 |      0.49 |          0.70 |
-
-: Model performance metrics (larger values are better) {#tbl-best-max-metrics}
+![Model performance metrics (smaller values are better)](imgs/fig-best-min-metrics.png){#fig-best-min-metrics}
 
 :::
-\twocolumn
-
-\onecolumn
 

### 2023-05-04 docs/hes_final/article.qmd
Actually push images for tables

@@
 
-We evaluated the models against a set of common performance metrics (see @fig-best-max-metrics, @fig-best-min-metrics, @fig-roc, 
-@fig-prc). Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches 
+We evaluated the models against a set of common performance metrics (see @fig-best-max-metrics, @fig-best-min-metrics, @fig-roc). 
+Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches 
 produced comparative results. Each of the models outperformed the other two in a subset of these metrics.
@@
 
-\onecolumn
-
-::: {layout="[[1, 1]]"}
-
 ![Model performance metrics (larger values are better)](imgs/fig-best-max-metrics.png){#fig-best-max-metrics}
@@
 
-:::
-
-::: {layout="[[1, 1]]"}
-
 ![Receiver Operator Characteristics](imgs/fig-roc.png){#fig-roc}
 
-![Precision Recall Curve](imgs/fig-prc.png){#fig-prc}
-
-:::
-::: {layout="[[1, 1]]"}
+Further investigation showed that while overall numeric metrics look similar, the Transformer-based text-only model appeared to generalized better across the globe.
+Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the generalizability could be a reason to prefer Transformer architecture and resampled dataset.
 
@@
 
-:::
-::: {layout="[[1, 1]]"}
-
 ![Transformer model accuracy on test data](imgs/fig-transformer-test-accuracy.png){#fig-transformer-test-accuracy}
-
-:::
-\twocolumn
-
-Further investigation showed that while overall numeric metrics look similar, the Transformer-based text-only model appeared to generalized better across the globe.
-Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the generalizability could be a reason to prefer Transformer architecture and resampled dataset.
 

### 2023-05-04 docs/hes_final/article.qmd
update metrics table imgs

@@
 
-We evaluated the models against a set of common performance metrics (see @fig-best-max-metrics, @fig-best-min-metrics, @fig-roc). 
-Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches 
-produced comparative results. Each of the models outperformed the other two in a subset of these metrics.
+We evaluated the models against a set of common performance metrics (see @fig-best-min-metrics, @fig-best-max-metrics, @fig-roc).
+Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results.
+Each of the models outperformed the other two in a subset of these metrics.
+
+![Model metrics (smaller values are better)](imgs/fig-best-min-metrics.png){#fig-best-min-metrics}
+
+![Model metrics (large values are better)](imgs/fig-best-max-metrics.png){#fig-best-max-metrics}
 
@@
 
+![Receiver Operator Characteristics](imgs/fig-roc.png){#fig-roc}
+
 We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives.
@@
 The relative ranks of the models are the opposite for these two metrics with each of the two models with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
-
-![Model performance metrics (larger values are better)](imgs/fig-best-max-metrics.png){#fig-best-max-metrics}
-
-![Model performance metrics (smaller values are better)](imgs/fig-best-min-metrics.png){#fig-best-min-metrics}
-
-![Receiver Operator Characteristics](imgs/fig-roc.png){#fig-roc}
-
-Further investigation showed that while overall numeric metrics look similar, the Transformer-based text-only model appeared to generalized better across the globe.
-Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the generalizability could be a reason to prefer Transformer architecture and resampled dataset.
 
@@
 
+Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world.
 While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making.
@@
 The results suggest that text embeddings provide a strong enough signal to make such predictions.
-Different data sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common numetic metrics.
-Transformer-based model generalized better across geographies while statistical model provided a size/cost of deployment advantage.
+Different data sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common numeric metrics.
+The neural network-based models appear to generalize better across geographies while statistical model provided a size/cost of deployment advantage.
 

### 2023-05-04 docs/hes_final/article.qmd
spelling corrections

@@
 While we believed this to be the best dataset available for our purposes, prior research has revealed some limitations that we must consider.
-GDELT’s coverage overrepresents the western world, particularly the United States and Europe.
+GDELT’s coverage over-represents the western world, particularly the United States and Europe.
 GDELT does represent Russia, India and China better than other event datasets, but it continues to underrepresent other parts of the world, particularly the global south.
@@
 flowchart TD
-    GSG[GDELT Global Similarty Graph] -->|sample 20% of daily files| Merge{Merge on article URL,\n select relevant features}
+    GSG[GDELT Global Similarity Graph] -->|sample 20% of daily files| Merge{Merge on article URL,\n select relevant features}
     GKG[GDELT Global Knowledge Graph] -->|sample 20% of daily files| Merge
@@
 
-This new dataset offered major advantages in combatting the curse of dimensionality, however, it also presented a distorted view of the world which overrepresents countries that receive more media coverage and underrepresents those that receive less, amplifying the biases already present in the GDELT dataset.
+This new dataset offered major advantages in combatting the curse of dimensionality, however, it also presented a distorted view of the world which over-represents countries that receive more media coverage and underrepresents those that receive less, amplifying the biases already present in the GDELT dataset.
 Despite this limitation, the dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target.
@@
 Our target was defined differently from related work which made direct comparison less meaningful.
-The closest approach used a presense or absense of any armed conflict related fatality as their binary target [@10.1093/jeea/jvac025].
+The closest approach used a presence or absence of any armed conflict related fatality as their binary target [@10.1093/jeea/jvac025].
 They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in peaceful countries) with their text based (LDA topic extraction) model.
@@
 The first and most apparent is bias in the training data.
-All of our models showed varying performance by region, and overreliance on such models by decision makers could lead to resources being diverted away from regions that need them, but which were underrepresented in the training data and thus had less accurate results.
+All of our models showed varying performance by region, and over-reliance on such models by decision makers could lead to resources being diverted away from regions that need them, but which were underrepresented in the training data and thus had less accurate results.
 
@@
 Large volumes of fake news content could be used to overwhelm the true signal from news articles to obscure the intentions of malicious actors, or to cause panic.
-Finally, and perhaps most concerningly, is the potential for conflict prediction models to deliver self-fulfilling prophesies.
+Finally, and perhaps most concerning, is the potential for conflict prediction models to deliver self-fulfilling prophesies.
 If policy makers become convinced of impending conflict in their own or neighboring countries because of the predictions of an AI model, they may take actions that make conflict more likely.
@@
 Building an application with a user interface could be a natural extension of this work.
-We envision a dashboard where a user can view a world map where countries are heatmap colored according to the estimated probability of escalating conflict.
+We envision a dashboard where a user can view a world map where countries are colored according to the estimated probability of escalating conflict.
 Countries with highest risk can be flagged or shown to a user in a separate view.

### 2023-05-05 docs/hes_final/article.qmd

@@
 Different data sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common numeric metrics.
-The neural network-based models appear to generalize better across geographies while statistical model provided a size/cost of deployment advantage.
+The neural network-based models appear to generalize better across geographies while the statistical model provided a size/cost of deployment advantage.
 

### 2023-05-05 docs/hes_final/article.qmd
tweak letter caps
Title Tweak and Emails

@@
 title: Text Is All You Need
-subtitle: Predicting conflict escalation with news article text content
+subtitle: Predicting Conflict Escalation Using Global News Content
 author:
   - name: Diana Kolusheva
-    email: email@h
+    email: dik431@g.harvard.edu
     affiliation: Harvard University
   - name: Chad Stoughton
-    email: email@h
+    email: ch0793@g.harvard.edu
     affiliation: Harvard University

### 2023-05-05 docs/hes_final/article.qmd
Minor Grammar edits

@@
 The first approach used supervised dimensionality reduction and a feed-forward neural network conditioned with country context characteristics.
-In the second approach we used a compact dataset of monthly embedding averages per country per month and applied statistical classification models such as K-Nearest Neighbors and XGBoost.
-In the third approach we stacked multiple article-level embeddings and applied Transformer-based models to extract a signal directly from the text corpus.
+In the second approach, we used a compact dataset of monthly embedding averages per country per month and applied statistical classification models such as K-Nearest Neighbors and XGBoost.
+In the third approach we stacked multiple article-level embeddings and applied transformer-based models to extract a signal directly from the text corpus.
 See the @fig-ffnn, @fig-xgboost, and @fig-transformer for details of modeling approaches.
@@
 
-Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world.
+Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world—see @fig-ffnn-test-accuracy, @fig-xgboost-test-accuracy, and @fig-transformer-test-accuracy.
 While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making.
@@
 Large volumes of fake news content could be used to overwhelm the true signal from news articles to obscure the intentions of malicious actors, or to cause panic.
+
 Finally, and perhaps most concerning, is the potential for conflict prediction models to deliver self-fulfilling prophesies.
@@
 
-Due to time and resource constraints, our work used only 20% of the daily embeddings available through  GDELT, and future research could focus on scaling the methods outline.
-We also believe that continued experimentation with data sampling and aggregation approaches, as well as model tuning, could yield models that outperform those outlined here without additional data.
+Future research could focus on scaling the methods outline.
+Due to time and resource constraints, our work used only 20% of the daily embeddings available through  GDELT, and increasing the scale of training data could yield better results.
+We also believe that continued experimentation with data sampling and aggregation approaches, as well as model tuning, could produce models that outperform those outlined here without additional data.
 
@@
 
-Building an application with a user interface could be a natural extension of this work.
+Building an application with a user interface could be a natural extension of this research.
 We envision a dashboard where a user can view a world map where countries are colored according to the estimated probability of escalating conflict.

### 2023-05-05 docs/hes_final/article.qmd
Minor Grammar edits redux
simplified author block

@@
 # https://stackoverflow.com/a/75043433
-template-partials: 
-- title.tex
+#template-partials: 
+#- title.tex
 # https://github.com/quarto-dev/quarto-cli/discussions/2786
@@
   - text: |
-      \usepackage[noblocks]{authblk}
-      \renewcommand*{\Authsep}{, }
-      \renewcommand*{\Authand}{, }
-      \renewcommand*{\Authands}{, }
-      \renewcommand\Affilfont{\small}
       \usepackage{float}
@@
 We evaluated the models against a set of common performance metrics (see @fig-best-min-metrics, @fig-best-max-metrics, @fig-roc).
-Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results.
+Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparable results.
 Each of the models outperformed the other two in a subset of these metrics.

### 2023-05-05 docs/hes_final/article.qmd
Two typo corrections

@@
 Modern methodologies that resemble those we outline here began to appear in the literature in the early 2010s, when researchers began using holdout datasets for model validation [@10.2307/40730711].
-This shift followed work by Ward et al that sharply criticized the use of statistical significance tests on in-sample prediction to validate conflict models [@doi:10.1177/0022343309356491].
+This shift followed work by Ward et al.
+that sharply criticized the use of statistical significance tests on in-sample prediction to validate conflict models [@doi:10.1177/0022343309356491].
 Our methodology closely mirrors that which emerged in the following decade, including that described in more contemporary work, such as Mueller and Ruah [@RePEc:cam:camdae:2220].
@@
 Our task of modeling conflict escalation required a precise definition of what conflict escalation means.
-We began with raw event and casualty counts from the [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/), who's team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914].
+We began with raw event and casualty counts from the [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/), whose team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914].
 ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and protests.

### 2023-05-07 docs/hes_final/presentation.qmd
add presentation draft

---
title: Text Is All You Need
subtitle: Predicting Conflict Escalation Using Global News Content
author:
  - name: Diana Kolusheva
    email: dik431@g.harvard.edu
    affiliation: Harvard University
  - name: Chad Stoughton
    email: ch0793@g.harvard.edu
    affiliation: Harvard University
  - name: Evan Wheeler
    email: evw217@g.harvard.edu
    affiliation: Harvard University
lang: en
format:
  revealjs:
    incremental: true

---
## Introduction
- The task of forecasting conflict escalation is of great interest to policy makers, executives, and social scientists. 
- We introduce a conflict escalation definition and a novel binary classification metric.
- We outline 3 methods to extract generalizable conflict signals directly from a corpus of over 2.5 million document embeddings derived from global news sources using contemporary natural language processing and machine learning methods.

## Conflict forecasting

- Predicting conflict has long been of interest to researchers and practitioners of international relations, and statistical and computational methods to do so can be found in the literature as early as 1944. 
- Modern methodologies that resemble those we outline here began to appear in the literature in the early 2010s, when researchers began using holdout datasets for model validation. 
- This shift followed work by Ward et al. that sharply criticized the use of statistical significance tests on in-sample prediction to validate conflict models.

## Challenge of global models

- Today, much of the work in this area has continued to focus on country or region-specific models to predict conflict. 
- This introduces a “small-data problem” where relatively few observations of conflict are available for a given country. 
- Some efforts have been made to overcome this by building generalizable models that can theoretically be used to make predictions on any country.

:::{.notes}
- We followed the latter approach in developing our model, hypothesizing that the larger number of observations available to a global model will allow it to extract more generalizable signals of conflict, and that such a model will be better able to predict unforeseen risks in countries with little available training data.
:::

## Methodology

- Our approach is text-first, making predictions primarily or exclusively from text, with few or no other features added.
- To the best of our knowledge, no conflict prediction model described in the literature makes use of text embeddings.
- Previous efforts have included statistical analysis of text, and unsupervised Latent Dirichlet Allocation (LDA) for topic extraction.

## Three approaches to extracting conflict signals from text embeddings

- The first approach used supervised dimensionality reduction and a feed-forward neural network conditioned with country context characteristics.
- In the second approach, we used a compact dataset of monthly embedding averages per country per month and applied statistical models such as K-Nearest Neighbors and XGBoost.
- In the third approach we stacked multiple article-level embeddings and applied transformer-based models to extract a signal exclusively from the text corpus.

## Defining Conflict Escalation

- Our task of modeling conflict escalation required a precise definition of what conflict escalation means. 
- We began with raw event and casualty counts from the [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/), whose team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities .

::: {.aside}
- Prior work has treated this as a binary classification task, asking whether or not conflict escalated in a given month, and we adopt this approach as well.
- In contrast to prior work, however, we do not treat the simple presence of a fatality due to conflict as sufficient to indicate conflict escalation. 
:::

::: {.notes}
- ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and protests. With these totals in-hand, we needed to create a target to label our training data. 
- To achieve our objective of creating a generalizable conflict model, we required a metric that was consistent across varying local conditions. 
- Small changes in the absolute number of monthly fatalities can have widely different implications depending on the country and time period. 
:::

## Introducing our 'spike' conflict metric

- We derived a metric designed to highlight relative “spikes” in political violence. Calculating whether a spike occurred in a given month is a two-step process. 
- First, we calculated the trend in violence for a given month, defined by the slope of OLS regression of the number of ACLED-documented fatalities in target month & preceding 2 months.
- Next, we calculated whether that slope was in the 75th percentile of such slopes in the target country for the preceding 2 years.

:::{.notes}
This trend allowed us to estimate whether violence was increasing or decreasing in the target country at any time.
:::
---

![The Spike metric applied to Ukraine](imgs/Ukraine_Spikes.png){#fig-spike}

- Our 'spike' binary metric is sensitive to small increases in political violence in peaceful countries/periods, while also filtering out small fluctuations in long-standing conflicts.

:::{.notes}

- @fig-spike shows how our metric captures conflict fluctuations in Ukraine. Prior to the Russian invasion of February 2022, small fluctuations trigger a spike. However, much larger fluctuations in the period following the invasion are filtered out, as the spike formula adjusts to the higher level of baseline conflict.

- The increase in violence during the Ukrainian counter-offensive in the late summer again triggers a spike, and this reflects significant changes in the situation on the ground.
:::

## News Content Data

- The [Global Database of Events Language and Tone (GDELT) Project](https://www.gdeltproject.org/) provides public APIs of current events derived from news articles scraped from sources around the world . 
- In 2021, GDELT introduced its Global Similarity Graph Document Embeddings dataset, which precomputes 512-dimension document-level embeddings of their scraped news articles using the Universal Sentence Encoder.

:::{.notes}
- GDELT’s coverage over-represents the western world, particularly the United States and Europe.
- GDELT does represent Russia, India and China better than other event datasets, but it continues to underrepresent other parts of the world, particularly the global south.
- English is the most common language in the GDELT dataset, comprising more than 40% of the articles in its corpus.
:::

## Training Dataset

::: {#fig-dgp layout-ncol=1}
```{mermaid}
%%| fig-width: 5
%%| fig-height: 5
flowchart TD
    GSG[GDELT Global Similarity Graph] -->|sample 20% of daily files| Merge{Merge on article URL,\n select relevant features}
    GKG[GDELT Global Knowledge Graph] -->|sample 20% of daily files| Merge

    Merge --> GDELT(GDELT dataset)

    ACLED[ACLED monthly conflict totals] --> Spike{Calculate our lagged\n 'spike' binary variable}

    GDELT --> Data(Time-partitioned train, val, & test datasets)
    Spike --> Data
```
Data generation process
:::

:::{.notes}
- Our training data was derived from these GDELT and ACLED datasets. 
- GDELT news article embeddings served as our primary independent variable, while the “spike” metric derived from ACLED’s conflict data served as our target variable.
- The spike metric was used to label article data on a three-month time lag, allowing us to make predictions of future events.
- Our training dataset thus ran from January 2020 through October 2022. Data from November 2022 through March 2023 was used for model validation and testing.
:::

## Conflict Prediction Using Text and Country Features

- Our first approach employed a feed forward neural network model. 
- Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner.

---

![Scatterplot of 2-dimension UMAP-reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){#fig-umap2d-ds-targets}

:::{.notes}
- Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space. 
- The fit UMAP model was applied to the validation and test data without supervision. These reduced, 128-dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons. The model's 4577 parameters were trained for 3 epochs.
:::

## Additional country context inputs

- 2 numerical indicators from UNICEF's State of the World's Children report: 'Under-five mortality rate 2019' & 'Adolescent population 2020 Proportion of total population (%) Total'
- 3 one-hot-encoded categorical indicators: Human Development Index ranges, categories of gross national income per capita, & categories of Fragile States Index ranges)

:::{.notes}
- avoid the per-country 'small data problem' by allowing model to learn similarities between country contexts and generalize
:::

---

::: {#fig-ffnn layout-ncol=1}
```{mermaid}
%%| fig-width: 5
%%| fig-height: 5
flowchart TD
    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)

    Articles --> UMAP{Supervised UMAP}
    UMAP --> Compact(128 dimension representations)

    Compact --> FFNN{Feed Forward Neural Network}
    FFNN --> Pred(Prediction of conflict escalation in 3 months)

    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
    Spike(Conflict response binary variable) --> |included during training|UMAP

    Spike -->FFNN

    Context[Country context indicators] --> FFNN
```
Feed forward neural network model architecture
:::

## Conflict Prediction Using Text and Country Labels

- In this approach we prepared the dataset by averaging news article embeddings per country, per month. We also included one-hot encoded columns for countries, but no other country features were added. 
- Given the high dimensionality of the data (512-dimensional embedding vectors + 700 dummy variables for countries), this dataset suffered from the “curse of dimensionality”.

---

::: {#fig-xgboost layout-ncol=1}
```{mermaid}
%%| fig-width: 5
%%| fig-height: 5
flowchart TD
    GDELT[GDELT Embeddings] --> |average by country by month| AGG(Monthly news representation)
    Countries[Country labels] --> |one-hot encoding| AGG

    AGG --> |PCA| PCA(169 dimension representation)
    PCA --> KNN{KNN}
    PCA --> XGD{XGBoost}

    XGD --> AVG{Mean}
    KNN --> AVG
    AVG --> Pred(Prediction of conflict escalation in 3 months)

    ACLED[Monthly ACLED fatality counts] --> |compute our 'spike' variable| Spike(Conflict response binary variable)
    Spike --> XGD
    Spike --> KNN
```
Statistical model architecture
:::

:::{.notes}
- This was handled by performing Principal Component Analysis; we found that reducing the number of dimensions to represent 80% of variance (~169 components with our selected train-test cutoff) improved the overall performance.
- Another issue was a class imbalance: conflict spike was present in only 14% of the training samples. We used random over-sampling technique to upsample the positive examples.
- The model used in this approach is a stacked model that averages probabilities independently predicted by a K-Nearest Neighbors model and an XGBoost model. We used standard open-source versions of the underlying models. The intention behind averaging the predicted probabilities was to reduce the impact of individual model weaknesses. 
:::

## Conflict Prediction Using Text Alone

- Our third approach used a pair of encoder-only transformers, but for simplicity, we will refer to the first of these models as “the encoder” and the second as the “classifier”.
- The purpose of the encoder was to reduce the dimensionality of the embedding corpus by analyzing a subset of article embeddings and returning a single vector of substantially reduced size. This is roughly analogous to the role of a convolutional layer in a CNN. 
- The classifier could then be used to make final predictions on the resulting, distilled dataset.

---

- To prepare the training dataset, our full sample of the GDELT corpus was split into a single $512\times n$ matrix for each country and month where $n$ is the number of articles published in that month related to that country.
- Each of these matrices was then split into several random subsamples of 50 articles each. 
- Samples for countries in which $n<50$ were padded to $512\times50$.

:::{.notes}

- By limiting the size of each sample to 50 articles, we reduced the dimensionality of each sample.
- By splitting the corpus into multiple unique training examples per country per month, we were able to substantially increase the size of the training dataset.
- This new dataset offered major advantages in combatting the curse of dimensionality, however, it also presented a distorted view of the world which over-represents countries that receive more media coverage and underrepresents those that receive less, amplifying the biases already present in the GDELT dataset. 
- Despite this limitation, the dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target. 
:::

---

![Reducing Embedding Dimensionality with Transformer Encoder](imgs/Transformer_Viz.png){#fig-transformer-viz}

- Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons. 
- The output of these 16 neurons would be used in the classification stage.

---

- With the encoder's training complete, we returned to the original base dataset. Each country-month matrix was again split into $512\times50$ matrices.
- Each of these samples could then be passed into the encoder, which distilled it into a $length-16$ vector.
- Each vector was then stacked, resulting in a single training example for each country and month of size $16\times\frac{n}{50}$.

:::{.notes}
- This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target. 
- In contrast to our earlier approaches, the dual-transformer model was trained purely on news article text embeddings. The model was not conditioned by country, and no additional features were added beyond the text corpus itself.
:::

---

::: {#fig-transformer layout-ncol=1}
```{mermaid}
%%| fig-width: 5
%%| fig-height: 5
flowchart TD
    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)

    Articles --> Trans1{Encoder\n transformer}
    Trans1 -->|classifier head present during training| Compact(16 dimension representations)

    Compact --> Trans2{Binary\n classifier\n transformer}
    Trans2 --> Pred(Prediction of conflict escalation in 3 months)

    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
    Spike(Conflict response\n binary variable) --> Trans2

    Spike -->|included during model training| Trans1
```
Transformer model architecture
:::

## Results

- We evaluated the models against a set of common performance metrics (see @fig-best-min-metrics, @fig-best-max-metrics, @fig-roc). 
- Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparable results.
- Each of the models outperformed the other two in a subset of these metrics.

---

- We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives. At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful. 
- Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics.

---

![Model metrics (large values are better)](imgs/fig-best-max-metrics.png){#fig-best-max-metrics}

:::{.notes}
- The relative ranks of the models are the opposite for these two metrics with each of the two models with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
:::

---

![Model metrics (smaller values are better)](imgs/fig-best-min-metrics.png){#fig-best-min-metrics}

---

![Receiver Operator Characteristics](imgs/fig-roc.png){#fig-roc}

:::{.notes}
- Our target was defined differently from related work which made direct comparison less meaningful.
- The closest approach used a presence or absence of any armed conflict related fatality as their binary target. They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in peaceful countries) with their text based (LDA topic extraction) model. Our best ROC_AUC score is 69% (achieved by text + country labels XGBoost/KNN model).
:::

---

![FFNN model accuracy on test data](imgs/fig-ffnn-test-accuracy.png){#fig-ffnn-test-accuracy}

![XGBoost model accuracy on test data](imgs/fig-xgboost-test-accuracy.png){#fig-xgboost-test-accuracy}

![Transformer model accuracy on test data](imgs/fig-transformer-test-accuracy.png){#fig-transformer-test-accuracy}

:::{.notes}
- Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world—see @fig-ffnn-test-accuracy, @fig-xgboost-test-accuracy, and @fig-transformer-test-accuracy. 
:::

---

- While performance is often the main objective when developing prediction models, our practical goal was a tool that can assist UNICEF, so cost & complexity is an important consideration.
- Average monthly embeddings dataset & corresponding statistical model fit in memory, whereas article-level embeddings in our other methods require additional data engineering steps to be processed in batches.
- All three approaches produced promising results so the model choice depends on the user's goals, priority performance metrics, and available resources.

## Ethical Considerations
- This work carries significant ethical considerations, especially if it were to be operationalized. 
- First is bias in the training data: all of our models showed varying performance by region, and over-reliance on such models by decision makers could lead to resources being diverted away from regions that need them--but which were underrepresented in the training data and thus had less accurate results.

---

- If models like these were to become widely used, they may be vulnerable to adversarial attacks. Large volumes of fake news content could be used to overwhelm the true signal from news articles to obscure the intentions of malicious actors, or to cause panic. 
- Finally, and perhaps most concerning, is the potential for conflict prediction models to deliver self-fulfilling prophesies. If policy makers become convinced of impending conflict in their own or neighboring countries because of the predictions of an AI model, they may take actions that make conflict more likely.

---

- Reducing these risks requires emphasizing that these are models, not oracles. They are measuring the temperature of the global media conversation, not gazing into the future.
- Decision makers need to understand that though models like these can be useful tools, they are not a substitute for careful deliberation and dialogue.
- At their best, these models can provide an early warning for conflict, and if the right actions are taken, then they can be made to be wrong.

## Future work

- We foresee several ways to extend this work, and there are exciting opportunities for both further research and operationalizations.

## Future research directions

- Our work used only 20% of the daily embeddings available through GDELT: increasing size of training sets could yield better results. 
- Continued experimentation with data sampling and aggregation approaches, as well as model tuning, could produce models that outperform those outlined here without additional training data. 
- Finally, we believe deeper research into regional disparities in model performance is needed. Understanding the limitations of global media coverage for conflict prediction will be critical to any effort to operationalize this work.

## Conclusions

- In this work we explored the possibility of predicting future geopolitical conflict escalation at a country level using a global collection of news publications. We compared the usage of text-only data and data enriched with country labels and social indicators. 
- We introduce a conflict escalation definition and a novel binary classification metric.

---

- The results suggest that text embeddings provide a strong enough signal to make such predictions. 
- Different data sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common numeric metrics.

---

- The neural network-based models appear to generalize better across geographies while the statistical model provided a size/cost of deployment advantage.
- While this work did not produce a single best model for conflict escalation predictions, it demonstrated the possibility of such predictions from primarily text data. 
- We described several approaches for data aggregation and modeling. We hope this work will contribute to both UNICEF operations and political science research. 
- The performance results can serve as a benchmark for future research projects in this field.

### 2023-05-10 docs/hes_final/presentation.qmd
Trimmed Ethics Section Language

@@
 ## Ethical Considerations
-- This work carries significant ethical considerations, especially if it were to be operationalized. 
-- First is bias in the training data: all of our models showed varying performance by region, and over-reliance on such models by decision makers could lead to resources being diverted away from regions that need them--but which were underrepresented in the training data and thus had less accurate results.
+- Bias in the training data
+- Adversarial attacks
+- Self-fulfilling prophesies
+
+:::{.notes}
+- All models showed varying performance by region, could lead to resources being misallocated. 
+- Large volumes of fake news could overwhelm true signal, obscure intentions of malicious actors & cause panic.
+- AI predictions of impending conflict may lead policy makers to take actions that make conflict more likely. 
+:::
 
@@
 
-- If models like these were to become widely used, they may be vulnerable to adversarial attacks.
-Large volumes of fake news content could be used to overwhelm the true signal from news articles to obscure the intentions of malicious actors, or to cause panic.
-- Finally, and perhaps most concerning, is the potential for conflict prediction models to deliver self-fulfilling prophesies.
-If policy makers become convinced of impending conflict in their own or neighboring countries because of the predictions of an AI model, they may take actions that make conflict more likely.
+- Models, not oracles 
+- Not a substitute for careful deliberation and dialogue
+- At their best, these models can provide an early warning for conflict, and if the right actions are taken, then they can be made to be wrong
 
----
-
-- Reducing these risks requires emphasizing that these are models, not oracles. They are measuring the temperature of the global media conversation, not gazing into the future.
-- Decision makers need to understand that though models like these can be useful tools, they are not a substitute for careful deliberation and dialogue.
-- At their best, these models can provide an early warning for conflict, and if the right actions are taken, then they can be made to be wrong.
+:::{.notes}
+- Measuring the temperature of the global media conversation, not gazing into the future.
+:::
 

### 2023-05-10 docs/hes_final/presentation.qmd
Transformer Slides First Pass

@@
 
-- Our third approach used a pair of encoder-only transformers, but for simplicity, we will refer to the first of these models as “the encoder” and the second as the “classifier”.
-- The purpose of the encoder was to reduce the dimensionality of the embedding corpus by analyzing a subset of article embeddings and returning a single vector of substantially reduced size.
-This is roughly analogous to the role of a convolutional layer in a CNN.
+- Our third approach used a pair of encoder-only transformers
+- the “encoder” and the “classifier”.
+- The encoder was used to reduce the dimensionality of the embedding corpus by analyzing a subset of article embeddings and returning a single vector of substantially reduced size.
+- Analogous to the role of a convolutional layer in a CNN. 
 - The classifier could then be used to make final predictions on the resulting, distilled dataset.
@@
 
-- To prepare the training dataset, our full sample of the GDELT corpus was split into a single $512\times n$ matrix for each country and month where $n$ is the number of articles published in that month related to that country.
-- Each of these matrices was then split into several random subsamples of 50 articles each. 
-- Samples for countries in which $n<50$ were padded to $512\times50$.
+:::: {.columns}
+
+::: {.column width="50%"}
+![Transformer Model](imgs/Transformer_Viz.png){#fig-transformer-viz}
+:::
+
+::: {.column width="50%"}
+Training the Encoder
+
+- Each country/month sample was split into subsamples of 50 articles each
+- The same label was used for each subsample, creating multiple unique training samples per country per month 
+- This produced a larger, but more distorted corpus
+:::
+
+::::
 
@@
 
-![Reducing Embedding Dimensionality with Transformer Encoder](imgs/Transformer_Viz.png){#fig-transformer-viz}
+:::: {.columns}
+
+::: {.column width="50%"}
+![Transformer Model](imgs/Transformer_Viz.png){#fig-transformer-viz}
+:::
+
+::: {.column width="50%"}
+Training the Classifier
 
 - Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons. 
-- The output of these 16 neurons would be used in the classification stage.
+- Each vector was then stacked, resulting in a single training example for each country and month of size $16\times\frac{n}{50}$. 
+:::
 
----
-
-- With the encoder's training complete, we returned to the original base dataset. Each country-month matrix was again split into $512\times50$ matrices.
-- Each of these samples could then be passed into the encoder, which distilled it into a $length-16$ vector.
-- Each vector was then stacked, resulting in a single training example for each country and month of size $16\times\frac{n}{50}$.
+::::
 

### 2023-05-10 docs/hes_final/presentation.qmd
Transformer Slide Edits

@@
 - the “encoder” and the “classifier”.
-- The encoder was used to reduce the dimensionality of the embedding corpus by analyzing a subset of article embeddings and returning a single vector of substantially reduced size.
-- Analogous to the role of a convolutional layer in a CNN. 
+- The encoder was used to distil the embedding corpus by analyzing a subset of article corpus and reducing it's dimensionality.
 - The classifier could then be used to make final predictions on the resulting, distilled dataset.

### 2023-05-11 docs/hes_final/presentation.qmd
Removed some redundant slides

@@
 
-## Three approaches to extracting conflict signals from text embeddings
-
-- The first approach used supervised dimensionality reduction and a feed-forward neural network conditioned with country context characteristics.
-- In the second approach, we used a compact dataset of monthly embedding averages per country per month and applied statistical models such as K-Nearest Neighbors and XGBoost.
-- In the third approach we stacked multiple article-level embeddings and applied transformer-based models to extract a signal exclusively from the text corpus.
-
 ## Defining Conflict Escalation
@@
 
-## Introducing our 'spike' conflict metric
+## The 'Spike' conflict metric
 
-- We derived a metric designed to highlight relative “spikes” in political violence. Calculating whether a spike occurred in a given month is a two-step process. 
-- First, we calculated the trend in violence for a given month, defined by the slope of OLS regression of the number of ACLED-documented fatalities in target month & preceding 2 months.
-- Next, we calculated whether that slope was in the 75th percentile of such slopes in the target country for the preceding 2 years.
+- We derived a metric to highlight relative “spikes” in political violence.
+- Two step calculation
+  + First, we calculated the trend in violence for a given month
+    - Defined by the slope of OLS regression of the number of ACLED-documented fatalities in target month & preceding 2 months. 
+  + Next, we calculated whether that slope was in the 75th percentile of such slopes in the target country for the preceding 2 years.
 
@@
 :::
----
+
+### Spike Metric Cont...
 
@@
 
-- Our 'spike' binary metric is sensitive to small increases in political violence in peaceful countries/periods, while also filtering out small fluctuations in long-standing conflicts.
+- Sensitive to small increases in political violence in peaceful countries/periods, while also filtering out small fluctuations in long-standing conflicts.
 
@@
 
----
-
+## Mitigating Risk
 - Models, not oracles 

### 2023-05-11 docs/hes_final/presentation.qmd
tweak spike slide headings
Update statistical model slide

@@
 
-## The 'Spike' conflict metric
+## Our 'Spike' conflict metric
 
@@
 
-### Spike Metric Cont...
+## 'Spike' conflict metric example
 
@@
 
-- In this approach we prepared the dataset by averaging news article embeddings per country, per month.
-We also included one-hot encoded columns for countries, but no other country features were added.
-- Given the high dimensionality of the data (512-dimensional embedding vectors + 700 dummy variables for countries), this dataset suffered from the “curse of dimensionality”.
+:::: {.columns}
 
----
-
+::: {.column width="50%"}
 ::: {#fig-xgboost layout-ncol=1}
@@
 :::
+:::
+
+::: {.column width="50%"}
+- Average article embeddings per country per month
+- One-hot encoded country labels
+- Reduce dimensionality with PCA
+- Train two models: K Nearest Neighbors and XGBoost
+- Average their predictions
+:::
+
+::::
 
 :::{.notes}
+- In this approach we prepared the dataset by averaging news article embeddings per country, per month.
+We also included one-hot encoded columns for countries, but no other country features were added.
+- Given the high dimensionality of the data (512-dimensional embedding vectors + 700 dummy variables for countries), this dataset suffered from the “curse of dimensionality”. 
 - This was handled by performing Principal Component Analysis; we found that reducing the number of dimensions to represent 80% of variance (~169 components with our selected train-test cutoff) improved the overall performance.

### 2023-05-11 docs/hes_final/presentation.qmd
Intro Slides Rework

--- 2023-05-11 docs/hes_final/presentation.qmd
@@
 ## Introduction
-- The task of forecasting conflict escalation is of great interest to policy makers, executives, and social scientists. 
-- We introduce a conflict escalation definition and a novel binary classification metric.
-- We outline 3 methods to extract generalizable conflict signals directly from a corpus of over 2.5 million document embeddings derived from global news sources using contemporary natural language processing and machine learning methods.
+- Forecasting conflict escalation is of interest to policy makers, executives, and social scientists. 
+- In our paper, we introduce:
+  + A novel binary classification metric for conflict escalation
+  + Three methods to extract generalizable conflict signals directly from document embeddings 
+    - Text with country characteristics  
+    - Text with country labels
+    - Text alone
 
-## Conflict forecasting
-
-- Predicting conflict has long been of interest to researchers and practitioners of international relations, and statistical and computational methods to do so can be found in the literature as early as 1944.
-- Modern methodologies that resemble those we outline here began to appear in the literature in the early 2010s, when researchers began using holdout datasets for model validation. 
-- This shift followed work by Ward et al. that sharply criticized the use of statistical significance tests on in-sample prediction to validate conflict models.
+:::{.notes}
+- Embeddings derived from global news sources using contemporary natural language processing and machine learning methods.
+- Methds use contemporary NLP and ML methods
+:::
 
@@
 
-- Today, much of the work in this area has continued to focus on country or region-specific models to predict conflict. 
-- This introduces a “small-data problem” where relatively few observations of conflict are available for a given country. 
-- Some efforts have been made to overcome this by building generalizable models that can theoretically be used to make predictions on any country.
+- Most contemporary work focuses on country or region-specific models. 
+  + This introduces a “small-data problem” 
+- We chose to persue a global model, a relatively new area of reseatch in this domain
+  + This introduces a new problem of accounting for local conditions
 
@@
 
-## Methodology
-
-- Our approach is text-first, making predictions primarily or exclusively from text, with few or no other features added.
-- To the best of our knowledge, no conflict prediction model described in the literature makes use of text embeddings.
-- Previous efforts have included statistical analysis of text, and unsupervised Latent Dirichlet Allocation (LDA) for topic extraction.
-
 ## Defining Conflict Escalation
 
-- Our task of modeling conflict escalation required a precise definition of what conflict escalation means. 
-- We began with raw event and casualty counts from the [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/), whose team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities .
+- Modeling conflict escalation requires a precise definition of what conflict escalation means. 
+- We treat this as a binary classification task
+- A global model requires a metric that is consistent acoss local conditions. 
+  + The presence or absence of conflict is not sufficient.
 
-::: {.aside}
-- Prior work has treated this as a binary classification task, asking whether or not conflict escalated in a given month, and we adopt this approach as well.
-- In contrast to prior work, however, we do not treat the simple presence of a fatality due to conflict as sufficient to indicate conflict escalation. 
+::: aside 
+- We used data from from the [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/) for our raw event and fatality counts
 :::
@@
 ::: {.notes}
+- prior work also treats this as binary classification
+- prior work uses the presence or absense of fatalities. This may work in local models
+- One study used the presence of fatalities if preceeded by the absense of fatalities
+- ACLED's team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities. 
 - ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and protests.
@@
 
-## The 'Spike' conflict metric
+## Our "Spike" Conflict Metric
+:::: {.columns}
 
-- We derived a metric to highlight relative “spikes” in political violence.
+::: {.column width="50%"}
+![The Spike metric applied to Ukraine](imgs/Ukraine_Spikes.png){#fig-spike}
+
+Spike Formula:
+$S = [\beta_m \geq Q_3(\beta)]$
+:::
+
+::: {.column width="50%"}
+- Highlights relative “spikes” in political violence.
+
+- Sensitive to small increases in political violence in peaceful countries/periods
+- Filters out small fluctuations in long-standing conflicts.
+:::
+
+::::
+
+:::{.notes}
+- Formula Explanation:
+  + $\beta_m$ is the slope of the OLS regression for the given month
+  + #Q_3(\beta)]$ is the third quartile of such slopes for the preceeding two years
+
 - Two step calculation
@@
   + Next, we calculated whether that slope was in the 75th percentile of such slopes in the target country for the preceding 2 years.
-
-:::{.notes}
-This trend allowed us to estimate whether violence was increasing or decreasing in the target country at any time.
-:::
-
-### Spike Metric Cont...
-
-![The Spike metric applied to Ukraine](imgs/Ukraine_Spikes.png){#fig-spike}
-
-- Sensitive to small increases in political violence in peaceful countries/periods, while also filtering out small fluctuations in long-standing conflicts.
-
-:::{.notes}
-
 - @fig-spike shows how our metric captures conflict fluctuations in Ukraine.
@@
 However, much larger fluctuations in the period following the invasion are filtered out, as the spike formula adjusts to the higher level of baseline conflict.
-
 - The increase in violence during the Ukrainian counter-offensive in the late summer again triggers a spike, and this reflects significant changes in the situation on the ground.
 :::
+
+## Text-First Approach
+
+- In contrast to prior work, our approach is text-first
+  + Conflict signals are extracted primarily or exclusively from text
+- Previous efforts have included statistical analysis of text and topic extraction as supplementary features
+- To the best of our knowledge, no conflict prediction model described in the literature makes use of text embeddings.
 
@@
 - The [Global Database of Events Language and Tone (GDELT) Project](https://www.gdeltproject.org/) provides public APIs of current events derived from news articles scraped from sources around the world .
-- In 2021, GDELT introduced its Global Similarity Graph Document Embeddings dataset, which precomputes 512-dimension document-level embeddings of their scraped news articles using the Universal Sentence Encoder.
+- In 2021, GDELT introduced its Global Similarity Graph Document Embeddings dataset.
+  + Provides 512-dimension document-level embeddings of scraped news articles using the Universal Sentence Encoder.
 
@@
 ```
+:::{.notes}
+- Spike is attached to embeddings on a 3-month lag
+:::
+
 Data generation process
@@
 - Our third approach used a pair of encoder-only transformers
-- the “encoder” and the “classifier”.
-- The encoder was used to distil the embedding corpus by analyzing a subset of article corpus and reducing it's dimensionality.
-- The classifier could then be used to make final predictions on the resulting, distilled dataset.
+- The “encoder” and the “classifier”.
+- The encoder distils the embedding corpus by analyzing a subset of article corpus and reducing it's dimensionality.
+- The classifier makes final predictions on the distilled dataset.
+---
 
----
+### The Encoder
 
@@
 ::: {.column width="50%"}
-Training the Encoder
 
-- Each country/month sample was split into subsamples of 50 articles each
-- The same label was used for each subsample, creating multiple unique training samples per country per month 
+- Trained using a dataset that drew multiple samples per country per month of 50 articles each 
+- The same label was used for each subsample, creating multiple unique training examples for each country and month 
 - This produced a larger, but more distorted corpus
@@
 
+### Training the Classifier
+
 :::: {.columns}
@@
 ::: {.column width="50%"}
-Training the Classifier
+- Once the Encoder was trained, it's penultimate layer of 16 dense neurons was exposed and the model was frozen
+- Each country/month matrix was then fed into in batches of 50 articles, yeilding a single training example for each country and month of size $16\times\frac{n}{50}$
 
-- Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons. 
-- Each vector was then stacked, resulting in a single training example for each country and month of size $16\times\frac{n}{50}$. 
 :::
@@
 The model was not conditioned by country, and no additional features were added beyond the text corpus itself.
-:::
-
----
-
-::: {#fig-transformer layout-ncol=1}
-```{mermaid}
-%%| fig-width: 5
-%%| fig-height: 5
-flowchart TD
-    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)
-
-    Articles --> Trans1{Encoder\n transformer}
-    Trans1 -->|classifier head present during training| Compact(16 dimension representations)
-
-    Compact --> Trans2{Binary\n classifier\n transformer}
-    Trans2 --> Pred(Prediction of conflict escalation in 3 months)
-
-    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
-    Spike(Conflict response\n binary variable) --> Trans2
-
-    Spike -->|included during model training| Trans1
-```
-Transformer model architecture
 :::

### 2023-05-11 docs/hes_final/presentation.qmd
Update research/future work/conclusions slides

--- 2023-05-11 docs/hes_final/presentation.qmd
@@
 
-## Results
-
-- We evaluated the models against a set of common performance metrics (see @fig-best-min-metrics, @fig-best-max-metrics, @fig-roc). 
-- Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparable results.
-- Each of the models outperformed the other two in a subset of these metrics.
-
----
-
-- We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives.
-At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful.
-- Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics.
-
----
+## Results: metrics
 
@@
 - The relative ranks of the models are the opposite for these two metrics with each of the two models with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
+- We evaluated the models against a set of common performance metrics (see @fig-best-min-metrics, @fig-best-max-metrics, @fig-roc). 
+- Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparable results.
+- Each of the models outperformed the other two in a subset of these metrics. 
+- We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives.
+At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful.
+- Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics.
 :::
@@
 
-![Model metrics (smaller values are better)](imgs/fig-best-min-metrics.png){#fig-best-min-metrics}
+## Results: comparison to prior work
 
----
+:::: {.columns}
 
+::: {.column width="50%"}
 ![Receiver Operator Characteristics](imgs/fig-roc.png){#fig-roc}
+:::
 
+::: {.column width="50%"}
+
+- Our best ROC_AUC is 0.69
+- Comparing to 0.83 overall and 0.75 for hard cases in paper by Mueller and Rauh 2022 (another binary classification problem)
+- Comparison is not very meaningful because our target is different
+:::
+::::
 :::{.notes}
@@
 
-![FFNN model accuracy on test data](imgs/fig-ffnn-test-accuracy.png){#fig-ffnn-test-accuracy}
+## Results: generalizability
 
-![XGBoost model accuracy on test data](imgs/fig-xgboost-test-accuracy.png){#fig-xgboost-test-accuracy}
+![FFNN model accuracy on test data](imgs/fig-ffnn-test-accuracy.png){#fig-ffnn-test-accuracy height="100px"}
 
-![Transformer model accuracy on test data](imgs/fig-transformer-test-accuracy.png){#fig-transformer-test-accuracy}
+![XGBoost model accuracy on test data](imgs/fig-xgboost-test-accuracy.png){#fig-xgboost-test-accuracy height="100px"}
+
+![Transformer model accuracy on test data](imgs/fig-transformer-test-accuracy.png){#fig-transformer-test-accuracy height="100px"}
 
@@
 
+## Results: practicality
+
+- Practical goal -- a tool that can assist UNICEF
+- Cost & complexity are important
+- Average monthly embeddings dataset & corresponding statistical model fit in memory
+- Article-level embeddings in our other methods require additional data engineering steps to be processed in batches
+
+:::{.notes}
 - While performance is often the main objective when developing prediction models, our practical goal was a tool that can assist UNICEF, so cost & complexity is an important consideration.
 - Average monthly embeddings dataset & corresponding statistical model fit in memory, whereas article-level embeddings in our other methods require additional data engineering steps to be processed in batches.
-- All three approaches produced promising results so the model choice depends on the user's goals, priority performance metrics, and available resources.
+:::
+
+## So which model is best?
+
+- All three approaches produced promising results
+- The model choice depends on:
+  - user's goals
+  - priority performance metrics
+  - available resources
 
@@
 
-- We foresee several ways to extend this work, and there are exciting opportunities for both further research and operationalizations.
+- Increase size of training set (we only used 20%)
+- Continue experimentation with data sampling and aggregation approaches and model tuning
+- Research regional disparities in model performance
+- Build an application with user interface
+- Build a more robust and automated data pipeline
 
-## Future research directions
-
+:::{.notes}
 - Our work used only 20% of the daily embeddings available through GDELT: increasing size of training sets could yield better results. 
@@
 Understanding the limitations of global media coverage for conflict prediction will be critical to any effort to operationalize this work.
+:::
 
@@
 
-- In this work we explored the possibility of predicting future geopolitical conflict escalation at a country level using a global collection of news publications.
-We compared the usage of text-only data and data enriched with country labels and social indicators.
-- We introduce a conflict escalation definition and a novel binary classification metric.
+- Text-only data vs data enriched with country labels and social indicators
+- Conflict escalation definition
+- Text embeddings provide a strong enough signal for predictions
+- Comparable results across different data sampling and modeling approaches
+- Neural network-based models -- better generalization across geographies
+- Statistical model -- smaller size/cost of deployment
 
----
+## So what?
 
-- The results suggest that text embeddings provide a strong enough signal to make such predictions. 
-- Different data sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common numeric metrics.
+- We hope this work will contribute to both UNICEF operations and political science research
+- The performance results can serve as a benchmark for future research projects in this field
 
+:::{.notes}
+We explored the possibility of predicting future geopolitical conflict escalation using a global collection of news publications
+:::
 ---

### 2023-05-11 docs/hes_final/presentation.qmd

--- 2023-05-11 docs/hes_final/presentation.qmd
@@
 
-## Results
-
-- We evaluated the models against a set of common performance metrics (see @fig-best-min-metrics, @fig-best-max-metrics, @fig-roc). 
-- Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparable results.
-- Each of the models outperformed the other two in a subset of these metrics.
-
 ---
 
-- We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives.
-At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful.
-- Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics.
+::: {#fig-transformer layout-ncol=1}
+```{mermaid}
+%%| fig-width: 5
+%%| fig-height: 5
+flowchart TD
+    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)
 
----
+    Articles --> Trans1{Encoder\n transformer}
+    Trans1 -->|classifier head present during training| Compact(16 dimension representations)
+
+    Compact --> Trans2{Binary\n classifier\n transformer}
+    Trans2 --> Pred(Prediction of conflict escalation in 3 months)
+
+    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
+    Spike(Conflict response\n binary variable) --> Trans2
+
+    Spike -->|included during model training| Trans1
+```
+Transformer model architecture
+:::
+
+## Results: metrics
 

### 2023-05-11 docs/hes_final/presentation.qmd
Transformer Slide Tweaks

--- 2023-05-11 docs/hes_final/presentation.qmd
@@
 - Trained using a dataset that drew multiple samples per country per month of 50 articles each 
-- The same label was used for each subsample, creating multiple unique training examples for each country and month 
-- This produced a larger, but more distorted corpus
+- This created multiple unique training examples for each country and month 
+- Once trained, the Encoder's weights were frozen and penultimate layer was exposed
 :::
@@
 
-### Training the Classifier
+### The Classifier
 
@@
 ::: {.column width="50%"}
-- Once the Encoder was trained, it's penultimate layer of 16 dense neurons was exposed and the model was frozen
-- Each country/month matrix was then fed into in batches of 50 articles, yeilding a single training example for each country and month of size $16\times\frac{n}{50}$
+- Each country/month embedding matrix was passed into the Encoder in batches of 50 articles
+- This produced a single distilled training example for each country and month
+- This final, undistorted data set was then used for classification
 

### 2023-05-11 docs/hes_final/presentation.qmd
Push more to notes

--- 2023-05-11 docs/hes_final/presentation.qmd
@@
 :::{.notes}
-We explored the possibility of predicting future geopolitical conflict escalation using a global collection of news publications
-:::
----
-
+- We explored the possibility of predicting future geopolitical conflict escalation using a global collection of news publications
 - The neural network-based models appear to generalize better across geographies while the statistical model provided a size/cost of deployment advantage.
@@
 - The performance results can serve as a benchmark for future research projects in this field.
+:::

### 2023-05-11 docs/hes_final/presentation.qmd
make bullets of ffnn features

--- 2023-05-11 docs/hes_final/presentation.qmd
@@
 
-- 2 numerical indicators from UNICEF's State of the World's Children report: 'Under-five mortality rate 2019' & 'Adolescent population 2020 Proportion of total population (%) Total'
-- 3 one-hot-encoded categorical indicators: Human Development Index ranges, categories of gross national income per capita, & categories of Fragile States Index ranges)
+- 2 numerical indicators from UNICEF's State of the World's Children report: 
+  + 'Under-five mortality rate 2019' 
+  + 'Adolescent population 2020 Proportion of total population (%) Total'
+- 3 one-hot-encoded categorical indicators: 
+  + Human Development Index ranges 
+  + gross national income per capita
+  + Fragile States Index ranges
 
@@
 - Trained using a dataset that drew multiple samples per country per month of 50 articles each 
-- This created multiple unique training examples for each country and month 
-- Once trained, the Encoder's weights were frozen and penultimate layer was exposed
+- The same label was used for each subsample, creating multiple unique training examples for each country and month 
+- This produced a larger, but more distorted corpus
 :::
@@
 
-### The Classifier
+### Training the Classifier
 
@@
 ::: {.column width="50%"}
-- Each country/month embedding matrix was passed into the Encoder in batches of 50 articles
-- This produced a single distilled training example for each country and month
-- This final, undistorted data set was then used for classification
+- Once the Encoder was trained, it's penultimate layer of 16 dense neurons was exposed and the model was frozen
+- Each country/month matrix was then fed into in batches of 50 articles, yeilding a single training example for each country and month of size $16\times\frac{n}{50}$
 

### 2023-05-11 docs/hes_final/presentation.qmd
fix typo

@@
   + This introduces a “small-data problem” 
-- We chose to persue a global model, a relatively new area of reseatch in this domain
+- We chose to persue a global model, a relatively new area of research in this domain
   + This introduces a new problem of accounting for local conditions

### 2023-05-11 docs/hes_final/presentation.qmd
move ffnn context features to speaker notes

@@
 
-## Additional country context inputs
-
-- 2 numerical indicators from UNICEF's State of the World's Children report: 
-  + 'Under-five mortality rate 2019' 
-  + 'Adolescent population 2020 Proportion of total population (%) Total'
-- 3 one-hot-encoded categorical indicators: 
-  + Human Development Index ranges 
-  + gross national income per capita
-  + Fragile States Index ranges
-
-:::{.notes}
-- avoid the per-country 'small data problem' by allowing model to learn similarities between country contexts and generalize
-:::
-
 ---
@@
 Feed forward neural network model architecture
+:::
+
+:::{.notes}
+- avoid the per-country 'small data problem' by allowing model to learn similarities between country contexts and generalize
+- 2 numerical indicators from UNICEF's State of the World's Children report: 
+  + 'Under-five mortality rate 2019' 
+  + 'Adolescent population 2020 Proportion of total population (%) Total'
+- 3 one-hot-encoded categorical indicators: 
+  + Human Development Index ranges 
+  + gross national income per capita
+  + Fragile States Index ranges
 :::

### 2023-05-11 docs/hes_final/presentation.qmd
Map Edits

--- 2023-05-11 docs/hes_final/presentation.qmd
@@
 
-Spike Formula:
-$S = [\beta_m \geq Q_3(\beta)]$
 :::
@@
 :::{.notes}
-- Formula Explanation:
-  + $\beta_m$ is the slope of the OLS regression for the given month
-  + #Q_3(\beta)]$ is the third quartile of such slopes for the preceeding two years
-
 - Two step calculation
@@
 
-- 2 numerical indicators from UNICEF's State of the World's Children report: 
-  + 'Under-five mortality rate 2019' 
-  + 'Adolescent population 2020 Proportion of total population (%) Total'
-- 3 one-hot-encoded categorical indicators: 
-  + Human Development Index ranges 
-  + gross national income per capita
-  + Fragile States Index ranges
+- 2 numerical indicators from UNICEF's State of the World's Children report: 'Under-five mortality rate 2019' & 'Adolescent population 2020 Proportion of total population (%) Total'
+- 3 one-hot-encoded categorical indicators: Human Development Index ranges, categories of gross national income per capita, & categories of Fragile States Index ranges)
 
@@
 
-![FFNN model accuracy on test data](imgs/fig-ffnn-test-accuracy.png){#fig-ffnn-test-accuracy height="100px"}
+::: {#fig-elephants layout-ncol=2}
 
-![XGBoost model accuracy on test data](imgs/fig-xgboost-test-accuracy.png){#fig-xgboost-test-accuracy height="100px"}
+![FFNN](imgs/fig-ffnn-test-accuracy.png){#fig-ffnn-test-accuracy}
 
-![Transformer model accuracy on test data](imgs/fig-transformer-test-accuracy.png){#fig-transformer-test-accuracy height="100px"}
+![XGBoost](imgs/fig-xgboost-test-accuracy.png){#fig-xgboost-test-accuracy}
+
+![Transformer](imgs/fig-transformer-test-accuracy.png){#fig-transformer-test-accuracy}
+
+Model Accuracy on Test Set
+:::
 
@@
 :::{.notes}
-- We explored the possibility of predicting future geopolitical conflict escalation using a global collection of news publications
+We explored the possibility of predicting future geopolitical conflict escalation using a global collection of news publications
+:::
+---
+
 - The neural network-based models appear to generalize better across geographies while the statistical model provided a size/cost of deployment advantage.
@@
 - The performance results can serve as a benchmark for future research projects in this field.
-:::

### 2023-05-11 docs/hes_final/presentation.qmd

@@
 
-## Additional country context inputs
-
-- 2 numerical indicators from UNICEF's State of the World's Children report: 'Under-five mortality rate 2019' & 'Adolescent population 2020 Proportion of total population (%) Total'
-- 3 one-hot-encoded categorical indicators: Human Development Index ranges, categories of gross national income per capita, & categories of Fragile States Index ranges)
-
-:::{.notes}
-- avoid the per-country 'small data problem' by allowing model to learn similarities between country contexts and generalize
-:::
-
 ---

### 2023-05-11 docs/hes_final/presentation.qmd
Reorder ethics slides
edit speaker notes, changes after dress rehersal

@@
 :::{.notes}
-- Embeddings derived from global news sources using contemporary natural language processing and machine learning methods.
-- Methds use contemporary NLP and ML methods
+- Embeddings derived from global news sources
 :::
@@
 :::{.notes}
-- We followed the latter approach in developing our model, hypothesizing that the larger number of observations available to a global model will allow it to extract more generalizable signals of conflict, and that such a model will be better able to predict unforeseen risks in countries with little available training data.
+- small data - some countries have not experienced much armed conflict, so very little training data
+- local conditions - baseline monthly conflict fatalities vary widely from country to country, so importance of changes is relative
 :::
@@
 - Modeling conflict escalation requires a precise definition of what conflict escalation means. 
-- We treat this as a binary classification task
+  + Many definitions in the literature.
 - A global model requires a metric that is consistent acoss local conditions. 
@@
 ::: {.notes}
-- prior work also treats this as binary classification
+- prior work treats this as binary classification
 - prior work uses the presence or absense of fatalities. This may work in local models
-- One study used the presence of fatalities if preceeded by the absense of fatalities
-- ACLED's team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities. 
-- ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and protests.
-With these totals in-hand, we needed to create a target to label our training data.
-- To achieve our objective of creating a generalizable conflict model, we required a metric that was consistent across varying local conditions. 
-- Small changes in the absolute number of monthly fatalities can have widely different implications depending on the country and time period. 
+- for a global conflict model, we required a metric that was consistent across varying local conditions. 
 :::
@@
 - Highlights relative “spikes” in political violence.  
-
-- Sensitive to small increases in political violence in peaceful countries/periods
+- Sensitive to small increases in political violence in peaceful countries/periods.
 - Filters out small fluctuations in long-standing conflicts.
@@
 - @fig-spike shows how our metric captures conflict fluctuations in Ukraine. 
-Prior to the Russian invasion of February 2022, small fluctuations trigger a spike.
-However, much larger fluctuations in the period following the invasion are filtered out, as the spike formula adjusts to the higher level of baseline conflict.
-- The increase in violence during the Ukrainian counter-offensive in the late summer again triggers a spike, and this reflects significant changes in the situation on the ground.
-:::
-
-## Text-First Approach
-
-- In contrast to prior work, our approach is text-first
-  + Conflict signals are extracted primarily or exclusively from text
-- Previous efforts have included statistical analysis of text and topic extraction as supplementary features
-- To the best of our knowledge, no conflict prediction model described in the literature makes use of text embeddings.
-
-## News Content Data
-
-- The [Global Database of Events Language and Tone (GDELT) Project](https://www.gdeltproject.org/) provides public APIs of current events derived from news articles scraped from sources around the world .
-- In 2021, GDELT introduced its Global Similarity Graph Document Embeddings dataset.
-  + Provides 512-dimension document-level embeddings of scraped news articles using the Universal Sentence Encoder.
-
-:::{.notes}
-- GDELT’s coverage over-represents the western world, particularly the United States and Europe.
-- GDELT does represent Russia, India and China better than other event datasets, but it continues to underrepresent other parts of the world, particularly the global south.
-- English is the most common language in the GDELT dataset, comprising more than 40% of the articles in its corpus.
+- Prior to the Russian invasion of February 2022, small fluctuations trigger a spike. 
+- Larger absolute fluctuations of fatalities in following period are not considered 'spikes', as the spike formula adjusts to a higher conflict baseline. 
+- The increase in fatalities during the Ukrainian counter-offensive in the late summer again triggers a spike, and this reflects significant escalation of the situation in the country.
 :::
@@
 ```
-:::{.notes}
-- Spike is attached to embeddings on a 3-month lag
-:::
-
 Data generation process
@@
 :::{.notes}
-- Our training data was derived from these GDELT and ACLED datasets. 
+- In contrast to prior work, our approach is text-first
+  + Conflict signals are extracted primarily or exclusively from text
+- Our training data was derived from 
+- Global Database of Events Language and Tone (GDELT) Project
+- Armed Conflict Location & Event Data Project
 - GDELT news article embeddings served as our primary independent variable, while the “spike” metric derived from ACLED’s conflict data served as our target variable.
-- The spike metric was used to label article data on a three-month time lag, allowing us to make predictions of future events.
-- Our training dataset thus ran from January 2020 through October 2022. Data from November 2022 through March 2023 was used for model validation and testing.
+- The spike metric was used to label article data with a three-month time lag, allowing us to make predictions of future events.
+- Previous efforts have included statistical analysis of text and topic extraction as supplementary features
+- To the best of our knowledge, no conflict prediction model described in the literature makes use of text embeddings.
 :::
@@
 
-- Our first approach employed a feed forward neural network model. 
+- Our first modeling approach employed a feed forward neural network model. 
 - Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner.
@@
 :::{.notes}
-- Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space. 
+- Reduction of 512 dimension embeddings to 128 dimension representation using UMAP
+- dimensionality reduction process conditioned on our spike variable to encourage separation in the reduced embedding space
 - The fit UMAP model was applied to the validation and test data without supervision. 
-These reduced, 128-dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons.
-The model's 4577 parameters were trained for 3 epochs.
 :::
@@
 :::{.notes}
-- avoid the per-country 'small data problem' by allowing model to learn similarities between country contexts and generalize
+- 128-dimension reduced embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons.
+- intended to avoid the 'small data problem' in some countries by allowing the model to learn similarities between country contexts and generalize
 - 2 numerical indicators from UNICEF's State of the World's Children report: 
@@
 
+## Ethical Considerations
+- Bias in the training data
+- Adversarial attacks
+- Self-fulfilling prophesies
+
+:::{.notes}
+- GDELT’s coverage over-represents the western world, particularly the United States and Europe.
+- GDELT does represent Russia, India and China better than other event datasets, but it continues to underrepresent other parts of the world, particularly the global south.
+- English is the most common language in the GDELT dataset, comprising more than 40% of the articles in its corpus.
+- All models showed varying performance by region, could lead to resources being misallocated. 
+- Large volumes of fake news could overwhelm true signal, obscure intentions of malicious actors & cause panic.
+- AI predictions of impending conflict may lead policy makers to take actions that make conflict more likely. 
+:::
+
+## Mitigating Risk
+- Models, not oracles 
+- Not a substitute for careful deliberation and dialogue
+- At their best, these models can provide an early warning for conflict, and if the right actions are taken, then they can be made to be wrong
+
+:::{.notes}
+- Measuring the temperature of the global media conversation, not gazing into the future.
+:::
+
 ## Results: metrics
@@
 - The relative ranks of the models are the opposite for these two metrics with each of the two models with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
-- We evaluated the models against a set of common performance metrics (see @fig-best-min-metrics, @fig-best-max-metrics, @fig-roc). 
+- We evaluated the models against a set of common performance metrics. 
 - Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparable results.
@@
 
+- Comparison is not very meaningful because our target is different
+- Comparing to 0.83 overall and 0.75 for hard cases in paper by Mueller and Rauh 2022 (another binary classification problem)
 - Our best ROC_AUC is 0.69
-- Comparing to 0.83 overall and 0.75 for hard cases in paper by Mueller and Rauh 2022 (another binary classification problem)
-- Comparison is not very meaningful because our target is different
 :::
@@
 :::{.notes}
-- Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world—see @fig-ffnn-test-accuracy, @fig-xgboost-test-accuracy, and @fig-transformer-test-accuracy.
+- Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world. 
 :::
@@
   - available resources
-
-## Ethical Considerations
-- Bias in the training data
-- Adversarial attacks
-- Self-fulfilling prophesies
-
-:::{.notes}
-- All models showed varying performance by region, could lead to resources being misallocated. 
-- Large volumes of fake news could overwhelm true signal, obscure intentions of malicious actors & cause panic.
-- AI predictions of impending conflict may lead policy makers to take actions that make conflict more likely. 
-:::
-
-## Mitigating Risk
-- Models, not oracles 
-- Not a substitute for careful deliberation and dialogue
-- At their best, these models can provide an early warning for conflict, and if the right actions are taken, then they can be made to be wrong
-
-:::{.notes}
-- Measuring the temperature of the global media conversation, not gazing into the future.
-:::
 
@@
 :::
+
+## Thank you

### 2023-05-11 docs/hes_final/presentation.qmd
Update notes in results

--- 2023-05-11 docs/hes_final/presentation.qmd
@@
 
+## Ethical Considerations
+- Bias in the training data
+- Adversarial attacks
+- Self-fulfilling prophesies
+
+:::{.notes}
+- All models showed varying performance by region, could lead to resources being misallocated. 
+- Large volumes of fake news could overwhelm true signal, obscure intentions of malicious actors & cause panic.
+- AI predictions of impending conflict may lead policy makers to take actions that make conflict more likely. 
+:::
+
+## Mitigating Risk
+- Models, not oracles 
+- Not a substitute for careful deliberation and dialogue
+- At their best, these models can provide an early warning for conflict, and if the right actions are taken, then they can be made to be wrong
+
+:::{.notes}
+- Measuring the temperature of the global media conversation, not gazing into the future.
+:::
+
 ## Results: metrics
@@
 :::{.notes}
-- The relative ranks of the models are the opposite for these two metrics with each of the two models with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
-- We evaluated the models against a set of common performance metrics (see @fig-best-min-metrics, @fig-best-max-metrics, @fig-roc). 
+
+- We evaluated the models against a set of common performance metrics
 - Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparable results.
@@
 - Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics.
+- The relative ranks of the models are the opposite for these two metrics with each of the two models with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
 :::
@@
 :::{.notes}
+- Area Under the Receiver Operating Characteristic Curve
 - Our target was defined differently from related work which made direct comparison less meaningful.
@@
 :::{.notes}
-- Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world—see @fig-ffnn-test-accuracy, @fig-xgboost-test-accuracy, and @fig-transformer-test-accuracy.
+- We wanted to build a globally generalizable model.
+Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world.
 :::
@@
 - Cost & complexity are important
-- Average monthly embeddings dataset & corresponding statistical model fit in memory
-- Article-level embeddings in our other methods require additional data engineering steps to be processed in batches
+- Average monthly embeddings dataset & corresponding statistical model fit in memory and can be trained on CPU
+- Article-level embeddings in our other methods require additional data engineering steps to be processed in batches and need GPU to train
 

### 2023-05-11 docs/hes_final/presentation.qmd

--- 2023-05-11 docs/hes_final/presentation.qmd
@@
 :::{.notes}
-- The relative ranks of the models are the opposite for these two metrics with each of the two models with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
-- We evaluated the models against a set of common performance metrics. 
+
+- We evaluated the models against a set of common performance metrics
 - Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparable results.
@@
 :::{.notes}
-- Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world. 
+- We wanted to build a globally generalizable model.
+Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world.
 :::
